### 2015-07-06 MSH.tex
First commit

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{pdflscape}
\modulolinenumbers[5]

\journal{Geochemica and Cosmochemica Acta}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

\title{The compositions of high-pressure water-rich melts in the system MgO-SiO$_2$-H$_2$O}
%\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}

%% Group authors per affiliation:
\author{R. Myhill, D. J. Frost}
\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany}
\cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
\ead{myhill.bob@gmail.com}

\author{D. Novella}
\address{Laboratoire Magmas et Volcans, Universit\'{e} Blaise Pascal, 5 Rue Kessler, 63038 Clermond-Ferrand, France}

%% or include affiliations in footnotes:
%\author[mymainaddress,mysecondaryaddress]{Elsevier Inc}
%\ead[url]{www.elsevier.com}

%\author[mysecondaryaddress]{Global Customer Service\corref{mycorrespondingauthor}}

%\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
%\address[mysecondaryaddress]{360 Park Avenue South, New York}

\begin{abstract}
High pressure multianvil experiments were conducted at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O. Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source. The topology of the liquidus is tightly constrained at much higher water contents than previously possible. In the subsystem MgO-H$_2$O, the invariant point between brucite, periclase and liquid occurs at ca. 1210 $^{\circ}$C and 67 mol\% H$_2$O. Forsterite and clinoenstatite both melt congruently in the temperature range investigated, although the fo-cen and cen-stv cotectics migrate close to the fo-H$_2$O and cen-H$_2$O binaries at high water contents. The chondrodite liquidus field approaches the fo-H$_2$O binary at ca. 1200 $^{\circ}$C. Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O, then shallowing up to 50 mol\%, similar to behaviour previously observed at 2 GPa.
\end{abstract}

\begin{keyword}
high pressure \sep mantle \sep hydrous melting \sep water \sep liquidus
\end{keyword}

\end{frontmatter}

\linenumbers

\section{Introduction}
The subduction of hydrous materials has a strong influence on the formation of melts within the deep Earth. The dehydration of hydrous crustal minerals is thought to drive the melting of the mantle wedge that feeds arc volcanoes. Dehydration of the underlying lithospheric mantle may also enable embrittlement of the slab, allowing it to deform at significantly higher strain rates than would be possible if it were dry.

There has been some controversy over whether slabs transport water beyond about 300 km depth. Certainly most of the ca. 7 km thick mafic crust is probably too warm to sustain hydrous minerals. However, hydrated ultramafics can potentially carry water into the lower mantle. Supporting evidence for an ultradeep part of the water cycle exists in the form of hydrous ringwoodite in diamond, which apparently formed under cold conditions.

The compositions of melts formed by the breakdown of hydrous phases is important for a full understanding of the deep water system. The thermodynamics of melting in water-rich systems cannot be determined without good estimates of melt compositions. Furthermore, the thermal stability of hydrous phases cannot be predicted from their properties without also having an activity-composition model for the melt phase. More generally, the density of hydrous melts will be highly composition dependent, impacting on their ability to rise through the mantle. Finally, reactions between migrating melts and the surrounding solid will be controlled by the \emph{P-T-X} dependence on melt composition.

Although the phase relations in hydrous systems in the upper mantle are now remarkably well-constrained, constraining melt compositions has remained a major challenge in high-pressure experimental petrology. The primary problem is that hydrous melts are unquenchable, which means that their compositions must be estimated indirectly. Compositions of partial hydrous melts have been estimated using mass balance, diamond traps \citep{MSUP2007}, and the Mg:Si ratio has been estimated by defocusing the beam of an electron probe microanalyser (EPMA) \citep[e.g.][]{YII2004}. Unfortunately, all three of these techniques are accompanied by significant uncertainties, especially in terms of the water content. An alternative technique is to constrain the liquidus temperature at a given composition. However, solid hydrous compounds do not reach the water concentrations expected from hydrous melts in the vicinity of a subducting slab. Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form. Extrapolation is prone to large errors, especially when high temperatures facilitate water loss through capsule walls. Water masses can be accurately calculated when added to large volume chambers, but not to the small volumes of high pressure capsules. Thus, constraining the water contents of melts in hydrous systems remains a key problem in experimental petrology.

In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$) as a novel water source. Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}. It breaks down at higher temperatures, releasing four molecules of H$_2$O and forming platinum (IV) oxide PtO$_2$. Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments.

\section{Experimental and analytical methods}
Starting compositions were created from a mixture of high purity brucite (Mg(OH)$_2$, SiO$_2$ and platinic acid (H$_2$Pt(OH)$_6$). SiO$_2$ was dried overnight at 1000 $^{\circ}$C, while brucite was heated to 250 $^{\circ}$C overnight. Both powders were stored in a desiccating oven at 130 $^{\circ}$C. Platinic acid is hygroscopic, so was stored in a vacuum desiccator. Powders were weighed and ground dry in an agate mortar for 30 minutes, using a mask, goggles and gloves to avoid physical contact with the platinic acid. Starting compositions were stored in glass vials in a vacuum desiccator. Compositions are listed in Table \ref{table:compositions}.

\begin{table}[ht!]
\caption{Starting compositions}
\label{table:compositions}
\begin{tabular}{l|lll|lll}
 & \multicolumn{3}{c}{Oxide proportions (mol\%)} & \multicolumn{3}{|c}{Compound fractions (mol/mol)} \\
\hline
 & \multicolumn{1}{c}{MgO} & \multicolumn{1}{c}{SiO$_2$} & \multicolumn{1}{c}{H$_2$O} & \multicolumn{1}{|c}{Mg(OH)$_2$} & \multicolumn{1}{c}{SiO$_2$} & \multicolumn{1}{c}{H$_2$Pt(OH)$_6$} \\
\hline
br 5.0 & 50.000 & 0.000 & 50.000 & 1.000 & 0.000 & 0.000 \\
br 5.5 & 45.000 & 0.000 & 55.000 & 0.947 & 0.000 & 0.053 \\
br 6.0 & 40.000 & 0.000 & 60.000 & 0.889 & 0.000 & 0.111 \\
br 6.5 & 35.000 & 0.000 & 65.000 & 0.824 & 0.000 & 0.176 \\
br 7.0 & 30.000 & 0.000 & 70.000 & 0.750 & 0.000 & 0.250 \\
en 6.0 & 30.000 & 30.000 & 40.000 & 0.480 & 0.480 & 0.040 \\
en 7.0 & 26.667 & 26.667 & 46.667 & 0.457 & 0.457 & 0.086 \\
en 8.0 & 23.333 & 23.333 & 53.333 & 0.431 & 0.431 & 0.138 \\
en 9.0 & 20.000 & 20.000 & 60.000 & 0.400 & 0.400 & 0.200 \\
fo 11.5 & 36.000 & 18.000 & 46.000 & 0.637 & 0.319 & 0.044 \\
fo 13.0 & 32.000 & 16.000 & 52.000 & 0.604 & 0.302 & 0.094 \\
fo 14.5 & 28.000 & 14.000 & 58.000 & 0.566 & 0.283 & 0.152 \\
fo 16.0 & 24.000 & 12.000 & 64.000 & 0.522 & 0.261 & 0.217 \\
q 1.5 & 0.000 & 85.000 & 15.000	& 0.000	& 0.958	& 0.042 \\
q 2.0 & 0.000 & 80.000 & 20.000 & 0.000 & 0.941 & 0.059 \\
q 2.5 & 0.000 & 75.000 & 25.000 & 0.000 & 0.923 & 0.077 \\
q 3.0 & 0.000 & 70.000 & 30.000 & 0.000 & 0.903 & 0.097 \\
q 3.5 & 0.000 & 65.000 & 35.000 & 0.000 & 0.881 & 0.119 \\
q 4.0 & 0.000 & 60.000 & 40.000 & 0.000 & 0.857 & 0.143 \\
q 4.5 & 0.000 & 55.000 & 45.000 & 0.000 & 0.830 & 0.170 \\
q 5.0 & 0.000 & 50.000 & 50.000 & 0.000 & 0.800 & 0.200 \\
br+q 2.25 & 25.000 & 50.000 & 25.000 & 25.000 & 50.000 & 0.000 \\
br+q 2.70 & 30.000 & 40.000 & 30.000 & 30.000 & 40.000 & 0.000 \\
br+q 3.2 & 35.556 & 28.889 & 35.556 & 35.556 & 28.889 & 0.000 \\
br+q 3.4 & 37.778 & 24.444 & 37.778 & 37.778 & 24.444 & 0.000
\end{tabular}
\end{table}

Capsules were created from 2 mm-diameter Pt$_{90}$Rh$_{10}$ and Au rods. They were cut by wire saw into 1 mm thick disks, into which were spark-eroded six holes in three rows, 250 microns wide and 700 microns deep. The capsules were cleaned by cycling between an acetone ultrasonic bath (15 minutes) and 1000 $^{\circ}$C furnace (20 minutes) three times. Any remaining contamination was removed with W$_{75}$Re$_{25}$ needle followed by a further trip to the ultrasonic bath. Capsule chambers were filled with powders of different compositions using a W$_{75}$Re$_{25}$ needle. Small pieces of tape were used to cover the other holes to avoid contamination.

Multianvil experiments were performed in the 5000 tonne press at the Bayerisches Geoinstitut (BGI). Cr-doped MgO octahedral multianvil assemblies with 18 mm edge length were used (Figure \ref{fig:assembly}). Two capsules were loaded into each assembly, with the open ends of the chambers facing each other and separated by six 0.05 mm thick Pt$_{90}$Rh$_{10}$ or Au foils. The assemblies were compressed to 13 GPa over four hours between eight tungsten carbide cubes with trucations of edge-length 11 mm. Pressure calibrations and details of the press can be found in \cite{FPTLDR2004} and \cite{KF2005}. The assemblies were then resistively heated via the stepped LaCrO$_3$ heater. The temperature was recorded using a W$_{97}$Re$_{3}$--W$_{75}$Re$_{25}$ (Type D) thermocouple inserted axially into the assembly. To avoid water loss, higher temperature runs were heated for a shorter duration. The experiments were quenched by cutting power to the heater. The assemblies were then decompressed to room pressure over 1000 minutes.

Capsules were recovered from the assembly, separated by wire saw and then ground by hand to reveal the tops of each capsule chamber. The wire saw was then used again to split the two sets of three chambers from each capsule. Each half-capsule was mounted in epoxy and mirror-polished. After revealing the edge of the capsule chambers, it was necessary to impregnate them with epoxy under vacuum before further polishing, in order to fill in the porous spaces created by the hydrous melt. Grinding under running water helped remove plucked grains from the polishing surface. Prepared and cleaned samples were then coated with a 10 nm thick carbon layer.

Analysis of run products was conducted via scanning electron microscope (SEM), using BSE imaging and EDS for phase identification (via the INCA software package).

\section{Results}
Experimental details and results are shown in Table \ref{table:experiments}. Superliquidus runs revealed networks of dendritic crystals of brucite, forsterite, enstatite and stishovite. Below the liquidus, solid phases tended to aggregate at the cold end of the capsules. Periclase formed equant crystals often associated with quench overgrowths which appear brighter in BSE images. Forsterite formed tabular crystals about 100 $\mu$m long. Enstatite crystals were much smaller, forming ca. 5 $\mu$m equant grains. Stishovite also formed equant to blocky grains, but crystals only separated from the melts in chambers with a high water content. In addition to the difference in texture and shape between quench crystals liquidus phases, subliquidus phases typically contained many small inclusions of Pt-Rh oxides. A typical set of run products is shown in Figure \ref{fig:sem}.

\begin{figure}[h!]
  \centering
      %\includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{The 18/11 octahedral assembly design used in this study.}
  \label{fig:assembly}
\end{figure}

\begin{landscape}
\begin{table}[ht!]
\caption{Experimental run conditions and run products determined by SEM/EDS. Compositions are listed in order of increasing molar H$_2$O content. Unused compositions for each run are marked by an en-dash. Minor solids are listed in brackets. Mineral abbreviations are as follows: br - brucite, per - periclase, co - hydroxychondrodite, en - clinoenstatite, s - stishovite.}
\label{table:experiments}
\begin{tabular}{llllccccc}
Expt \# & P (GPa) & T ($^{\circ}$C) & t (min) & brucite+PtAc & forsterite+PtAcid & enstatite+PtAcid & quartz+PtAcid & br+q \\
Z1063 & 13 & 1200 & 40 & -/br/br/br/L & co/fo/fo/(fo) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1085 & 13 & 1300 & 40 & -/-/per/L/L & -/fo/(en)/L & -/-/-/- & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1079 & 13 & 1300 & 35 & -/per/per/L/L & fo/fo/en/(en) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1058 & 13 & 1400 & 30 & -/per/(per)/L/L & fo/fo/L/L & en/en/en/(en) & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1060 & 13 & 1500 & 20 & -/-/-/-/- & fo/L/L/L & en/en/L/L & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1140 & 13 & 1600 & 10 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/s/(s) & s+en/en/en/en \\
Z1084 & 13 & 1650 & 10 & -/-/-/-/- & -/-/-/- & L/L/-/- & -/-/-/-/-/-/-/- & -/-/-/- \\
Z1207 & 13 & 1700 & 8 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/(s)/L & s+en/(en)/L/L \\
Z1209 & 13 & 1800 & 5 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/(s?)/L/L/L & s/L/L/L \\
Z1091 & 13 & 1900 & 5 & (per)/-/-/-/- & -/-/-/- & -/-/-/- & -/-/-/L/-/L/L/L & L/L/-/-
\end{tabular}
\end{table}
\end{landscape}

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/Z1058_1400C_13GPa_fo}
  \caption{BSE image of typical run products. Experimental chambers from Experiment Z1058, conducted at 1400 $^{\circ}$C and 13 GPa. Chambers have compositions along the forsterite-water binary, with water contents increasing from left to right.}
  \label{fig:sem}
\end{figure}

\section{Discussion}
\subsection{Modelling water solubility in melts}

Oxygen in MgO-SiO$_2$-H$_2$O melts can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group. The general equation describing the reaction between melt species is
\begin{equation}
0.5 \textrm{H}_2\textrm{O} + 0.5 \textrm{O}_{br}^{2-} \rightleftharpoons \textrm{OH}_{tm}^-
\label{eqn:speciation}
\end{equation}

where the subscripts br and tm respectively refer to bridging and terminal oxygens. The equilibrium between these species can be described with an equilibrium constant $K$ \cite[e.g.][]{Stolper1982}
\begin{equation}
K = \frac{X_{\textrm{OH}_{tm}^-}}{\sqrt{X_{\textrm{H}_2\textrm{O}} X_{\textrm{O}_{br}^{2-}} } }
\label{eqn:equilibrium_constant}
\end{equation}

The equilibrium constant is related to the energy $\Delta G^{\circ}$ required to protonate a bridging oxygen
\begin{equation}
K = \exp\left(\frac{-\Delta G^{\circ}}{RT}\right)
\end{equation}

These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers. A high proportion of monomers and dimers exist in relatively dilute solutions, but not in the concentrated solutions investigated in this study.

A number of models have been presented to investigate the behaviour of concentrated hydrous melts. \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation. For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made purely of monomers. If r is less then the number of oxygens per formula unit no distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment. Fixing $r$ places an implicit constraint on the maximum value of $K$ in water-rich compositions. For example, if $r=2$ in a hydrous forsteritic melt, then water contents exceeding 50\% require that $K<\infty$. In practise, the equilibrium constant $K$ has been shown to be dependent on temperature and composition. \cite{SK1995} obtained a good fit to isochemical data with an expression of the form
\begin{equation}
\Delta G^{\circ} = a + b\,T
\end{equation}

\cite{THH2012} simplified the model of \cite{SS1985} by assuming that all hydrogen in the melt exists as OH$^-$, and that protonation is equally likely on any oxygen ($K=\infty$, $\Delta G^{\circ}=-\infty$, $r$ maximised). Deviations from this model must increase with increasing water content, and such a model cannot describe any melt more water-rich than the Mg(OH)$_2$-Si(OH)$_4$ binary.

To describe melting in the SiO$_2$-H$_2$O system, \cite{HM2012} generalised the  aforementioned models by allowing non-ideality between the two endmembers. The excess term $\Delta G^{\circ}$(P,T) for the formation of OH$^-$ groups was assumed to be independent of composition; all oxygens are equally available for protonation, and the local extent of protonation does not affect the energy required for each further protonation.

In all of these models, the activity of component $x$ with the composition of the liquidus phase $X$ can be calculated from the melt model and independently from the entropy of melting of $X$. At equilibrium, these two activities must be equal:
\begin{equation}
RT \ln a_{x,L}^X = G_{X_l} - G_{X_{s}} = - \int_T^{T_{melt}} S_{X_{l}}-S_{X_s} \, dT
\end{equation}
where $T_{melt}$ is the temperature of melting of the liquidus phase under anhydrous conditions.

\citep{NM2008}
\citep{DM2010}
\citep{ZK1998}

\subsection{Subsystems}
\subsubsection{MgO-H$_2$O}
The dehydration of brucite at high pressure was previously studied by \cite{FIYKFO2005}. They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa. We observed large brucite crystals at 1200 $^{\circ}$C and 13 GPa; so assuming that errors from both studies are $\sim$ 50 $^{\circ}$C, decomposition takes place just above 1200 $^{\circ}$C. In the present study, the composition of the fluid coexisting with brucite and periclase is constrained to be 64--66 mol\% H$_2$O. The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase. The best experimental estimate of the melting point of MgO is currently that of \cite{ZF2008}, who extrapolated to the MgO endmember using melting phase relations in the system (Mg,Fe)O. Their melting curve is essentially identical to that obtained from large-scale molecular dynamics simulations \citep{CG1994}. Using these two studes, melting at 13 GPa is estimated to be $\sim 5373$ K. This value is higher than that from more recent ab-initio investigations \citep{Alfe2005,KKS2013}, and 2000 K higher than the melting temperature estimated experimentally by \cite{ZB1994}.

At 13 GPa, the entropy of melting of periclase is estimated to be $\sim$22 J/K/mol \cite{CG1994}. Assuming that the metastable entropy of melting at this pressure remains constant with temperature, the activity of periclase in the system can be determined and used to fit a melting model to the experimental data. Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -70000 - 15T$ provides a good fit to the data. However, this model yields a low activity of water in equilibrium with brucite and periclase ($a_{H_2O}\sim 0.34$) relative to that estimated using the thermodynamic data of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995} ($0.52$).

To simultaneously fit the melting data and the activity of water in melts in equilibrium with periclase and brucite, $\partial^2 G_{ex} / \partial X_{H_2O}^2$ needs to be less positive at compositions more water-rich than brucite. This can be achieved by imposing a binary subregular model (W$_{MgO-H_2O}$=38000 J/mol, W$_{H_2O-MgO}$=0 J/mol). This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system. The second critical endpoint of the MgO-H$_2$O is probably around 13 GPa \citep{MSUP2007}, so large positive interaction enthalpies are justified in this region of the phase diagram. Alternatively, a regular solution model with MgO, Mg(OH)$_2$ and H$_2$O species in the melt can adequately fit the data, with parameters G$_{br,melt} = -41000. - 10*(T-(1210.+273.15))$, W$_{MgO-H_2O}=0$, W$_{MgO-Mg(OH)_2}=-10000$, W$_{Mg(OH)_2-H_2O}=-22000$ (all in J/mol). The resulting fits are shown in Figure \ref{fig:MH}. The large differences between the two model-types at high temperature is a result of the much larger entropy of mixing in the \cite{SS1985} models close to the anhydrous endmembers.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/per-H2O}
  \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results. The position of the dry periclase melting point is taken from \cite{ZF2008}, with an entropy of dry melting from \cite{CG1994}. Model fits in grey use the model of \cite{SS1985}, with r=1 and K=$\infty$ (bottom), K=K(T) (expression in main text) and K=0 (all H2O as molecular, top). The black line includes an interaction term to match the fluid in equilibrium with brucite during it's decomposition to periclase. The black dotted line represents a regular solution model with MgO, Mg(OH)$_2$ and H$_2$O as distinct species in the melt (parameters in the main text).}
  \label{fig:MH}
\end{figure}

\clearpage
\subsubsection{Mg$_2$SiO$_4$-H$_2$O}
The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}. The anhydrous melting of forsterite is incongruent. The temperature of the fo $\rightarrow$ per + L reaction is taken from \citep{PW1993}, with the temperature of the metastable fo $\rightarrow$ L reaction extrapolated from the same study. The liquidus is calculated by interpolation between the melting temperature of periclase \citep{ZF2008} and an estimate of the temperature and composition of melt in equilibrium with periclase and forsterite by interpolating between the invariant fo-per-foL point \citep{PW1993} and position of the per liquidus at 16 GPa \citep{LF2012}.

One chamber from an experiment run at 1300 $^{\circ}$C revealed a single very small crystal of enstatite. Another run at  $^{\circ}$C revealed large crystals of hydroxychondrodite in the chamber with the lowest water contents. Small differences in Mg:Si ratio could explain the differences in solid phases between the separate chambers, so it is assumed that the enstatite+forsterite and chondrodite+forsterite cotectics migrate very close to the forsterite+H$_2$O binary at high water contents and approximately 1200 $^{\circ}$C.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/fo-H2O}
  \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The position of the dry melting point is taken from \citep{PW1993}. \citep{LF2012}.}
  \label{fig:foH}
\end{figure}
\clearpage
\subsubsection{MgSiO$_3$-H$_2$O}

The melting of clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}. The lack of stishovite as a liquidus phase is in agreement with \cite{YII2004}, who reported a clinoenstatite + stishovite + liquid field appearing at 13.5 GPa. 
\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/en-H2O}
  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The position of the dry melting point is taken from \cite{PG1990}.}
  \label{fig:eoH}
\end{figure}

\clearpage
\subsubsection{SiO$_2$-H$_2$O}
The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}. The anhydrous melting temperature of coesite and the temperature of the coesite-stishovite transition at 13 GPa are taken from \cite{ZLGHF1993}. The metastable melting temperature of stishovite is taken from the melting curve of \cite{Millotetal2015}, which is consistent with the shock data of \cite{LAM1983} and static melting data of \cite{ZLGHF1993}.

The liquidus in this system provides quite severe constraints on thermodynamics in the anhydrous system. Given that SiO$_2$-H$_2$O liquids approach ideality with increasing pressure and temperature \citep{HM2012}, we feel justified in using the ideal mixing model of \cite{SS1985}. Using this model constrains the entropy of melting of stishovite to be between about 8 and 19 J/K/mol, not significantly higher than room pressure values \citep{HM2012,ZLGHF1993}. A value of 13.5 J/K/mol with G$_{OH}$=35000-20T J/mol yields the model shown in Figure  \ref{fig:SH}. Unfortunately, these entropy values of are much higher than the values estimated from the melting curve of SiO$_2$. The melting curve of coesite has a negligable dT/dP at 13 GPa \citep{ZLGHF1993}, constraining the $\Delta$V of melting to be approximately zero. Using the volume of coesite at the coesite-stishovite-liquid triple junction as a proxy for the volume of the melt, published melting curves of stishovite \citep{Millotetal2015, ZLGHF1993} imply an entropy of melting of $\sim$120 J/K/mol. There are two potential resolutions to this problem. The first is that the assumptions of the hydrous melting model are inappropriate for the high pressure SiO$_2$ system. The second is that the thermodynamic properties in the anhydrous system need to be revisited.

%NB: Most likely culprit the change in S with temperature?? Need a melt and solid model to test this.

% We use the formulation of \cite{HM2012} to model SiO$_2$-H$_2$O melts at 13 GPa. The volume of stishovite is taken from \cite{HP2011}. The melting curve of coesite at this pressure has $dT/dP \sim 0$ \citep{ZLGHF1993}, so the volume of anhydrous SiO$_2$ at this pressure is roughly equal to the volume of coesite via the Clausium-Clapeyron relation $dT/dP = \Delta V / \Delta S$. Using these volumes, the metastable entropy of melting of stishovite $\Delta S_{stv-L}$ is estimated as 8.5 J/K/mol. For completeness, we also calculate the melting point and entropy of melting of ice VII at 13 GPa from \emph{P-V-T} data collected along the melting curve \citep{FFH2004}.

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/SiO2-H2O}
  \caption{The silica-water phase diagram at 13 GPa, based on current experimental results. The liquidus curve is compared with previous experimentally-derived models computed at 1 GPa and 2 GPa \citep{HM2012}. An entropy of 13.5 J/K/mol is chosen to fit the data. Model fits use the model of \cite{SS1985}, with r=2 and K=$\infty$ (dotted grey), K=K(T) (black, expression in main text) and K=0 (dashed grey, all H2O as molecular).}
  \label{fig:SH}
\end{figure}

%\citep{SU2001}
%\citep{KOM2005}
%\citep{AA1980}
%\citep{Inoue1994}
%\citep{Wunder1998}
%\citep{AFRH2001}
%\citep{Nagano2002}
%\citep{MSUP2007}
%\citep{MFY2002}
%\citep{Sta2004}
%\citep{MFY2004}
%\citep{SUTG2001}

\clearpage
\subsection{The MgO-SiO$_2$-H$_2$O system}
The set of experimental run products can be used to create a preliminary liquidus diagram (Figure \ref{fig:liquidus}).

\cite{YII2004} suggested that there was a sharp break in the liquidus at the enstatite-stishovite cotectic (their Figure 6). However, this prediction breaks Schreinemakers Rules (specifically, that no one assemblage can occupy more than 180$^{\circ}$ around an invariant point).

\begin{figure}[ht!]
  \centering
      \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
  \caption{A preliminary liquidus diagram at 13 GPa, based on current experimental results.}
  \label{fig:liquidus}
\end{figure}

\subsection{Hydrous melting in high pressure subduction zone systems}
Subducting slabs are thought of transport significant quantities of mineral-bound water into the mantle transition zone. Much of this water is lost within the upper 200--300 km of the mantle through the breakdown of hydrous minerals such as antigorite and lawsonite. However, some mineral-bound water is thought to be preserved within a series of dense hydrous magnesium silicate (DHMS) phases. This observation was recently confirmed by the discovery of hydrous ringwoodite in diamond, which must have formed in the lower half of the mantle transition zone. The amount of water in the ringwoodite suggested that last equilibration occured under cold conditions probably associated with subduction. A key question, then, is how this material reached the surface sufficiently quickly to avoid retrogression to olivine, and how it remained cold during ascent to avoid the creation of hydrous melts. One effective way to allow melts to rapidly ascend to the surface is to channelise them. In turn, channelisation is promoted by reaction-infiltration and positive deformation-infiltration feedbacks.

The results of this study provide important constraints on the relationships between hydrous melt and the surrounding environment under the conditions at the top of the mantle transition zone. Within a single lithology, melts percolating upwards into hotter surroundings will consume small amounts of the surrounding solids, encouraging channelisation. At such high pressures, the change in fluid composition with temperature is gradual along the forsterite-enstatite and enstatite-stishovite cotectics, and the stishovite melting curve. Thus, it is unlikely that melt fraction will rapidly change in a single lithology.

A more profound effect occurs when hydrous fluids in equilibrium with peridotites infiltrate more Si-rich rocks. The enstatite-stishovite cotectic has a much greater $\partial T/\partial X_{H_2O}$ than the forsterite-enstatite cotectic at high temperatures. This implies that fluids entering silica-rich rocks will be corrosive, consuming enstatite and stishovite to reach equilibrium with their new surroundings. Despite being volumetrically minor components of subduction zone systems, silica-rich lithologies could play a dominant role in channelising hydrous fluids. The motion of these fluids could also significantly weaken the uppermost slab, effectively decoupling it from the overlying mantle wedge. Such effects promote trench parallel flow, encouraging slab deformation and rollback (trench migration relative to the surrounding mantle).

\clearpage
\section*{References}

\bibliography{references_MSH}

\end{document}

### 2015-07-07 MSH.tex
Update

@@
 
-\title{The compositions of high-pressure water-rich melts in the system MgO-SiO$_2$-H$_2$O}
+\title{Hydrous transition zone melting  
+%\title{High-pressure water-rich melting in the system MgO-SiO$_2$-H$_2$O: Insights into hydrous transition zone melting}
 %\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}
@@
 \begin{abstract}
-High pressure multianvil experiments were conducted at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O.
+
+In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O.
 Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source.
-The topology of the liquidus is tightly constrained at much higher water contents than previously possible.
-In the subsystem MgO-H$_2$O, the invariant point between brucite, periclase and liquid occurs at ca.
-1210 $^{\circ}$C and 67 mol\% H$_2$O.
-Forsterite and clinoenstatite both melt congruently in the temperature range investigated, although the fo-cen and cen-stv cotectics migrate close to the fo-H$_2$O and cen-H$_2$O binaries at high water contents.
-The chondrodite liquidus field approaches the fo-H$_2$O binary at ca.
-1200 $^{\circ}$C.
-Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O, then shallowing up to 50 mol\%, similar to behaviour previously observed at 2 GPa.
+As MgO:SiO$_2$ ratios decrease, the lquidus curve develops an increasingly pronounced concave-down topology.
+At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents.
+Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O.
+This melting point depression exceeds estimates from published models, and requires either significant non-ideality between species in the melt, or distinct differences in the ease of protonation of bridging oxygens.
+We discuss the implications of this work on the migration of deep-seated slab fluids, and the partitioning of water between nominally anhydrous phases and melts.
 \end{abstract}
@@
 \section{Introduction}
-The subduction of hydrous materials has a strong influence on the formation of melts within the deep Earth.
-The dehydration of hydrous crustal minerals is thought to drive the melting of the mantle wedge that feeds arc volcanoes.
-Dehydration of the underlying lithospheric mantle may also enable embrittlement of the slab, allowing it to deform at significantly higher strain rates than would be possible if it were dry.
+Subduction zones are primary vectors for water transport into the deep Earth.
+This water is progressively lost as fluids and then melts as the surrounding mantle warms the slab.
+The most prominent surface expression of this melting is the presence of arc volcanoes about 100 km above the slab surface, but hydrous minerals within the crust should allow some water to be transported to 250 km, and peridotite hydration should allow water to penetrate into the lowermost mantle transition zone, providing it remains low enough in temperature.
+A startling confirmation of such deep water is the discovery of water-saturated ringwoodite in a kimberlitic diamond from Juina, Brazil.
 
-There has been some controversy over whether slabs transport water beyond about 300 km depth.
+Several key questions present themselves.
+When water-rich melts are created in subduction zones, what are their compositions?
+How do their compositions and resulting physical properties (density, viscosity, melt-solid dihedral angle) change as they migrate?
+Where will they migrate?
+Their migration paths will have major implications for figuring out the water budget of the Earth - if they are trapped close to or within the slab, hydrating the whole of the mantle transition zone will be much less likely.
+
+There is another reason to be interested in the compositions of hydrous melts.
+The thermal stability of dense hydrous magnesium silicate (DHMS) phases and the partitioning of water between silicate melt and nominally anhydrous minerals (NAMS) control water transport into the deep mantle.
+At shallow depths, the fluids released during dehydration of these phases contain only a few percent of dissolved solutes.
+As such, it is reasonable to simplify thermodynamic calculations by approximating the fluid product of dehydration as pure water.
+As pressure increases, several processes ensure that this is no longer the case.
+Firstly, the second critical endpoint of peridotitic and eclogitic systems occur at <10 GPa, after which there is no longer a distinction between melt and fluid, and melt compositions smoothly migrate towards lower water contents with increasing temperature.
+Secondly, the thermal stability of DHMS phases increases.
+As a result, the final breakdown of DHMS phases at mantle transition zone pressures occurs when MgO-SiO$_2$-H$_2$O melts have H$_2$O activities significantly lower than one, which will also strongly influence the water content of NAMS.
+Accurate thermodynamic estimates of water storage capacities in mantle rocks thus require models for both the silicate phases and the coexisting melts.
+
+% There has been some controversy over whether slabs transport water beyond about 300 km depth.
 Certainly most of the ca.
@@
 
-The compositions of melts formed by the breakdown of hydrous phases is important for a full understanding of the deep water system.
+% The compositions of melts formed by the breakdown of hydrous phases is important for a full understanding of the deep water system.
 The thermodynamics of melting in water-rich systems cannot be determined without good estimates of melt compositions.
@@
 
-Although the phase relations in hydrous systems in the upper mantle are now remarkably well-constrained, constraining melt compositions has remained a major challenge in high-pressure experimental petrology.
+Constraining melt compositions remains a major challenge in high-pressure experimental petrology.
 The primary problem is that hydrous melts are unquenchable, which means that their compositions must be estimated indirectly.
@@
 Unfortunately, all three of these techniques are accompanied by significant uncertainties, especially in terms of the water content.
-An alternative technique is to constrain the liquidus temperature at a given composition.
-However, solid hydrous compounds do not reach the water concentrations expected from hydrous melts in the vicinity of a subducting slab.
+An alternative technique is to constrain the liquidus temperature at a given composition, thus circumventing the need to estimate water concentrations in the melt after the experiment (when much of the water has been lost).
+The problem with this technique is that at ambient pressures, hydrous phases are limited in their maximum water content.
+In the MgO-SiO$_2$ system, the lack of a thermally stable stoichiometric SiO$_2$.H$_2$O compound means that brucite has been used as the water source.
+This limits water concentrations in model peridotitic systems (Mg$\sim$1.4 Si) to $\sim$18 wt\% H$_2$O.
 Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form.
-Extrapolation is prone to large errors, especially when high temperatures facilitate water loss through capsule walls.
-Water masses can be accurately calculated when added to large volume chambers, but not to the small volumes of high pressure capsules.
-Thus, constraining the water contents of melts in hydrous systems remains a key problem in experimental petrology.
+Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible.
 
@@
 \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation.
-For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made purely of monomers.
-If r is less then the number of oxygens per formula unit no distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment.
+For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers.
+No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment.
 Fixing $r$ places an implicit constraint on the maximum value of $K$ in water-rich compositions.
-For example, if $r=2$ in a hydrous forsteritic melt, then water contents exceeding 50\% require that $K<\infty$.
+For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K<\infty$.
 In practise, the equilibrium constant $K$ has been shown to be dependent on temperature and composition.
@@
 
-In all of these models, the activity of component $x$ with the composition of the liquidus phase $X$ can be calculated from the melt model and independently from the entropy of melting of $X$.
-At equilibrium, these two activities must be equal:
+In these models, equilibrium melt compositions are calculated by calculating the activity of a component $x$ in the melt which has the composition of the anhydrous liquidus phase $X$ from the Gibbs energy of melting of $X$ at the temperature of interest:
 \begin{equation}
@@
 where $T_{melt}$ is the temperature of melting of the liquidus phase under anhydrous conditions.
+
+In this study, we use a modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} using melt endmembers MgO and SiO$_2$.
+For MgO melt, only the reference Helmholtz Free energy needed to be adjusted.
+For the SiO$_2$ melt, the bulk modulus of the SiO$_2$ endmember was increased to satisfy the experimental constraint that coexisting melt and coesite have similar volumes at 13--14 GPa.
+As a result, SiO$_2$ volumes at lower mantle conditions will be significantly higher than predicted, but this detail is not important for the present study.
+The volume at 1 bar is fit to experimental data.
+The entropy at reference conditions is adjusted by the value corresponding to the difference between the experimentally-derived and MD-calculated stishovite values.
+Remarkably, the value of the entropy derived in this way is essentially identical to that calculated for the experimental data of \cite{} near the coesite-stishovite-melt triple point.
+As such, we are confident that this value is accurate.
+It should be noted that the recent melting curve proposed by \cite{} (using a Simon's equation) predicts a much higher entropy of melting at 14 GPa.
+This is a result of downweighting the data in the study of \cite{}, which indicates a controversially high dT/dP.
+Nevertheless, this study is not inconsistent with higher pressure studies \citep{} if the gradient of the melting curve shallows between 30 and 70 GPa, which may be the consequence of changing Si-coordination and resultant densification in the melt.
+
+A subregular solution model was adopted to describe the free energy between the MgO and SiO$_2$ melts.
+The two interaction parameters W$_{MS}$ and W$_{SM}$ 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{} and melting temperatures at the fo, hen and fo-hen cotectic compositions.
+The temperature dependence around 14 GPa was calculated by assuming an ideal excess entropy contribution, and then fitting to the known melting curves of forsterite and enstatite.
+The pressure dependencies were fit to the melting curves.
+It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower.
+However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).
+
+All of the calculations in the work were undertaken with the mineral physics toolkit Burnman \cite{}.
 
@@
 They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa.
-We observed large brucite crystals at 1200 $^{\circ}$C and 13 GPa; so assuming that errors from both studies are $\sim$ 50 $^{\circ}$C, decomposition takes place just above 1200 $^{\circ}$C.
+We observed large brucite crystals at 1200 $^{\circ}$C and 13 GPa; the temperature difference is probably within experimental uncertainties.
 In the present study, the composition of the fluid coexisting with brucite and periclase is constrained to be 64--66 mol\% H$_2$O.
 The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase.
-The best experimental estimate of the melting point of MgO is currently that of \cite{ZF2008}, who extrapolated to the MgO endmember using melting phase relations in the system (Mg,Fe)O.
-Their melting curve is essentially identical to that obtained from large-scale molecular dynamics simulations \citep{CG1994}.
-Using these two studes, melting at 13 GPa is estimated to be $\sim 5373$ K.
-This value is higher than that from more recent ab-initio investigations \citep{Alfe2005,KKS2013}, and 2000 K higher than the melting temperature estimated experimentally by \cite{ZB1994}.
+The extremely high melting point of MgO means that the best experimental estimate of the melting point at high pressure (7 GPa) is derived by extrapolation from iron-bearing (Mg,Fe)O, using the compositions of coexisting melt and solid to constrain the shape of the two-phase region \citep{ZF2008}.
+Their melting curve is within error of that obtained from early molecular dynamics simulations \citep{CG1994}, which suggest a melting temperature of $\sim$ 5373 K at 13 GPa.
+In this study, we follow more recent MD simulations \citep{KKS2013, DKS2013} which predict periclase melting at a lower temperature of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}, but significantly higher than early experimental estimates \citep[$\sim$ 3100 K][]{ZB1994}.
 
-At 13 GPa, the entropy of melting of periclase is estimated to be $\sim$22 J/K/mol \cite{CG1994}.
-Assuming that the metastable entropy of melting at this pressure remains constant with temperature, the activity of periclase in the system can be determined and used to fit a melting model to the experimental data.
-Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -70000 - 15T$ provides a good fit to the data.
-However, this model yields a low activity of water in equilibrium with brucite and periclase ($a_{H_2O}\sim 0.34$) relative to that estimated using the thermodynamic data of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995} ($0.52$).
-
-To simultaneously fit the melting data and the activity of water in melts in equilibrium with periclase and brucite, $\partial^2 G_{ex} / \partial X_{H_2O}^2$ needs to be less positive at compositions more water-rich than brucite.
-This can be achieved by imposing a binary subregular model (W$_{MgO-H_2O}$=38000 J/mol, W$_{H_2O-MgO}$=0 J/mol).
+At 13 GPa, the calculated entropy of melting of periclase is $27.5$ J/K/mol, somewhat higher than the value from previous MD simulations \cite[$\sim$22 J/K/mol][]{CG1994}.
+This higher value will moderate the differences in melting temperature between the two studies.
+Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol).
+This asymmetry is required to fit the activity of water at the periclase-brucite-liquid triple point to the solid thermodynamic models of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995} ($0.52$).
 This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system.
 The second critical endpoint of the MgO-H$_2$O is probably around 13 GPa \citep{MSUP2007}, so large positive interaction enthalpies are justified in this region of the phase diagram.
-Alternatively, a regular solution model with MgO, Mg(OH)$_2$ and H$_2$O species in the melt can adequately fit the data, with parameters G$_{br,melt} = -41000.
-- 10*(T-(1210.+273.15))$, W$_{MgO-H_2O}=0$, W$_{MgO-Mg(OH)_2}=-10000$, W$_{Mg(OH)_2-H_2O}=-22000$ (all in J/mol).
-The resulting fits are shown in Figure \ref{fig:MH}.
-The large differences between the two model-types at high temperature is a result of the much larger entropy of mixing in the \cite{SS1985} models close to the anhydrous endmembers.
 
@@
 One chamber from an experiment run at 1300 $^{\circ}$C revealed a single very small crystal of enstatite.
-Another run at  $^{\circ}$C revealed large crystals of hydroxychondrodite in the chamber with the lowest water contents.
-Small differences in Mg:Si ratio could explain the differences in solid phases between the separate chambers, so it is assumed that the enstatite+forsterite and chondrodite+forsterite cotectics migrate very close to the forsterite+H$_2$O binary at high water contents and approximately 1200 $^{\circ}$C.
+Another run at 1200 $^{\circ}$C revealed large crystals of hydroxychondrodite in the chamber with the lowest water contents.
+Small differences in Mg:Si ratio could explain the differences in solid phases between the separate chambers, so it is assumed that the enstatite+forsterite and chondrodite+forsterite cotectics cross the forsterite+H$_2$O binary at high water contents and 1200--1300 $^{\circ}$C.
+This supposition is supported by \cite{DDFK2005}, who observe a double-crossing of the liquid and wadsleyite-water tie-line at 15 GPa, first at ca.
+1100 $^{\circ}$C, and again at 1400 $^{\circ}$C.
+Very similar Mg:Si ratios were observed by \citep{LSOK2011} at 15 and 18.5 GPa.
 
@@
 
-\subsection{Hydrous melting in high pressure subduction zone systems}
-Subducting slabs are thought of transport significant quantities of mineral-bound water into the mantle transition zone.
-Much of this water is lost within the upper 200--300 km of the mantle through the breakdown of hydrous minerals such as antigorite and lawsonite.
-However, some mineral-bound water is thought to be preserved within a series of dense hydrous magnesium silicate (DHMS) phases.
-This observation was recently confirmed by the discovery of hydrous ringwoodite in diamond, which must have formed in the lower half of the mantle transition zone.
-The amount of water in the ringwoodite suggested that last equilibration occured under cold conditions probably associated with subduction.
-A key question, then, is how this material reached the surface sufficiently quickly to avoid retrogression to olivine, and how it remained cold during ascent to avoid the creation of hydrous melts.
-One effective way to allow melts to rapidly ascend to the surface is to channelise them.
-In turn, channelisation is promoted by reaction-infiltration and positive deformation-infiltration feedbacks.
+\section{Discussion}
+\subsection{Water partitioning in the mantle transition zone}
+% Olivine, wadsleyite and ringwoodite
 
-The results of this study provide important constraints on the relationships between hydrous melt and the surrounding environment under the conditions at the top of the mantle transition zone.
-Within a single lithology, melts percolating upwards into hotter surroundings will consume small amounts of the surrounding solids, encouraging channelisation.
-At such high pressures, the change in fluid composition with temperature is gradual along the forsterite-enstatite and enstatite-stishovite cotectics, and the stishovite melting curve.
-Thus, it is unlikely that melt fraction will rapidly change in a single lithology.
+\cite{DDFK2005}
+\citep{LSOK2011}
+\citep{OMY2000}
 
-A more profound effect occurs when hydrous fluids in equilibrium with peridotites infiltrate more Si-rich rocks.
-The enstatite-stishovite cotectic has a much greater $\partial T/\partial X_{H_2O}$ than the forsterite-enstatite cotectic at high temperatures.
-This implies that fluids entering silica-rich rocks will be corrosive, consuming enstatite and stishovite to reach equilibrium with their new surroundings.
-Despite being volumetrically minor components of subduction zone systems, silica-rich lithologies could play a dominant role in channelising hydrous fluids.
-The motion of these fluids could also significantly weaken the uppermost slab, effectively decoupling it from the overlying mantle wedge.
-Such effects promote trench parallel flow, encouraging slab deformation and rollback (trench migration relative to the surrounding mantle).
+\subsection{Deep water release from subduction zones}
+% Densities
+\cite{MF2010}
+\cite{BK2003}
+\cite{MSK2008}
+
+% Viscosities
+% Shear viscosity 
+% Density
+
+%So at least in my models there are three quantities that are important for the pattern of melt migration, the permeability, the compaction viscosity and the shear viscosity (if you have elasticity/plasticity, other things come in, too).
+Everywhere in the equations where you have the permeability, it's divided by the melt viscosity, and the permeability itself is normally assumed to scale with phi^3 (or less common phi^2), and at least I didn't see any parametrisations describing the effect of temperature on the permeability, so I would assume it's small.
+So for the permeability, you can just use the values you get for your different P-T-X conditions, and see if the overall term permeability/melt_viscosity increases or decreases.
+
+%For the viscosities of the solid you have the influence of temperature on the viscosity (which I think is similar for shear and compaction viscosity) and you have the effect of porosity, eta ~ exp(alpa * phi) with phi~-27 for the shear viscosity (I think this is what Rich Katz uses in his models, but I would have to look it up to be sure), and xi ~ phi0/phi for the compaction viscosity, and you can do the same here, compare how they change for your different P-T-X conditions.
 

### 2015-07-07 MSH.tex
Add

@@
 
-\title{Hydrous transition zone melting  
-%\title{High-pressure water-rich melting in the system MgO-SiO$_2$-H$_2$O: Insights into hydrous transition zone melting}
+\title{Hydrous melting and partitioning in the deep upper mantle: insights from experiments in the system MgO-SiO$_2$-H$_2$O}
 %\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}
@@
 Constraining melt compositions remains a major challenge in high-pressure experimental petrology.
-The primary problem is that hydrous melts are unquenchable, which means that their compositions must be estimated indirectly.
-Compositions of partial hydrous melts have been estimated using mass balance, diamond traps \citep{MSUP2007}, and the Mg:Si ratio has been estimated by defocusing the beam of an electron probe microanalyser (EPMA) \citep[e.g.][]{YII2004}.
-Unfortunately, all three of these techniques are accompanied by significant uncertainties, especially in terms of the water content.
-An alternative technique is to constrain the liquidus temperature at a given composition, thus circumventing the need to estimate water concentrations in the melt after the experiment (when much of the water has been lost).
+The primary problem is that hydrous melts are unquenchable, which means that their water concentrations must be estimated indirectly.
+Compositions of partial hydrous melts have been estimated using a mixture of mass balance and defocused-beam electron probe measurements, which to provide Mg:Si ratios and a rough estimate of water content from deficits in totals.
+Such measurements are conducted on melt pockets \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011} or melts isolated within diamond traps \citep{MSUP2007}.
+Unfortunately, these techniques are accompanied by significant uncertainties, especially in terms of water content.
+An alternative technique is to constrain the liquidus temperature at a given composition, thus circumventing the need to estimate water concentrations in the melt \emph{ex-situ} (when much of the water has been lost).
 The problem with this technique is that at ambient pressures, hydrous phases are limited in their maximum water content.
-In the MgO-SiO$_2$ system, the lack of a thermally stable stoichiometric SiO$_2$.H$_2$O compound means that brucite has been used as the water source.
-This limits water concentrations in model peridotitic systems (Mg$\sim$1.4 Si) to $\sim$18 wt\% H$_2$O.
+For example, in the MgO-SiO$_2$ system, maximum water contents in a solid starting material are provided by a mixture of brucite and quartz.
 Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form.
@@
 
-Capsules were created from 2 mm-diameter Pt$_{90}$Rh$_{10}$ and Au rods.
-They were cut by wire saw into 1 mm thick disks, into which were spark-eroded six holes in three rows, 250 microns wide and 700 microns deep.
+Capsules were created from 2 mm-diameter Pt$_{90}$Rh$_{10}$ and Au rods, cut by wire saw into 1 mm thick disks.
+Into each disk were spark-eroded two rows of three holes, each 250 microns in diameter and 700 microns deep.
 The capsules were cleaned by cycling between an acetone ultrasonic bath (15 minutes) and 1000 $^{\circ}$C furnace (20 minutes) three times.
-Any remaining contamination was removed with W$_{75}$Re$_{25}$ needle followed by a further trip to the ultrasonic bath.
+Any remaining contamination was removed with a W$_{75}$Re$_{25}$ needle followed by a further trip to the ultrasonic bath.
 Capsule chambers were filled with powders of different compositions using a W$_{75}$Re$_{25}$ needle.
-Small pieces of tape were used to cover the other holes to avoid contamination.
+Small pieces of tape were used to temporarily cover the other holes to avoid contamination.
 
@@
 
-\citep{NM2008}
-\citep{DM2010}
-\citep{ZK1998}
-
 \subsection{Subsystems}
@@
 %\citep{AA1980}
-%\citep{Inoue1994}
 %\citep{Wunder1998}
@@
 %\citep{SUTG2001}
+%\citep{NM2008}
+%\citep{DM2010}
+%\citep{ZK1998}
 
@@
 
-\cite{DDFK2005}
-\citep{LSOK2011}
-\citep{OMY2000}
+Partitioning of hydrogen between mineral and melt phases can be described using a partition coefficient D:
+\begin{equation}
+  D^{a/b}_{H_2O} = c^a_{H_2O} / c^b_{H_2O}
+\end{equation}
+where $c^i_{H_2O}$ is the concentration of H$_2$O (in wt\%) in phase $i$ \citep{KB2006}.
+In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA.
+In the case of melts, which are unquenchable, $c$ in partitioning studies has most commonly been estimated by mass balance or deficit in microprobe totals.
+In two studies on Fe-free wadsleyite, \cite{DDFK2005} and  \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits.
+A similar study on ringwoodite was undertaken by \cite{OMY2000}.
+One striking observation is the extreme differences in water content estimated for melts in equilibrium with wadsleyite at 15 GPa and ringwoodite at 17-20 GPa.
+For example, melt in equilibrium with wadsleyite at 1400 C was estimated to contain 10.6-13.3 wt wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite suggest water contents of 57-66 wt\% H$_2$O.
+Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs.
+97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate.
+We note that two of the three values provided by \cite{OMY2000} fall above the temperatures indicating that H$_2$O is the only H-bearing species in the melt \citep{SS1985}, and that the melting temperatures of wadsleyite + 20.4 wt\% H$_2$O (40 mol\% H$_2$O in Figure \ref{fig:partitioning}) in the wadsleyite studies fall below that of the ame composition at 5.5 GPa \citep{Inoue1994}.
+Both of these interpretations appear unlikely.
+
+Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that the thermodynamic properties of the melt remain essentially constant between 13 and 20 GPa.
+Water concentrations in the solid are taken from the original studies.
+We restrict our analysis to data from 1100--1400 C, where the Mg:Si ratios of the melts are similar to that in forsterite.
+The average $D^{wad/melt}_{H_2O}$ (0.44) and $D^{rw/melt}_{H_2O}$ values (0.6) produce the lines in Figure \ref{fig:partitioning}.
+Note that $D^{wad/rw}_{H_2O}$ = 0.72, implying that at equilibrium, the lower transition zone should contain more water than the upper transition zone; the opposite conclusion to that reached by \cite{DDFK2005} using the same data XXXX check this!!.
+This prediction could explain observations of low velocity layers at ca.
+520 km depth \citep{JDH2010}, in an extension of the water filter model of \cite{BK2003}.
+
+An accurate thermodynamic model for hydrous melts can also reveal the mechanism of hydrogen incorporation into nominally anhydrous minerals.
+This arises from the equilibrium constant for the reaction between oxygen in the silicate, water in the melt, and OH groups in the solid: XXXX correct this!!
+\begin{equation}
+  K = \frac{a_{OH}^n}{a_{H_2O}^{n/2}a_O^{n/2}}
+\end{equation}
+
+Isolated protons (n=0.5), proton pairs (n=1) and proton . Making the approximation that $a_O = 1$, which is probably reasonable at low H concentrations in the solid 
+Making the assumption that concentration of H$_2$O in the solid is proportional to the activity of OH in the melt, we therefore have
+\begin{equation}
+  c \propto a_{H_2O}^n
+\end{equation}
+where $n$ relates to the mechanisms of hydrogen incorporation into the solid \citep{KB2006}
 

### 2015-07-08 MSH.tex
Update

@@
 The average $D^{wad/melt}_{H_2O}$ (0.44) and $D^{rw/melt}_{H_2O}$ values (0.6) produce the lines in Figure \ref{fig:partitioning}.
-Note that $D^{wad/rw}_{H_2O}$ = 0.72, implying that at equilibrium, the lower transition zone should contain more water than the upper transition zone; the opposite conclusion to that reached by \cite{DDFK2005} using the same data XXXX check this!!.
+Note that $D^{wad/rw}_{H_2O}$ = 0.72, implying that at equilibrium, the lower transition zone should contain more water than the upper transition zone.
 This prediction could explain observations of low velocity layers at ca.
 520 km depth \citep{JDH2010}, in an extension of the water filter model of \cite{BK2003}.
+We also note that there is currently no strong evidence to argue for repartitioning of water between the upper and lower transition zone due to the secular cooling of the Earth as proposed by \cite{DDFK2005}.
 
 An accurate thermodynamic model for hydrous melts can also reveal the mechanism of hydrogen incorporation into nominally anhydrous minerals.
-This arises from the equilibrium constant for the reaction between oxygen in the silicate, water in the melt, and OH groups in the solid: XXXX correct this!!
+This arises from the equilibrium constant for the reaction between oxygen in the silicate, water in the melt, and OH groups in the solid:
 \begin{equation}
-  K = \frac{a_{OH}^n}{a_{H_2O}^{n/2}a_O^{n/2}}
+  K = \frac{a_{(OH)_{2n}}}{a_{H_2O}^{n}a_O^{n}}
 \end{equation}
-
-Isolated protons (n=0.5), proton pairs (n=1) and proton . Making the approximation that $a_O = 1$, which is probably reasonable at low H concentrations in the solid 
-Making the assumption that concentration of H$_2$O in the solid is proportional to the activity of OH in the melt, we therefore have
+The value of n depends on the H$^+$ substitution mechanism: isolated protonation yields n=0.5, proton pairs n=1 and 4-proton clusters n=2.
+Making the approximation that $a_O$ is constant, that the concentration of OH in the solid is proportional to the activity of (OH)$_{2n}$, and that the equilibrium constant has the usual Arrhenius form \citep{KB2006}, we therefore have
 \begin{equation}
-  c \propto a_{H_2O}^n
+  c = A a_{H_2O}^{n} \exp \left( -\frac{\Delta H^{1 bar} + \Delta V^{solids} P}{R T}\right)
 \end{equation}
-where $n$ relates to the mechanisms of hydrogen incorporation into the solid \citep{KB2006}
 

### 2015-07-09 MSH.tex
Update

--- 2015-07-07 MSH.tex
@@
 
-\title{Hydrous melting and partitioning in the deep upper mantle: insights from experiments in the system MgO-SiO$_2$-H$_2$O}
+\title{Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments}
 %\tnotetext[mytitlenote]{Fully documented templates are available in the elsarticle package on \href{http://www.ctan.org/tex-archive/macros/latex/contrib/elsarticle}{CTAN}.}
@@
 Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source.
-As MgO:SiO$_2$ ratios decrease, the lquidus curve develops an increasingly pronounced concave-down topology.
+As MgO:SiO$_2$ ratios decrease, the liquidus curve develops an increasingly pronounced concave-down topology.
 At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents.
@@
 \section{Introduction}
-Subduction zones are primary vectors for water transport into the deep Earth.
-This water is progressively lost as fluids and then melts as the surrounding mantle warms the slab.
-The most prominent surface expression of this melting is the presence of arc volcanoes about 100 km above the slab surface, but hydrous minerals within the crust should allow some water to be transported to 250 km, and peridotite hydration should allow water to penetrate into the lowermost mantle transition zone, providing it remains low enough in temperature.
-A startling confirmation of such deep water is the discovery of water-saturated ringwoodite in a kimberlitic diamond from Juina, Brazil.
+Hydrous melts have a remarkable influence on Earth's surface and interior.
+Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes.
+The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}.
+This deep water cycle is probably the source for some kimberlitic eruptions, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}.
+The ringwoodite stability field requires that the kimberlite must have been sourced from at least ca.
+500 km depth.
+In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage.
+It has also been argued that neutral or negatively buoyant water-rich melts are to blame for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.
 
-Several key questions present themselves.
-When water-rich melts are created in subduction zones, what are their compositions?
-How do their compositions and resulting physical properties (density, viscosity, melt-solid dihedral angle) change as they migrate?
-Where will they migrate?
-Their migration paths will have major implications for figuring out the water budget of the Earth - if they are trapped close to or within the slab, hydrating the whole of the mantle transition zone will be much less likely.
-
-There is another reason to be interested in the compositions of hydrous melts.
-The thermal stability of dense hydrous magnesium silicate (DHMS) phases and the partitioning of water between silicate melt and nominally anhydrous minerals (NAMS) control water transport into the deep mantle.
-At shallow depths, the fluids released during dehydration of these phases contain only a few percent of dissolved solutes.
-As such, it is reasonable to simplify thermodynamic calculations by approximating the fluid product of dehydration as pure water.
-As pressure increases, several processes ensure that this is no longer the case.
-Firstly, the second critical endpoint of peridotitic and eclogitic systems occur at <10 GPa, after which there is no longer a distinction between melt and fluid, and melt compositions smoothly migrate towards lower water contents with increasing temperature.
-Secondly, the thermal stability of DHMS phases increases.
-As a result, the final breakdown of DHMS phases at mantle transition zone pressures occurs when MgO-SiO$_2$-H$_2$O melts have H$_2$O activities significantly lower than one, which will also strongly influence the water content of NAMS.
-Accurate thermodynamic estimates of water storage capacities in mantle rocks thus require models for both the silicate phases and the coexisting melts.
-
-% There has been some controversy over whether slabs transport water beyond about 300 km depth.
-Certainly most of the ca.
-7 km thick mafic crust is probably too warm to sustain hydrous minerals.
-However, hydrated ultramafics can potentially carry water into the lower mantle.
-Supporting evidence for an ultradeep part of the water cycle exists in the form of hydrous ringwoodite in diamond, which apparently formed under cold conditions.
-
-% The compositions of melts formed by the breakdown of hydrous phases is important for a full understanding of the deep water system.
-The thermodynamics of melting in water-rich systems cannot be determined without good estimates of melt compositions.
-Furthermore, the thermal stability of hydrous phases cannot be predicted from their properties without also having an activity-composition model for the melt phase.
-More generally, the density of hydrous melts will be highly composition dependent, impacting on their ability to rise through the mantle.
-Finally, reactions between migrating melts and the surrounding solid will be controlled by the \emph{P-T-X} dependence on melt composition.
+Processes involving deep-seated hydrous melts are controlled to a great extent by melt composition.
+Composition is a key variable in determining melt density, viscosity and solid-solid-melt dihedral angles, which together with the temperature and equivalent properties of the solid control the time- and lengthscales on which melt can separate from the melt.
+They also govern the extent to which this melt can become channelised, reducing interaction with its surroundings and increasing ascent rates.
+Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure.
+Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth.
+Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O.
+As pressure increases, immiscibility between fluid and melt decreases, and the thermal stability of hydrous phases increases.
+As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one.
+Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.
 
@@
 
-In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$) as a novel water source.
+In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$), as a novel water source.
 Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}.
@@
 \section{Experimental and analytical methods}
-Starting compositions were created from a mixture of high purity brucite (Mg(OH)$_2$, SiO$_2$ and platinic acid (H$_2$Pt(OH)$_6$).
-SiO$_2$ was dried overnight at 1000 $^{\circ}$C, while brucite was heated to 250 $^{\circ}$C overnight.
+Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (H$_2$Pt(OH)$_6$).
+Before weighing, quartz was dried at 1000 $^{\circ}$C for 12 hours, while brucite was heated to 250 $^{\circ}$C, also over 12 hours.
 Both powders were stored in a desiccating oven at 130 $^{\circ}$C.
@@
 
+\section{Modelling water solubility in melts}
+
+Oxygen in MgO-SiO$_2$-H$_2$O melts can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group.
+The general equation describing the reaction between melt species is
+\begin{equation}
+0.5 \textrm{H}_2\textrm{O} + 0.5 \textrm{O}_{br}^{2-} \rightleftharpoons \textrm{OH}_{tm}^-
+\label{eqn:speciation}
+\end{equation}
+
+where the subscripts br and tm respectively refer to bridging and terminal oxygens. The equilibrium between these species can be described with an equilibrium constant $K$ \cite[e.g.][]{Stolper1982}
+\begin{equation}
+K = \frac{X_{\textrm{OH}_{tm}^-}}{\sqrt{X_{\textrm{H}_2\textrm{O}} X_{\textrm{O}_{br}^{2-}} } }
+\label{eqn:equilibrium_constant}
+\end{equation}
+
+The equilibrium constant is related to the energy $\Delta G^{\circ}$ required to protonate a bridging oxygen
+\begin{equation}
+K = \exp\left(\frac{-\Delta G^{\circ}}{RT}\right)
+\end{equation}
+
+These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers.
+A high proportion of monomers and dimers exist in relatively dilute solutions, but not in the concentrated solutions investigated in this study.
+
+A number of models have been presented to investigate the behaviour of concentrated hydrous melts.
+\cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation.
+For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers.
+No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment.
+Fixing $r$ places an implicit constraint on the maximum value of $K$ in water-rich compositions.
+For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K<\infty$.
+In practise, the equilibrium constant $K$ has been shown to be dependent on temperature and composition.
+\cite{SK1995} obtained a good fit to isochemical data with an expression of the form
+\begin{equation}
+\Delta G^{\circ} = a + b\,T
+\end{equation}
+
+\cite{THH2012} simplified the model of \cite{SS1985} by assuming that all hydrogen in the melt exists as OH$^-$, and that protonation is equally likely on any oxygen ($K=\infty$, $\Delta G^{\circ}=-\infty$, $r$ maximised).
+Deviations from this model must increase with increasing water content, and such a model cannot describe any melt more water-rich than the Mg(OH)$_2$-Si(OH)$_4$ binary.
+
+To describe melting in the SiO$_2$-H$_2$O system, \cite{HM2012} generalised the aforementioned models by allowing non-ideality between the two endmembers.
+The excess term $\Delta G^{\circ}$(P,T) for the formation of OH$^-$ groups was assumed to be independent of composition; all oxygens are equally available for protonation, and the local extent of protonation does not affect the energy required for each further protonation.
+
+In these models, equilibrium melt compositions are calculated by calculating the activity of a component $x$ in the melt which has the composition of the anhydrous liquidus phase $X$ from the Gibbs energy of melting of $X$ at the temperature of interest:
+\begin{equation}
+RT \ln a_{x,L}^X = G_{X_l} - G_{X_{s}} = - \int_T^{T_{melt}} S_{X_{l}}-S_{X_s} \, dT
+\end{equation}
+where $T_{melt}$ is the temperature of melting of the liquidus phase under anhydrous conditions.
+
+In this study, we use a modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} using melt endmembers MgO and SiO$_2$.
+For MgO melt, only the reference Helmholtz free energy needed to be adjusted.
+We note that the melting point of MgO at high pressure is rather uncertain.
+In the present model \citep{DKS2013}, MgO melts at $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}.
+Experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}.
+We note that if the present model is incorrect, it is probably the calculated entropy of melting which is at fault.
+A higher melting temperature would therefore imply a lower entropy of melting, such that the values in $G_{\textit{melt}}$ - $G_{\textit{solid}}$ at low temperatures will have smaller errors, than would otherwise be expected from such large melting point uncertainties.
+For the SiO$_2$ melt, the bulk modulus of the SiO$_2$ endmember was increased to satisfy the experimental constraint that coexisting melt and coesite have similar volumes at 13--14 GPa \citep{ZLGHF1993}.
+The volume at 1 bar is fit to experimental data.
+The entropy of melting at the coesite-stishovite-melt triple point ($\sim$40 J/K/mol) was kept the same.
+This value is identical to the value derived from the Clapeyron slope of the stishovite liquidus \cite{SL1995}.
+We note that this slope is much steeper than that recently proposed by \cite{Millotetal2015}.
+This difference is not the result of incompatibility between experimental results at different pressures \citep{SL1995, LAM1983, Millotetal2015}, but instead the result of using a 2-parameter Simon's equation to fit the rather abrupt change in dT/dP implied by the experiments at 20--40 GPa.
+
+A subregular solution model was adopted to describe the free energy between the MgO and SiO$_2$ melts.
+The two interaction parameters W$_{MS}$ and W$_{SM}$ at 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions.
+The temperature dependence around 14 GPa was calculated by assuming an ideal excess entropy contribution, and then fitting to the known melting curves of forsterite and enstatite.
+The pressure dependencies were fit to the melting curves.
+It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower.
+However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).
+
+The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/}][]{CHRU2014}.
+
+\clearpage
 \section{Results}
+\subsection{Ex-situ observations}
 Experimental details and results are shown in Table \ref{table:experiments}.
@@
 
-\section{Discussion}
-\subsection{Modelling water solubility in melts}
-
-Oxygen in MgO-SiO$_2$-H$_2$O melts can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group.
-The general equation describing the reaction between melt species is
-\begin{equation}
-0.5 \textrm{H}_2\textrm{O} + 0.5 \textrm{O}_{br}^{2-} \rightleftharpoons \textrm{OH}_{tm}^-
-\label{eqn:speciation}
-\end{equation}
-
-where the subscripts br and tm respectively refer to bridging and terminal oxygens. The equilibrium between these species can be described with an equilibrium constant $K$ \cite[e.g.][]{Stolper1982}
-\begin{equation}
-K = \frac{X_{\textrm{OH}_{tm}^-}}{\sqrt{X_{\textrm{H}_2\textrm{O}} X_{\textrm{O}_{br}^{2-}} } }
-\label{eqn:equilibrium_constant}
-\end{equation}
-
-The equilibrium constant is related to the energy $\Delta G^{\circ}$ required to protonate a bridging oxygen
-\begin{equation}
-K = \exp\left(\frac{-\Delta G^{\circ}}{RT}\right)
-\end{equation}
-
-These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers.
-A high proportion of monomers and dimers exist in relatively dilute solutions, but not in the concentrated solutions investigated in this study.
-
-A number of models have been presented to investigate the behaviour of concentrated hydrous melts.
-\cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation.
-For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers.
-No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment.
-Fixing $r$ places an implicit constraint on the maximum value of $K$ in water-rich compositions.
-For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K<\infty$.
-In practise, the equilibrium constant $K$ has been shown to be dependent on temperature and composition.
-\cite{SK1995} obtained a good fit to isochemical data with an expression of the form
-\begin{equation}
-\Delta G^{\circ} = a + b\,T
-\end{equation}
-
-\cite{THH2012} simplified the model of \cite{SS1985} by assuming that all hydrogen in the melt exists as OH$^-$, and that protonation is equally likely on any oxygen ($K=\infty$, $\Delta G^{\circ}=-\infty$, $r$ maximised).
-Deviations from this model must increase with increasing water content, and such a model cannot describe any melt more water-rich than the Mg(OH)$_2$-Si(OH)$_4$ binary.
-
-To describe melting in the SiO$_2$-H$_2$O system, \cite{HM2012} generalised the aforementioned models by allowing non-ideality between the two endmembers.
-The excess term $\Delta G^{\circ}$(P,T) for the formation of OH$^-$ groups was assumed to be independent of composition; all oxygens are equally available for protonation, and the local extent of protonation does not affect the energy required for each further protonation.
-
-In these models, equilibrium melt compositions are calculated by calculating the activity of a component $x$ in the melt which has the composition of the anhydrous liquidus phase $X$ from the Gibbs energy of melting of $X$ at the temperature of interest:
-\begin{equation}
-RT \ln a_{x,L}^X = G_{X_l} - G_{X_{s}} = - \int_T^{T_{melt}} S_{X_{l}}-S_{X_s} \, dT
-\end{equation}
-where $T_{melt}$ is the temperature of melting of the liquidus phase under anhydrous conditions.
-
-In this study, we use a modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} using melt endmembers MgO and SiO$_2$.
-For MgO melt, only the reference Helmholtz Free energy needed to be adjusted.
-For the SiO$_2$ melt, the bulk modulus of the SiO$_2$ endmember was increased to satisfy the experimental constraint that coexisting melt and coesite have similar volumes at 13--14 GPa.
-As a result, SiO$_2$ volumes at lower mantle conditions will be significantly higher than predicted, but this detail is not important for the present study.
-The volume at 1 bar is fit to experimental data.
-The entropy at reference conditions is adjusted by the value corresponding to the difference between the experimentally-derived and MD-calculated stishovite values.
-Remarkably, the value of the entropy derived in this way is essentially identical to that calculated for the experimental data of \cite{} near the coesite-stishovite-melt triple point.
-As such, we are confident that this value is accurate.
-It should be noted that the recent melting curve proposed by \cite{} (using a Simon's equation) predicts a much higher entropy of melting at 14 GPa.
-This is a result of downweighting the data in the study of \cite{}, which indicates a controversially high dT/dP.
-Nevertheless, this study is not inconsistent with higher pressure studies \citep{} if the gradient of the melting curve shallows between 30 and 70 GPa, which may be the consequence of changing Si-coordination and resultant densification in the melt.
-
-A subregular solution model was adopted to describe the free energy between the MgO and SiO$_2$ melts.
-The two interaction parameters W$_{MS}$ and W$_{SM}$ 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{} and melting temperatures at the fo, hen and fo-hen cotectic compositions.
-The temperature dependence around 14 GPa was calculated by assuming an ideal excess entropy contribution, and then fitting to the known melting curves of forsterite and enstatite.
-The pressure dependencies were fit to the melting curves.
-It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower.
-However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).
-
-All of the calculations in the work were undertaken with the mineral physics toolkit Burnman \cite{}.
-
 \subsection{Subsystems}
@@
 The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase.
-The extremely high melting point of MgO means that the best experimental estimate of the melting point at high pressure (7 GPa) is derived by extrapolation from iron-bearing (Mg,Fe)O, using the compositions of coexisting melt and solid to constrain the shape of the two-phase region \citep{ZF2008}.
-Their melting curve is within error of that obtained from early molecular dynamics simulations \citep{CG1994}, which suggest a melting temperature of $\sim$ 5373 K at 13 GPa.
-In this study, we follow more recent MD simulations \citep{KKS2013, DKS2013} which predict periclase melting at a lower temperature of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}, but significantly higher than early experimental estimates \citep[$\sim$ 3100 K][]{ZB1994}.
 
-At 13 GPa, the calculated entropy of melting of periclase is $27.5$ J/K/mol, somewhat higher than the value from previous MD simulations \cite[$\sim$22 J/K/mol][]{CG1994}.
-This higher value will moderate the differences in melting temperature between the two studies.
 Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol).
@@
 The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}.
-The anhydrous melting temperature of coesite and the temperature of the coesite-stishovite transition at 13 GPa are taken from \cite{ZLGHF1993}.
-The metastable melting temperature of stishovite is taken from the melting curve of \cite{Millotetal2015}, which is consistent with the shock data of \cite{LAM1983} and static melting data of \cite{ZLGHF1993}.
-
-The liquidus in this system provides quite severe constraints on thermodynamics in the anhydrous system.
-Given that SiO$_2$-H$_2$O liquids approach ideality with increasing pressure and temperature \citep{HM2012}, we feel justified in using the ideal mixing model of \cite{SS1985}.
-Using this model constrains the entropy of melting of stishovite to be between about 8 and 19 J/K/mol, not significantly higher than room pressure values \citep{HM2012,ZLGHF1993}.
-A value of 13.5 J/K/mol with G$_{OH}$=35000-20T J/mol yields the model shown in Figure  \ref{fig:SH}.
-Unfortunately, these entropy values of are much higher than the values estimated from the melting curve of SiO$_2$.
-The melting curve of coesite has a negligable dT/dP at 13 GPa \citep{ZLGHF1993}, constraining the $\Delta$V of melting to be approximately zero.
-Using the volume of coesite at the coesite-stishovite-liquid triple junction as a proxy for the volume of the melt, published melting curves of stishovite \citep{Millotetal2015, ZLGHF1993} imply an entropy of melting of $\sim$120 J/K/mol.
-There are two potential resolutions to this problem.
-The first is that the assumptions of the hydrous melting model are inappropriate for the high pressure SiO$_2$ system.
-The second is that the thermodynamic properties in the anhydrous system need to be revisited.
 
@@
 \section{Discussion}
+\subsection{Melting in the MgO-SiO$_2$-H$_2$O system}
 \subsection{Water partitioning in the mantle transition zone}
@@
 In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA.
-In the case of melts, which are unquenchable, $c$ in partitioning studies has most commonly been estimated by mass balance or deficit in microprobe totals.
+In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to hydrogen.
 In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits.
 A similar study on ringwoodite was undertaken by \cite{OMY2000}.
-One striking observation is the extreme differences in water content estimated for melts in equilibrium with wadsleyite at 15 GPa and ringwoodite at 17-20 GPa.
-For example, melt in equilibrium with wadsleyite at 1400 C was estimated to contain 10.6-13.3 wt wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite suggest water contents of 57-66 wt\% H$_2$O.
+The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:partitioning}).
+For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O.
 Modelled Mg$_2$SiO$_4$ melting temperatures increase by only 40 K between 15 and 20 GPa, and entropies of melting at these two pressures are not very different (92.4 vs.
 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate.
-We note that two of the three values provided by \cite{OMY2000} fall above the temperatures indicating that H$_2$O is the only H-bearing species in the melt \citep{SS1985}, and that the melting temperatures of wadsleyite + 20.4 wt\% H$_2$O (40 mol\% H$_2$O in Figure \ref{fig:partitioning}) in the wadsleyite studies fall below that of the ame composition at 5.5 GPa \citep{Inoue1994}.
+We note that two of the three values provided by \cite{OMY2000} fall above the temperatures indicating that H$_2$O is the only H-bearing species in the melt \citep{SS1985}, and that the melting temperatures of wadsleyite + 20.4 wt\% H$_2$O (X$_{H_2O}$=0.4 in Figure \ref{fig:partitioning}) in the wadsleyite studies fall below that of the same composition at 5.5 GPa \citep{Inoue1994}.
 Both of these interpretations appear unlikely.
@@
 Water concentrations in the solid are taken from the original studies.
-We restrict our analysis to data from 1100--1400 C, where the Mg:Si ratios of the melts are similar to that in forsterite.
-The average $D^{wad/melt}_{H_2O}$ (0.44) and $D^{rw/melt}_{H_2O}$ values (0.6) produce the lines in Figure \ref{fig:partitioning}.
-Note that $D^{wad/rw}_{H_2O}$ = 0.72, implying that at equilibrium, the lower transition zone should contain more water than the upper transition zone; the opposite conclusion to that reached by \cite{DDFK2005} using the same data XXXX check this!!.
-This prediction could explain observations of low velocity layers at ca.
-520 km depth \citep{JDH2010}, in an extension of the water filter model of \cite{BK2003}.
+We restrict our analysis to data from 1100--1400 $^{\circ}$C, where the Mg:Si ratios of the melts are similar to that in forsterite.
+At 15 GPa, $D^{wad/melt}_{H_2O}$ increases from 0.30 to 0.51 between 900 and 1200 $^{\circ}$C, and then decreases back to 0.32 at 1400 $^{\circ}$C.
+The three ringwoodite H$_2$O concentrations (20-23 GPa) indicate a decrease in $D^{rw/melt}_{H_2O}$ from 0.68 to 0.57 between 1300 and ca.
+1400 $^{\circ}$C.
 
-An accurate thermodynamic model for hydrous melts can also reveal the mechanism of hydrogen incorporation into nominally anhydrous minerals.
-This arises from the equilibrium constant for the reaction between oxygen in the silicate, water in the melt, and OH groups in the solid: XXXX correct this!!
+It is also possible to create a thermodynamic model for incorporation of H$_2$O into wadsleyite and ringwoodite, given some knowledge of the substitution mechanism.
+Hydrogen in wadsleyite and ringwoodite is incorporated as proton pairs substituting for Mg (Mg$^{2+}$ $\leftarrow$ 2H$^{+}$) \citep{Smyth1987, SHFJLM2003}.
+Ringwoodite potentially also exhibits a substitution involving Si (Si$^{4+}$ $\leftarrow$ Mg$^{2+}$ + 2H$^{+}$) \citep{KKMO2000}.
+Both of these reactions may be written in the form
 \begin{equation}
-  K = \frac{a_{OH}^n}{a_{H_2O}^{n/2}a_O^{n/2}}
+H_2O + O = (OH)_2
 \end{equation}
+which leads to the equilibrium relation 
+\begin{eqnarray}
+\Delta G = -R T \ln \left( \frac{a_{(OH)_2}}{a_{H_2O}a_{O}} \right), \\
+= \Delta H - (T-T_0) \Delta S + (P-P_0) \Delta V
+\end{eqnarray}
+where $\Delta G$ is the Gibbs energy associated with the reaction.
+The second equation is a linearised form of the expression describing the change in free energy with pressure and temperature.
+The first equation can be simplified by making the approximation that $a_O$ is constant and that the weight percentage of H$_2$O in the solid $c_{H_2O}$ is proportional to the activity of (OH)$_2$.
+Both of these assumptions are probably reasonable at low values of $c_{H_2O}$.
+Rearranging now produces the expression:
+\begin{equation}
+c_{H_2O} \propto a_{H_2O} \exp{ \left( -\frac{\Delta H - (T-T_0) \Delta S + (P-P_0) \Delta V}{R T} \right) }
+\end{equation}
+Further simplifications can be made by taking $T_0 = 0$ K, by realising that $P_0$ can be neglected at high pressures, and by incorporating the factor $\Delta S/R$ into the exponential prefactor.
+Finally, it is recognised that the pressure and temperature dependence of the Gibbs free energy of water $G_{H_2O}$ is highly nonlinear.
+For this reason, we split the Gibbs free energy change of the anhydrous and hydrous solid from that of the liquid, yielding the equation \citep{KB2006}:
+\begin{eqnarray}
+c_{H_2O} = A a_{H_2O} f_{H_2O} \exp{ \left( -\frac{\Delta H + P \Delta V^{solid}}{R T} \right) }, \\
+f_{H_2O} = \exp \left(\frac{\Delta G_{H_2O}}{RT} \right)
+\end{eqnarray}
 
-Isolated protons (n=0.5), proton pairs (n=1) and proton . Making the approximation that $a_O = 1$, which is probably reasonable at low H concentrations in the solid 
-Making the assumption that concentration of H$_2$O in the solid is proportional to the activity of OH in the melt, we therefore have
-\begin{equation}
-  c \propto a_{H_2O}^n
-\end{equation}
-where $n$ relates to the mechanisms of hydrogen incorporation into the solid \citep{KB2006}
+We calculate the fugacity of pure water $f_{H_2O}$ from the equation of state of \citep{PS1995}.
+We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on wadsleyite and ringwoodite using our melt model.
+The pressure data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a negligable pressure dependence on water content of wadsleyite, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}.
+A single data point at 18 GPa has much lower water contents, which would require $\Delta V \sim 1.25 \cdot 10^{-5}$ m$^3$/mol and would produce an almost linear curve which produces large misfits at 16 and 17 GPa.
+We therefore favour the lower estimate of $\Delta V$, and assume the same value for ringwoodite.
+Other estimated values are shown in Table \ref{table:partitioning}.
+
+\begin{table}[]
+\centering
+\caption{Thermodynamic models for water solubility in wadsleyite and ringwoodite}
+\label{table:partitioning}
+\begin{tabular}{lllll}
+Phase & $P_{\textit{\tiny{ref}}}$ (Pa) & A [kg/(kgPa)] & $\Delta H_{P_{\textit{\tiny{ref}}}}$ [J/mol] & $\Delta V$ [m$^3$/mol] \\
+\hline
+Wadsleyite & 1.5e10 & 4.06e-12 & 1.66e5 & 1e-5 \\
+Ringwoodite & 2.0e10 & 2.19e-12 & 2.04e5 & 1e-5 (fixed) \\
+\end{tabular}
+\end{table}
+
+These models allow us to predict the partition coefficient of water between wadsleyite and ringwoodite at the 520 km discontinuity ($\sim$18 GPa).
+The coefficent increases gradually with temperature between 0.6 at 1000 $^{\circ}$C to almost 1.0 at 2000 $^{\circ}$C.
+One implication is that repartitioning of water during secular cooling of the Earth is likely to be much more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}.
+The second implication is that water saturated peridotite in the lower mantle will partially melt during upwelling, which could explain observations of low velocity layers at ca.
+520 km depth \citep[e.g.][]{JDH2010}.
 
@@
 
+\section{Conclusions}
+
 \clearpage

### 2015-07-10 MSH.tex
Checklist
Update

@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
+      \includegraphics[width=0.8\textwidth]{figures/assembly}
   \caption{The 18/11 octahedral assembly design used in this study.}
@@
 97 J/K/mol), so unless the thermodynamics of mixing in the liquid change markedly between these pressures, estimates of $c^{melt}_{H_2O}$ for the ringwoodite or wadsleyite studies (or both) must be somewhat inaccurate.
-We note that two of the three values provided by \cite{OMY2000} fall above the temperatures indicating that H$_2$O is the only H-bearing species in the melt \citep{SS1985}, and that the melting temperatures of wadsleyite + 20.4 wt\% H$_2$O (X$_{H_2O}$=0.4 in Figure \ref{fig:partitioning}) in the wadsleyite studies fall below that of the same composition at 5.5 GPa \citep{Inoue1994}.
-Both of these interpretations appear unlikely.
+We note that two of the three melt compositions in equilibrium with ringwoodite fall above the temperatures indicating that hydrogen exists only as molecular H$_2$O \citep{SS1985}.
+In addition, the calculated liquidus temperatures in the wadsleyite studies at 15 GPa are lower than those reported at 5.5 GPa \citep{Inoue1994}.
+Both of these observations appear highly unlikely.
 
@@
 
-It is also possible to create a thermodynamic model for incorporation of H$_2$O into wadsleyite and ringwoodite, given some knowledge of the substitution mechanism.
+\begin{figure}[ht!]
+  \centering
+      %\includegraphics[width=0.8\textwidth]{figures/partitioning}
+  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa.
+Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.
+Inset: Modelled partition coefficients between wadsleyite, ringwoodite and melt at 18 GPa.}
+  \label{fig:partitioning}
+\end{figure}
+
+We now create a thermodynamic model for incorporation of H$_2$O into wadsleyite and ringwoodite.
 Hydrogen in wadsleyite and ringwoodite is incorporated as proton pairs substituting for Mg (Mg$^{2+}$ $\leftarrow$ 2H$^{+}$) \citep{Smyth1987, SHFJLM2003}.
@@
 The pressure data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a negligable pressure dependence on water content of wadsleyite, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}.
-A single data point at 18 GPa has much lower water contents, which would require $\Delta V \sim 1.25 \cdot 10^{-5}$ m$^3$/mol and would produce an almost linear curve which produces large misfits at 16 and 17 GPa.
+A single data point at 18 GPa has much lower water contents, which would require $\Delta V \sim 1.25 \cdot 10^{-5}$ m$^3$/mol.
+This value is inconsistent with the large water contents at 16 and 17 GPa, and is probably unreasonable given the reported values for olivine.
 We therefore favour the lower estimate of $\Delta V$, and assume the same value for ringwoodite.
-Other estimated values are shown in Table \ref{table:partitioning}.
+The model values are shown in Table \ref{table:partitioning}.
 
@@
 
-These models allow us to predict the partition coefficient of water between wadsleyite and ringwoodite at the 520 km discontinuity ($\sim$18 GPa).
-The coefficent increases gradually with temperature between 0.6 at 1000 $^{\circ}$C to almost 1.0 at 2000 $^{\circ}$C.
-One implication is that repartitioning of water during secular cooling of the Earth is likely to be much more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}.
-The second implication is that water saturated peridotite in the lower mantle will partially melt during upwelling, which could explain observations of low velocity layers at ca.
+These models allow us to predict the partition coefficients of water between wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa).
+$DD^{wad/ring}_{H_2O}$ increases from 0.6 at 1000 $^{\circ}$C to almost 1.0 at 2000 $^{\circ}$C.
+That these values are <1 indicate that water saturated peridotite in the lower mantle will partially melt during upwelling, which could explain observations of low velocity layers at ca.
 520 km depth \citep[e.g.][]{JDH2010}.
+The small temperature dependence implies that repartitioning of water during secular cooling of the Earth is likely to be more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}.
 

### 2015-07-12 MSH.tex
Update

@@
 \subsubsection{MgO-H$_2$O}
-The dehydration of brucite at high pressure was previously studied by \cite{FIYKFO2005}.
-They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa.
-We observed large brucite crystals at 1200 $^{\circ}$C and 13 GPa; the temperature difference is probably within experimental uncertainties.
-In the present study, the composition of the fluid coexisting with brucite and periclase is constrained to be 64--66 mol\% H$_2$O.
+Experimental results for the system MgO-H$_2$O are shown in Figure \ref{fig:MH}.
+The periclase-brucite-water invariant at 13 GPa is well constrained at 62--66 mol\% H$_2$O and $\sim$1200 $^{\circ}$C.
 The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase.
-
-Using the ideal OH$^-$ mixing model of \cite{SS1985} with no compositional dependence on the formation of OH groups, $G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol).
-This asymmetry is required to fit the activity of water at the periclase-brucite-liquid triple point to the solid thermodynamic models of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995} ($0.52$).
-This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system.
-The second critical endpoint of the MgO-H$_2$O is probably around 13 GPa \citep{MSUP2007}, so large positive interaction enthalpies are justified in this region of the phase diagram.
+The melting curve and equilibrium with brucite is modelled with the ideal OH$^-$ mixing model of \cite{SS1985}.
+$G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol).
+This asymmetry is required to fit the activity of water at the periclase-brucite-liquid triple point to the solid thermodynamic models of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995}.
+This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system, as observed in the small dT/dX of the brucite liquidus.
+This flat slope is in excellent agreement with the suggestion that the second critical endpoint in the MgO-H$_2$O system probably lies at ca.
+12--13 GPa \citep{MSUP2007}.
 
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/per-H2O}
+      \includegraphics[width=0.8\textwidth]{figures/periclase}
   \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results.
-The position of the dry periclase melting point is taken from \cite{ZF2008}, with an entropy of dry melting from \cite{CG1994}.
-Model fits in grey use the model of \cite{SS1985}, with r=1 and K=$\infty$ (bottom), K=K(T) (expression in main text) and K=0 (all H2O as molecular, top).
-The black line includes an interaction term to match the fluid in equilibrium with brucite during it's decomposition to periclase.
-The black dotted line represents a regular solution model with MgO, Mg(OH)$_2$ and H$_2$O as distinct species in the melt (parameters in the main text).}
+Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (the discussion in main text).
+Properties of the anhydrous solid and melt are taken from published literature \citep{SLB2011, DKS2013}.
+An additional subregular interaction term is required to match the composition and temperature of brucite dehydration.}
   \label{fig:MH}
 \end{figure}
+
+The dehydration of brucite at high pressure was previously studied by \cite{FIYKFO2005}.
+They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa.
+The temperature difference between their study and ours is therefore 50 $^{\circ}$C, within experimental uncertainties.
 
@@
 The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}.
-The anhydrous melting of forsterite is incongruent.
-The temperature of the fo $\rightarrow$ per + L reaction is taken from \citep{PW1993}, with the temperature of the metastable fo $\rightarrow$ L reaction extrapolated from the same study.
-The liquidus is calculated by interpolation between the melting temperature of periclase \citep{ZF2008} and an estimate of the temperature and composition of melt in equilibrium with periclase and forsterite by interpolating between the invariant fo-per-foL point \citep{PW1993} and position of the per liquidus at 16 GPa \citep{LF2012}.
-
-One chamber from an experiment run at 1300 $^{\circ}$C revealed a single very small crystal of enstatite.
-Another run at 1200 $^{\circ}$C revealed large crystals of hydroxychondrodite in the chamber with the lowest water contents.
-Small differences in Mg:Si ratio could explain the differences in solid phases between the separate chambers, so it is assumed that the enstatite+forsterite and chondrodite+forsterite cotectics cross the forsterite+H$_2$O binary at high water contents and 1200--1300 $^{\circ}$C.
-This supposition is supported by \cite{DDFK2005}, who observe a double-crossing of the liquid and wadsleyite-water tie-line at 15 GPa, first at ca.
-1100 $^{\circ}$C, and again at 1400 $^{\circ}$C.
-Very similar Mg:Si ratios were observed by \citep{LSOK2011} at 15 and 18.5 GPa.
+The anhydrous melting of forsterite is incongruent, so the high temperature part of the liquidus is metastable with respect to a mixture of periclase and liquid (see Figure \ref{fig:ternary}).
+Water-rich chambers from experiments run at 1300 $^{\circ}$C revealed small quantities of enstatite, indicating that the forsterite-enstatite cotectic crosses the Mg$_2$SiO$_4$-H$_2$O tie line at 1300-1400 $^{\circ}$C.
+This is in excellent agreement with results on the wadsleyite-water system run at 15 GPa \citep{DDFK2005, LSOK2011}.
+At 1200 $^{\circ}$C, one chamber contained large crystals of hydroxychondrodite.
+The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained by MgO-deficit in forsterite at these conditions.
+This MgO deficit may be the result of hydrogen incorporation by the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}.
+Chambers with higher water contents have forsterites with similar compositions, but melts which are slightly more SiO$_2$-rich.
 
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/fo-H2O}
-  \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results. The position of the dry melting point is taken from \citep{PW1993}. \citep{LF2012}.}
+      \includegraphics[width=0.8\textwidth]{figures/forsterite}
+      \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results.
+The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
+At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.}
   \label{fig:foH}
@@
 \clearpage
+
 \subsubsection{MgSiO$_3$-H$_2$O}
+The melting of high pressure clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}.
+Unlike the depression of the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}.
+The liquidus is in excellent agreement with that estimated previously \citep{YII2004}, as is the absence of stishovite in any of the sample chambers.
+We also note that although we do not have any evidence for melt-fluid immiscibility, the very small observed dT/dX at ca.
+1500 $^{\circ}$C results in a very rapid change from water-rich to water-poor fluids with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}.
 
-The melting of clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}.
-The lack of stishovite as a liquidus phase is in agreement with \cite{YII2004}, who reported a clinoenstatite + stishovite + liquid field appearing at 13.5 GPa.
 \begin{figure}[ht!]
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/en-H2O}
-  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results. The position of the dry melting point is taken from \cite{PG1990}.}
+      \includegraphics[width=0.8\textwidth]{figures/enstatite}
+  \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results.
+The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.}
   \label{fig:eoH}
@@
 The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}.
-
-%NB: Most likely culprit the change in S with temperature?? Need a melt and solid model to test this.
-
-% We use the formulation of \cite{HM2012} to model SiO$_2$-H$_2$O melts at 13 GPa.
-The volume of stishovite is taken from \cite{HP2011}.
-The melting curve of coesite at this pressure has $dT/dP \sim 0$ \citep{ZLGHF1993}, so the volume of anhydrous SiO$_2$ at this pressure is roughly equal to the volume of coesite via the Clausium-Clapeyron relation $dT/dP = \Delta V / \Delta S$.
-Using these volumes, the metastable entropy of melting of stishovite $\Delta S_{stv-L}$ is estimated as 8.5 J/K/mol.
-For completeness, we also calculate the melting point and entropy of melting of ice VII at 13 GPa from \emph{P-V-T} data collected along the melting curve \citep{FFH2004}.
+While the deviation between the predicted liquidus depression for molecular H$_2$O-free melts and the clinoenstatite observations were relatively small, the deviation for the stishovite data is extreme.
+There are several possible explanations for this.
+First, the (metastable) melting point of anhydrous stishovite could be overestimated by ca.
+300 $^{\circ}$C, or the entropy of melting could be 10--20 J/K/mol, rather than 40 J/K/mol.
+Neither of these possibilities is very likely; the coesite and stishovite melting curves constrain the volume and entropy of SiO$_2$ melt in the region of the triple point at ca.
+13.5 GPa.
+A third possibility is that approximately 30\% volume liquidus-phase stishovite crystals were missed during SEM analysis.
+This is also extremely unlikely; the equant and dendritic morphologies of liquidus and quench stishovite are instantly recognisable, and the thermal gradients in the capsule lead to melt separation on the timescales of the experiments, such that even a small (5\%) volume proportion of solids should be detected.
 
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/SiO2-H2O}
+      \includegraphics[width=0.8\textwidth]{figures/stishovite}
       \caption{The silica-water phase diagram at 13 GPa, based on current experimental results.
-The liquidus curve is compared with previous experimentally-derived models computed at 1 GPa and 2 GPa \citep{HM2012}.
-An entropy of 13.5 J/K/mol is chosen to fit the data.
-Model fits use the model of \cite{SS1985}, with r=2 and K=$\infty$ (dotted grey), K=K(T) (black, expression in main text) and K=0 (dashed grey, all H2O as molecular).}
+The thermodynamics of stishovite and the anhydrous melt are described in the text.
+At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.}
   \label{fig:SH}
 \end{figure}
+
+Having ruled out experimental error and uncertainties in the thermodynamics of the anhydrous system, the only possibility is that hydrous SiO$_2$ melts with relatively low water contents are stabilised well beyond the expected entropic stabilisation from random protonation of the silica network.
+The causes of this additional stability (and, to a lesser extent, that of hydrous MgSiO$_3$ melts) are well beyond the scope of the current study, but may be a fruitful line of investigation for molecular dynamics simulations \citep[e.g.][]{KS2010}.
 
@@
 \subsection{The MgO-SiO$_2$-H$_2$O system}
-The set of experimental run products can be used to create a preliminary liquidus diagram (Figure \ref{fig:liquidus}).
+The set of experimental analyses can be used to create a preliminary ternary liquidus diagram (Figure \ref{fig:ternary}).
 
@@
 
+\citep{MSUP2007}
+
 \begin{figure}[ht!]
@@
       \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
-  \caption{A preliminary liquidus diagram at 13 GPa, based on current experimental results.}
-  \label{fig:liquidus}
+  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results.}
+  \label{fig:ternary}
 \end{figure}
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/partitioning}
-  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa.
-Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.
-Inset: Modelled partition coefficients between wadsleyite, ringwoodite and melt at 18 GPa.}
+      \includegraphics[width=0.8\textwidth]{figures/wad_rw}
+  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
+  \label{fig:wad_rw}
+\end{figure}
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=0.8\textwidth]{figures/wad_rw_partitioning}
+  \caption{Modelled partition coefficient between wadsleyite and ringwoodite at 18 GPa.}
   \label{fig:partitioning}
@@
 \hline
-Wadsleyite & 1.5e10 & 4.06e-12 & 1.66e5 & 1e-5 \\
-Ringwoodite & 2.0e10 & 2.19e-12 & 2.04e5 & 1e-5 (fixed) \\
+Wadsleyite & 1.5e10 & 1.97e-12 & 1.60e5 & 1e-5 \\
+Ringwoodite & 2.0e10 & 1.46e-12 & 2.03e5 & 1e-5 (fixed) \\
 \end{tabular}
@@
 These models allow us to predict the partition coefficients of water between wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa).
-$DD^{wad/ring}_{H_2O}$ increases from 0.6 at 1000 $^{\circ}$C to almost 1.0 at 2000 $^{\circ}$C.
-That these values are <1 indicate that water saturated peridotite in the lower mantle will partially melt during upwelling, which could explain observations of low velocity layers at ca.
+$D^{wad/ring}_{H_2O}$ increases from $\sim$0.7 at 1000 $^{\circ}$C to almost $\sim$0.9 at 2000 $^{\circ}$C.
+These values imply that water saturated ringwoodite in the lower mantle will partially melt as it transforms to wadsleyite during upwelling, which could explain observations of low velocity layers at ca.
 520 km depth \citep[e.g.][]{JDH2010}.

### 2015-07-12 MSH.tex
Update

@@
 \begin{abstract}
-
+Hydrous melting at high pressures affects the physical properties, dynamics and chemical differentiation of the Earth.
+However, probing the compositions of hydrous melts at the conditions of the mantle transition zone has traditionally been challenging.
 In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O.
-Water-rich starting compositions were obtained by using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source.
+Water-rich starting compositions were created using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source.
 As MgO:SiO$_2$ ratios decrease, the liquidus curve develops an increasingly pronounced concave-down topology.
@@
 Z1079 & 13 & 1300 & 35 & -/per/per/L/L & fo/fo/en/(en) & en/en/en/en & -/-/-/-/-/-/-/- & -/-/-/- \\
-Z1058 & 13 & 1400 & 30 & -/per/(per)/L/L & fo/fo/L/L & en/en/en/(en) & -/-/-/-/-/-/-/- & -/-/-/- \\
-Z1060 & 13 & 1500 & 20 & -/-/-/-/- & fo/L/L/L & en/en/L/L & -/-/-/-/-/-/-/- & -/-/-/- \\
+Z1058 & 13 & 1400 & 30 & -/per/(per)/L/L & fo/(fo)/L/L & en/en/en/(en) & -/-/-/-/-/-/-/- & -/-/-/- \\
+Z1060 & 13 & 1500 & 20 & -/-/-/-/- & (fo)/L/L/L & en/en/L/L & -/-/-/-/-/-/-/- & -/-/-/- \\
 Z1140 & 13 & 1600 & 10 & -/-/-/-/- & -/-/-/- & -/-/-/- & s/s/s/s/s/s/s/(s) & s+en/en/en/en \\
@@
 The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
-At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.}
+At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.
+The preferred fit through the data is given by $G^{\circ} = 75 \cdot (1420-T)$ J/mol.}
   \label{fig:foH}
@@
 Unlike the depression of the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}.
-The liquidus is in excellent agreement with that estimated previously \citep{YII2004}, as is the absence of stishovite in any of the sample chambers.
-We also note that although we do not have any evidence for melt-fluid immiscibility, the very small observed dT/dX at ca.
+The liquidus is in excellent agreement with that estimated by \cite{YII2004}, as is the absence of stishovite in any of the sample chambers.
+We note that although we do not have any evidence for melt-fluid immiscibility, the very small observed dT/dX at ca.
 1500 $^{\circ}$C results in a very rapid change from water-rich to water-poor fluids with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}.
@@
   \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results.
-The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.}
+The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
+The line K=K(T) is given by $G^{\circ} = 120 \cdot (1750-T)$ J/mol.}
   \label{fig:eoH}
@@
 The thermodynamics of stishovite and the anhydrous melt are described in the text.
-At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.}
+At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.
+The line K=K(T) is given by $G^{\circ} = 100 \cdot (1200-T)$ J/mol.}
   \label{fig:SH}
@@
 The set of experimental analyses can be used to create a preliminary ternary liquidus diagram (Figure \ref{fig:ternary}).
+It should be noted that the periclase liquidus field and periclase-forsterite cotectic are poorly constrained by the data.
+The low temperature extension of the cotectic is chosen to fit the reported Mg:Si ratios of melts in equilibrium with forsterite and various DHMS phases at 13.5 GPa \citep{MSUP2007}.
 
-\cite{YII2004} suggested that there was a sharp break in the liquidus at the enstatite-stishovite cotectic (their Figure 6).
-However, this prediction breaks Schreinemakers Rules (specifically, that no one assemblage can occupy more than 180$^{\circ}$ around an invariant point).
-
-\citep{MSUP2007}
+% Yamada et al. Figure 6 breaks Schreinemakers rules at high water contents... no reason to mention this in text, though...
 
@@
       \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
-  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results.}
+  \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results.
+Contours are in $^{\circ}$C, and mineral abbreviations are liquidus fields as follows: per - periclase, fo - forsterite, en - clinoenstatite, stv - stishovite, coe - coesite, phB - phase B, br - brucite, chond - hydroxychondrodite, ice - ice VII.}
   \label{fig:ternary}
@@
 \section{Discussion}
-\subsection{Melting in the MgO-SiO$_2$-H$_2$O system}
+\subsection{Melting in deep subduction zones and atop the mantle transition zone}
+Our melt model in the MSH system serves as a useful tool for investigating melting at high pressures in the mantle.
+In particular, the forsterite-enstatite cotectic implies that hydrous melts in equilibrium with depleted peridotites at typical mantle temperatures of 1500 $^{\circ}$C will contain $\sim$ 48 mol\% H$_2$O and an Mg:Si ratio of about 2 (the same as forsterite), corresponding to about 26 wt\% H$_2$O.
+Such compositions are far too water-rich to be neutrally buoyant relative to a forsterite + enstatite assemblage \citep{MSK2008}, and will rise through the mantle.
+Adding further components to the system will modify this behaviour.
+In particular, the addition of iron lowers anhydrous melting temperatures and increases melt densities.
+Thus, we cannot directly use our results to confirm or refute the viability of the water-filter model proposed by \cite{BK2003}, which requires neutrally buoyant or dense melts at the 410 km discontinuity.
+However, the extremely low dT/dX revealed by our experiments implies that between 1300 and 1700 $^{\circ}$C, water contents in the melt drop from 62 mol\% (34 wt\%) to around 30 mol\% (16 wt\%).
+Such large decreases in water content may lead to an abrupt sign change in density, where melts are buoyant at relatively low temperatures but dense at high temperatures.
+This would explain the preservation of high pressure phases in diamonds \citep{Pearsonetal2014}, which must have risen quickly from the mantle transition zone within a buoyant melt \citep{Keppler2014}.
+Large decreases in water contents are also likely to have a profound effect on melt viscosity, and may encourage reaction-infiltration instabilities encouraging melt channelisation \citep{SKA2001}.
+
+% Densities
+%\cite{MF2010}
+%\cite{BK2003}
+%\cite{MSK2008}
+
+%So at least in my models there are three quantities that are important for the pattern of melt migration, the permeability, the compaction viscosity and the shear viscosity (if you have elasticity/plasticity, other things come in, too).
+Everywhere in the equations where you have the permeability, it's divided by the melt viscosity, and the permeability itself is normally assumed to scale with phi^3 (or less common phi^2), and at least I didn't see any parametrisations describing the effect of temperature on the permeability, so I would assume it's small.
+So for the permeability, you can just use the values you get for your different P-T-X conditions, and see if the overall term permeability/melt_viscosity increases or decreases.
+
+%For the viscosities of the solid you have the influence of temperature on the viscosity (which I think is similar for shear and compaction viscosity) and you have the effect of porosity, eta ~ exp(alpa * phi) with phi~-27 for the shear viscosity (I think this is what Rich Katz uses in his models, but I would have to look it up to be sure), and xi ~ phi0/phi for the compaction viscosity, and you can do the same here, compare how they change for your different P-T-X conditions.
+
 \subsection{Water partitioning in the mantle transition zone}
@@
 A similar study on ringwoodite was undertaken by \cite{OMY2000}.
-The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:partitioning}).
+The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:wad_rw_melt}).
 For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O.
@@
   \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
-  \label{fig:wad_rw}
-\end{figure}
-
-\begin{figure}[ht!]
-  \centering
-  \includegraphics[width=0.8\textwidth]{figures/wad_rw_partitioning}
-  \caption{Modelled partition coefficient between wadsleyite and ringwoodite at 18 GPa.}
-  \label{fig:partitioning}
+  \label{fig:wad_rw_melt}
 \end{figure}
@@
 
-These models allow us to predict the partition coefficients of water between wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa).
+These models allow us to predict the partition coefficients of water between wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa; Figure \ref{fig:partitioning}).
 $D^{wad/ring}_{H_2O}$ increases from $\sim$0.7 at 1000 $^{\circ}$C to almost $\sim$0.9 at 2000 $^{\circ}$C.
@@
 
-\subsection{Deep water release from subduction zones}
-% Densities
-\cite{MF2010}
-\cite{BK2003}
-\cite{MSK2008}
-
-% Viscosities
-% Shear viscosity 
-% Density
-
-%So at least in my models there are three quantities that are important for the pattern of melt migration, the permeability, the compaction viscosity and the shear viscosity (if you have elasticity/plasticity, other things come in, too).
-Everywhere in the equations where you have the permeability, it's divided by the melt viscosity, and the permeability itself is normally assumed to scale with phi^3 (or less common phi^2), and at least I didn't see any parametrisations describing the effect of temperature on the permeability, so I would assume it's small.
-So for the permeability, you can just use the values you get for your different P-T-X conditions, and see if the overall term permeability/melt_viscosity increases or decreases.
-
-%For the viscosities of the solid you have the influence of temperature on the viscosity (which I think is similar for shear and compaction viscosity) and you have the effect of porosity, eta ~ exp(alpa * phi) with phi~-27 for the shear viscosity (I think this is what Rich Katz uses in his models, but I would have to look it up to be sure), and xi ~ phi0/phi for the compaction viscosity, and you can do the same here, compare how they change for your different P-T-X conditions.
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=0.8\textwidth]{figures/wad_rw_partitioning}
+  \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa. Note the separate scales for solid/solid and solid/melt partitioning.}
+  \label{fig:partitioning}
+\end{figure}
 

### 2015-07-14 MSH.tex
Sent to Dan

@@
 A similar study on ringwoodite was undertaken by \cite{OMY2000}.
-The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:wad_rw_melt}).
+The T-X$_{H_2O}$ dependencies of the melts in equilibrium with wadsleyite or ringwoodite are strikingly different (Figure \ref{fig:fo_wad_rw_melt}).
 For example, melt in equilibrium with wadsleyite at 15 GPa and 1400 $^{\circ}$C was estimated to contain 10.6-13.3 wt\% H$_2$O, while melt compositions in equilibrium with ringwoodite at 20 GPa and 1400 $^{\circ}$C suggest water contents of 57-66 wt\% H$_2$O.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/wad_rw}
+      \includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
   \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
-  \label{fig:wad_rw_melt}
+  \label{fig:fo_wad_rw_melt}
 \end{figure}
 
-We now create a thermodynamic model for incorporation of H$_2$O into wadsleyite and ringwoodite.
-Hydrogen in wadsleyite and ringwoodite is incorporated as proton pairs substituting for Mg (Mg$^{2+}$ $\leftarrow$ 2H$^{+}$) \citep{Smyth1987, SHFJLM2003}.
-Ringwoodite potentially also exhibits a substitution involving Si (Si$^{4+}$ $\leftarrow$ Mg$^{2+}$ + 2H$^{+}$) \citep{KKMO2000}.
+We now create thermodynamic models for incorporation of H$_2$O into forsterite, wadsleyite and ringwoodite.
+For forsterite, we use high pressure (12--14 GPa) experimental data on the solids from \cite{SFNHB2006} and \cite{LSKO2009}; neither of these studies estimate coexisting melt compositions.
+Hydrogen in all three minerals is incorporated as proton pairs substituting for Mg (Mg$^{2+}$ $\leftrightarrow$ 2H$^{+}$) \citep{Smyth1987, SHFJLM2003}.
+Ringwoodite potentially also exhibits a substitution involving Si (Si$^{4+}$ $\leftrightarrow$ Mg$^{2+}$ + 2H$^{+}$) \citep{KKMO2000}.
 Both of these reactions may be written in the form
@@
 We calculate the fugacity of pure water $f_{H_2O}$ from the equation of state of \citep{PS1995}.
-We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on wadsleyite and ringwoodite using our melt model.
-The pressure data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a negligable pressure dependence on water content of wadsleyite, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}.
+We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on forsterite, wadsleyite and ringwoodite using our melt model.
+The spike in olivine water concentrations at 1250 $^{\circ}$C reported by \cite{SFNHB2006} is difficult to fit with reasonable enthalpies.
+The wadsleyite data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a low pressure effect on water content, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}.
 A single data point at 18 GPa has much lower water contents, which would require $\Delta V \sim 1.25 \cdot 10^{-5}$ m$^3$/mol.
@@
 \hline
+% Olivine & 1.2e10 & 1.74e-11 & 1.70e5 & 1e-5 \\ Not a bad fit for the Zhao enthalpy
+% Olivine & 1.2e10 & 5e-12 & 1.70e5 & 1e-5 \\ Zhao et al., 2004, water contents much too high
+Olivine & 1.2e10 & 3.23e-12 & 1.50e5 & 1e-5 \\
 Wadsleyite & 1.5e10 & 1.97e-12 & 1.60e5 & 1e-5 \\
-Ringwoodite & 2.0e10 & 1.46e-12 & 2.03e5 & 1e-5 (fixed) \\
+Ringwoodite & 2.0e10 & 1.46e-12 & 2.03e5 & 1e-5 (fixed)
 \end{tabular}
@@
 
-These models allow us to predict the partition coefficients of water between wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa; Figure \ref{fig:partitioning}).
-$D^{wad/ring}_{H_2O}$ increases from $\sim$0.7 at 1000 $^{\circ}$C to almost $\sim$0.9 at 2000 $^{\circ}$C.
-These values imply that water saturated ringwoodite in the lower mantle will partially melt as it transforms to wadsleyite during upwelling, which could explain observations of low velocity layers at ca.
-520 km depth \citep[e.g.][]{JDH2010}.
-The small temperature dependence implies that repartitioning of water during secular cooling of the Earth is likely to be more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}.
+These models allow us to predict the partition coefficients of water between olivine, wadsleyite and melt at the 410 km discontinuity ($\sim$14 GPa; Figure \ref{fig:partitioning_rw_wad}), and wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa; Figure \ref{fig:partitioning_rw_wad}).
+$D^{wad/fo}_{H_2O}$ displays a decrease from $\sim$4 at 1000 $^{\circ}$C to $\sim$1.8 at 2000 $^{\circ}$C, in good agreement with the estimate of \cite{LSOK2011} at low temperatures, but without a return to high values at high temperatures, where water contents are lower and prone to large relative errors.
+$D^{ring/wad}_{H_2O}$ decreases from $\sim$1.4 at 1000 $^{\circ}$C to $\sim$1.1 at 2000 $^{\circ}$C.
+These values imply that upwelling across both the 410 and the 520 km discontinuity will cause partial melting.
+Both boundaries have been associated with low seismic velocities, which may indicate zones of deep melting \citep[e.g.][]{JDH2010}.
+The small temperature dependence on the wadsleyite-ringwoodite partition coefficient implies that repartitioning of water during secular cooling within the mantle transition zone is likely to be more minor (and in the opposite direction) to that proposed by \cite{DDFK2005}.
 
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/wad_rw_partitioning}
-  \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa. Note the separate scales for solid/solid and solid/melt partitioning.}
-  \label{fig:partitioning}
+  \includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
+  \caption{Modelled partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity.
+Note the separate scales for solid/solid and solid/melt partitioning.}
+  \label{fig:partitioning_wad_fo}
+\end{figure}
+
+\begin{figure}[ht!]
+  \centering
+  \includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
+  \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity.
+Note the separate scales for solid/solid and solid/melt partitioning.}
+  \label{fig:partitioning_rw_wad}
 \end{figure}

### 2015-07-24 MSH.tex
Second draft

@@
 The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}.
-This deep water cycle is probably the source for some kimberlitic eruptions, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}.
-The ringwoodite stability field requires that the kimberlite must have been sourced from at least ca.
-500 km depth.
+This deep water cycle must extend to at least 500 km depth, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}.
 In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage.
@@
 Processes involving deep-seated hydrous melts are controlled to a great extent by melt composition.
-Composition is a key variable in determining melt density, viscosity and solid-solid-melt dihedral angles, which together with the temperature and equivalent properties of the solid control the time- and lengthscales on which melt can separate from the melt.
-They also govern the extent to which this melt can become channelised, reducing interaction with its surroundings and increasing ascent rates.
+Most importantly, the equilibrium proportion of H$_2$O in melts controls the extent of partial melting for a given bulk H$_2$O content.
+Together with solid-solid-melt dihedral angles, the extent of partial melting governs the permeability and therefore mobility of melt, and the ability to detect it via body wave seismology and electrical conductivity measurements.
+Composition is also a key variable in determining melt density and viscosity.
+Finally, the change in equilibrium melt compositions with pressure and temperature has a significant influence on melt channelisation, which reduces melt-solid interaction and increases ascent rates.
+
 Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure.
@@
 Constraining melt compositions remains a major challenge in high-pressure experimental petrology.
-The primary problem is that hydrous melts are unquenchable, which means that their water concentrations must be estimated indirectly.
-Compositions of partial hydrous melts have been estimated using a mixture of mass balance and defocused-beam electron probe measurements, which to provide Mg:Si ratios and a rough estimate of water content from deficits in totals.
-Such measurements are conducted on melt pockets \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011} or melts isolated within diamond traps \citep{MSUP2007}.
-Unfortunately, these techniques are accompanied by significant uncertainties, especially in terms of water content.
-An alternative technique is to constrain the liquidus temperature at a given composition, thus circumventing the need to estimate water concentrations in the melt \emph{ex-situ} (when much of the water has been lost).
-The problem with this technique is that at ambient pressures, hydrous phases are limited in their maximum water content.
-For example, in the MgO-SiO$_2$ system, maximum water contents in a solid starting material are provided by a mixture of brucite and quartz.
-Thus, studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form.
+The primary problem is that water-rich melts are unquenchable.
+During quench, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour on decompression and sample preparation.
+The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}.
+Mass balance alone depends on knowing the volume proportions of solid phases and quench phases, being able to recognise and subtract quench overgrowths where they exist, and knowing the effective ambient-pressure density of the melt.
+All of these are associated with significant uncertainties.
+Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit would imply a large water content).
+However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam.
+For this reason, defocused beam techniques should probably be restricted to estimating non-volatile element ratios (such as Mg:Si).
+
+Bracketing the liquidus requires only that the bulk composition during the experiment is known.
+This technique therefore seems a good alternative to estimating melt compositions ex-situ.
+The only problem is that at slab or ambient mantle temperatures, water contents in melts at high pressure are very high.
+Making up solid starting compositions with more water than a brucite-quartz mix has traditionally been avoided.
+Existing studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form if the sample chambers are sufficiently large \citep[e.g.][]{HM2012}.
 Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible.
-
 In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$), as a novel water source.
@@
 Pressure calibrations and details of the press can be found in \cite{FPTLDR2004} and \cite{KF2005}.
-The assemblies were then resistively heated via the stepped LaCrO$_3$ heater.
+The assemblies were then resistively heated via a stepped LaCrO$_3$ heater.
 The temperature was recorded using a W$_{97}$Re$_{3}$--W$_{75}$Re$_{25}$ (Type D) thermocouple inserted axially into the assembly.
@@
 \section{Modelling water solubility in melts}
-
-Oxygen in MgO-SiO$_2$-H$_2$O melts can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group.
+\subsection{Background}
+Oxygen in neutral MSH melt species can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group.
 The general equation describing the reaction between melt species is
 \begin{equation}
-0.5 \textrm{H}_2\textrm{O} + 0.5 \textrm{O}_{br}^{2-} \rightleftharpoons \textrm{OH}_{tm}^-
+\textrm{H}_2\textrm{O} + \textrm{O}_{br}^{2-} \rightleftharpoons 2 \textrm{OH}_{tm}^-
 \label{eqn:speciation}
@@
 \begin{equation}
-K = \frac{X_{\textrm{OH}_{tm}^-}}{\sqrt{X_{\textrm{H}_2\textrm{O}} X_{\textrm{O}_{br}^{2-}} } }
+K_{(1)} = \frac{{X_{\textrm{OH}_{tm}^-}}^2}{X_{\textrm{H}_2\textrm{O}} X_{\textrm{O}_{br}^{2-}} }
 \label{eqn:equilibrium_constant}
@@
 
-The equilibrium constant is related to the energy $\Delta G^{\circ}$ required to protonate a bridging oxygen
+The equilibrium constant is related to the energy $\Delta G_{(1)}$ required to protonate two bridging oxygens with a water molecule
 \begin{equation}
-K = \exp\left(\frac{-\Delta G^{\circ}}{RT}\right)
+K_{(1)} = \exp\left(\frac{-\Delta G_{(1)}}{RT}\right)
 \end{equation}
@@
 \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation.
-For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers.
+For example, an Mg$_2$SiO$_4$ melt with $r=4$ and $K_{(1)}=\infty$ represents a continuum between an anhydrous melt and one made of Mg(OH)$_2$ and Si(OH)$_4$ monomers.
 No distinction is made between isolated oligomers and partial depolymerisation of a silicate network; in other words, the energy required to form a terminal OH group is independent of local environment.
-Fixing $r$ places an implicit constraint on the maximum value of $K$ in water-rich compositions.
-For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K<\infty$.
-In practise, the equilibrium constant $K$ has been shown to be dependent on temperature and composition.
+Fixing $r$ places an implicit constraint on the maximum value of $K_{(1)}$ in water-rich compositions.
+For example, if $r=1$ in a hydrated MgO melt, then water contents exceeding 50 mol\% require that $K_{(1)}<\infty$.
+In practise, the equilibrium constant $K_{(1)}$ has been shown to be dependent on temperature and composition.
 \cite{SK1995} obtained a good fit to isochemical data with an expression of the form
 \begin{equation}
-\Delta G^{\circ} = a + b\,T
+\Delta G_{(1)} = a + b\,T
 \end{equation}
 
-\cite{THH2012} simplified the model of \cite{SS1985} by assuming that all hydrogen in the melt exists as OH$^-$, and that protonation is equally likely on any oxygen ($K=\infty$, $\Delta G^{\circ}=-\infty$, $r$ maximised).
+\cite{THH2012} simplified the model of \cite{SS1985} by assuming that all hydrogen in the melt exists as OH$^-$, and that protonation is equally likely on any oxygen ($K_{(1)}=\infty$, $\Delta G_{(1)}=-\infty$, $r$ maximised).
 Deviations from this model must increase with increasing water content, and such a model cannot describe any melt more water-rich than the Mg(OH)$_2$-Si(OH)$_4$ binary.
@@
 To describe melting in the SiO$_2$-H$_2$O system, \cite{HM2012} generalised the aforementioned models by allowing non-ideality between the two endmembers.
-The excess term $\Delta G^{\circ}$(P,T) for the formation of OH$^-$ groups was assumed to be independent of composition; all oxygens are equally available for protonation, and the local extent of protonation does not affect the energy required for each further protonation.
+The excess term $\Delta G_{(1)}$(P,T) for the formation of OH$^-$ groups was assumed to be independent of composition; all oxygens are equally available for protonation, and the local extent of protonation does not affect the energy required for each further protonation.
 
-In these models, equilibrium melt compositions are calculated by calculating the activity of a component $x$ in the melt which has the composition of the anhydrous liquidus phase $X$ from the Gibbs energy of melting of $X$ at the temperature of interest:
+These models can be used to calculate melt compositions by equating the gibbs free energy of the solid $G^s_{solid}$ with the chemical potential of the component with the composition of that solid in the melt $\mu^s_{melt}$:
 \begin{equation}
-RT \ln a_{x,L}^X = G_{X_l} - G_{X_{s}} = - \int_T^{T_{melt}} S_{X_{l}}-S_{X_s} \, dT
+G^s_{solid} = \mu^s_{melt} = G^s_{melt} - RT \ln a^s_{melt}
+\label{eqn:equilibrium}
 \end{equation}
-where $T_{melt}$ is the temperature of melting of the liquidus phase under anhydrous conditions.
+In this study, we use the thermodynamic dataset of \cite{SLB2011} to calculate $G^s_{solid}$ for periclase, forsterite and its high pressure polymorphs, orthoenstatite, high pressure clinoenstatite, stishovite and coesite.
+A modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} is used to calculate $G^s_{melt}$, and the \cite{SS1985} model is used to calculate $a^s_{melt}$.
+The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/}][]{CHRU2014}.
 
-In this study, we use a modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} using melt endmembers MgO and SiO$_2$.
-For MgO melt, only the reference Helmholtz free energy needed to be adjusted.
-We note that the melting point of MgO at high pressure is rather uncertain.
-In the present model \citep{DKS2013}, MgO melts at $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}.
-Experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}.
-We note that if the present model is incorrect, it is probably the calculated entropy of melting which is at fault.
-A higher melting temperature would therefore imply a lower entropy of melting, such that the values in $G_{\textit{melt}}$ - $G_{\textit{solid}}$ at low temperatures will have smaller errors, than would otherwise be expected from such large melting point uncertainties.
-For the SiO$_2$ melt, the bulk modulus of the SiO$_2$ endmember was increased to satisfy the experimental constraint that coexisting melt and coesite have similar volumes at 13--14 GPa \citep{ZLGHF1993}.
-The volume at 1 bar is fit to experimental data.
-The entropy of melting at the coesite-stishovite-melt triple point ($\sim$40 J/K/mol) was kept the same.
-This value is identical to the value derived from the Clapeyron slope of the stishovite liquidus \cite{SL1995}.
-We note that this slope is much steeper than that recently proposed by \cite{Millotetal2015}.
-This difference is not the result of incompatibility between experimental results at different pressures \citep{SL1995, LAM1983, Millotetal2015}, but instead the result of using a 2-parameter Simon's equation to fit the rather abrupt change in dT/dP implied by the experiments at 20--40 GPa.
+\subsection{The anhydrous melt model}
+Thermodynamic models of MgO-SiO$_2$ melts are currently in their infancy.
+Many of the techniques used to constrain the thermodynamic properties (particularly the equations of state) of solids cannot be used on liquids as a result of their highly dynamic behaviour.
+Therefore in this study, we use MgO and SiO$_2$ melt models derived from molecular dynamics simulations \citep{DKS2013}, and fit a simplified mixing curve based on experimental data.
+First, however, the thermodynamic properties of the endmember melts need to be adjusted to fit the experimentally derived thermodynamic data at the melting point.
+This step was unnecessary in the original study of \cite{DKS2013} because solid properties were also obtained by molecular dynamics simulations.
+Such simulations do not provide good estimates of absolute energies and entropies, and so the values of the solids were obtained by fixing the melting point at a given pressure to the experimentally derived values (0 GPa, 3070 K for periclase, 14 GPa, 3120 K for stishovite).
+It turns out that the MD-determined entropy of periclase is almost identical to the experimentally determined values at all pressures and temperatures \citep{SLB2011}, so we only needed to adjust the standard state free energy.
+This leads to a periclase melting point of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}.
+We notes that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}.
+Current thermodynamic models for periclase cannot be extrapolated to the extreme temperatures predicted by the \cite{ZF2008} data.
+However, assuming that the volume change of melting is roughly constant with temperature (1.8 cm$^3$/mol at 13 GPa), the entropy of melting at a melting point of 5373 K would be about 50\% of that for a melting point of 4180 K (13.25 vs.
+27.5 J/K/mol).
+At 13 GPa and 2000 K, MgO melt would actually be 15 kJ/mol \emph{more} stable as a result of this lower entropy.
 
-A subregular solution model was adopted to describe the free energy between the MgO and SiO$_2$ melts.
-The two interaction parameters W$_{MS}$ and W$_{SM}$ at 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions.
-The temperature dependence around 14 GPa was calculated by assuming an ideal excess entropy contribution, and then fitting to the known melting curves of forsterite and enstatite.
+The MD-determined entropy of stishovite is significantly lower than the experimentally determined value, so we increased the standard state entropy of the melt by 16.5 J/K/mol.
+Conveniently, we also have a way to test the volumes predicted by the model of \cite{DKS2013} at around 13 GPa.
+The experimentally-derived slope of the coesite melting curve is essentially zero at the triple point with stishovite at 13.7 GPa \citep{ZLGHF1993}.
+Therefore, from the Clausius-Clapeyron relationship $dT/dP = \Delta V / \Delta S$ the volumes of coesite and SiO$_2$ melt must be zero at this point.
+Unfortunately, the MD-derived SiO$_2$ melt model has its density crossover with coesite about 4 GPa lower.
+Therefore, we adjusted the bulk modulus and reference pressure so satisfy this constraint, and the volume at zero pressure.
+The adjustments required here may not be a failing of the MD simulations; it is likely that the coordination changes in the SiO$_2$ melt require a higher order Birch Murnaghan EoS than that used by \citep{DKS2013}.
+Our adjusted model is probably only accurate in a range of about $\pm$5 GPa around the triple point.
+The experimental data can also be used to judge the accuracy of the entropy of the SiO$_2$ melt model.
+The MD-derived entropy of melting is $\sim$40 J/K/mol at the triple point.
+This value is in perfect agreement with the value derived from the Clapeyron slope of the stishovite liquidus \citep{SL1995}.
+We note that the slope recently proposed by \cite{Millotetal2015} is much shallower, but their slope is a two parameter Simon's equation fit based on the triple point temperature at 13.7 GPa and the melting point at 500 GPa, and fits the data at 15--100 GPa \citep{SL1995, LAM1983} rather poorly.
+A Simon's equation fit is probably inappropriate over such extreme pressure ranges, especially where the melting point may be rather sensitive to changes in cation coordination.
+
+A subregular solution model was adopted to describe the free energy of intermediate-composition MgO-SiO$_2$ melts.
+The two interaction parameters W$_{MS}$ and W$_{SM}$ at 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions \citep{PG1990, PW1993}.
+We assumed an ideal entropic contribution to mixing.
 The pressure dependencies were fit to the melting curves.
@@
 
-The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/}][]{CHRU2014}.
+\subsection{The hydrous melt model}
+
+In this study, the anhydrous components in our hydrous melts are normalised to a one-cation basis (i.e.
+MgO, Mg$_{2/3}$Si$_{1/3}$O$_{4/3}$, Mg$_{1/2}$Si$_{1/2}$O$_{3/2}$, SiO$_2$).
+We follow \cite{SS1985} in determining the proportions of H$_2$O + O$_{br}$ and OH$_{tm}$  in the melt as a function of the molar proportion of H$_2$O $p_{H_2O}$ as follows:
+
+\begin{eqnarray}
+p' = \frac{p_{H_2O}}{p_{H_2O}  + r \left(1.-p_{H_2O} \right)} \\
+X_O = 1 - p' - \frac{ 0.5 - \sqrt{ 0.25 - \frac{K_{(1)}-4}{K_{(1)}} \left(p'-p'^2 \right)}}{\frac{K_{(1)}-4}{K_{(1)}} }\\
+X_{H_2O} = X_O + 2 p' - 1 \\
+X_{OH} = 1 - X_O - X_{H_2O}
+\end{eqnarray}
+\noindent $p'$ is the molar proportion of H$_2$O where the silicate is normalised on a one oxygen basis (rather than a one-cation basis).
+We assume that all bridging oxygens are equally likely to be protonated.
+The number of oxygens per formula unit $r$ thus increases monotonously from 1 to 2 across the MgO-SiO$_2$ binary ($r_{\textrm{per}}$ = 1, $r_{\textrm{fo, wad, rw}}$ = 4/3, $r_{\textrm{oen, cen}}$ = 3/2, and $r_{\textrm{coe, stv}}$ = 2).
+The ideal activities of the anhydrous and water components can then be calculated:
+\begin{eqnarray}
+a^{\textit{s}}_{\textit{ideal melt}} =X_O^r \\
+a^{H_2O}_{\textit{ideal melt}} = X_{H_2O}
+\end{eqnarray}
+
+The activity coefficients (i.e.
+the activity prefactor which provides the non-ideal contribution to mixing; $a = \gamma a_{ideal}$) are equal to one except in the case of the MgO-H$_2$O binary, where a simple Margules model is applied \citep{HM2012}
+\begin{equation}
+RT \ln \gamma_i = {p_j}^2 \left( 1 - p_j \right) \left( 2 W_{ji} - W_{ij} \right) \\
+\end{equation}
+\noindent where the subscripts $i$ and $j$ correspond to the components $s$ or H$_2$O.
+Equilibrium with the solid can now be found using Equation \ref{eqn:equilibrium}, noting that the dataset-derived values of $G^s_{solid}$ of the forsterite (Mg$_2$SiO$_4$) and enstatite (Mg$_2$Si$_2$O$_6$) polymorphs should be divided by 3 and 4 to reduce them to their one-cation equivalents.
 
@@
 The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase.
-The melting curve and equilibrium with brucite is modelled with the ideal OH$^-$ mixing model of \cite{SS1985}.
-$G^{\circ}= -75000$ provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol).
-This asymmetry is required to fit the activity of water at the periclase-brucite-liquid triple point to the solid thermodynamic models of \citep{HP2011} and the equation of state of water formulated by \cite{PS1995}.
-This model also has the effect of increasing the tendency for melt-fluid immiscibility at the water-rich end of the system, as observed in the small dT/dX of the brucite liquidus.
-This flat slope is in excellent agreement with the suggestion that the second critical endpoint in the MgO-H$_2$O system probably lies at ca.
+In addition to the thermodynamic models described in the methods section, the activity of H$_2$O at the brucite-periclase-melt triple point is calculated using the thermodynamic dataset of \citep{HP2011}, which includes models for brucite, periclase and water \citep{PS1995}.
+This provides an additional constraint on the melt model via the equation
+\begin{equation}
+RT \ln a_{H_2O} =  G_{br} - G_{per} - G_{H_2O}
+\end{equation}
+The brucite melting curve is obtained by finding the solution to the equation
+\begin{equation}
+(G_{MgO, melt} + RT \ln a_{MgO}) + (G_{H_2O, HP} + RT \ln a_{H_2O}) = G_{br, HP} - G_{per, HP} + G_{per, SLB}
+\end{equation}
+\noindent where the right hand side is the chemical potential of Mg(OH)$_2$ in the melt, and the right hand side is the Gibbs free energy of brucite with a correction for the different standard state energies of MgO between the \cite{HP2011} and \cite{SLB2011} datasets.
+$G_{(1)} = RT \ln K_{(1)} = -75000$ J/mol ($r$=1) provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol).
+The large positive interaction parameter increases the tendency for melt-fluid immiscibility at the water-rich end of the system.
+This is reflected in the flat slope of the brucite liquidus, and is in excellent agreement with the suggestion that the second critical endpoint in the MgO-H$_2$O system probably lies at ca.
 12--13 GPa \citep{MSUP2007}.
@@
   \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results.
-Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (the discussion in main text).
+Model fits to the data use the melt formulation of \cite{SS1985}, with r=1 and K=K(T) (see discussion in main text).
 Properties of the anhydrous solid and melt are taken from published literature \citep{SLB2011, DKS2013}.
@@
 The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}.
-The anhydrous melting of forsterite is incongruent, so the high temperature part of the liquidus is metastable with respect to a mixture of periclase and liquid (see Figure \ref{fig:ternary}).
+The model parameters shown are $G_{(1)} = 75 \cdot (1420-T)$ J/mol, $r=4/3$.
+The anhydrous melting of forsterite is incongruent, so the high temperature part of the liquidus is metastable with respect to a mixture of periclase and liquid (see also Figure \ref{fig:ternary}).
 Water-rich chambers from experiments run at 1300 $^{\circ}$C revealed small quantities of enstatite, indicating that the forsterite-enstatite cotectic crosses the Mg$_2$SiO$_4$-H$_2$O tie line at 1300-1400 $^{\circ}$C.
@@
 At 1200 $^{\circ}$C, one chamber contained large crystals of hydroxychondrodite.
-The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained by MgO-deficit in forsterite at these conditions.
-This MgO deficit may be the result of hydrogen incorporation by the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}.
-Chambers with higher water contents have forsterites with similar compositions, but melts which are slightly more SiO$_2$-rich.
+The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained an MgO-deficit in forsterite resulting from the incorporation of hydrogen via the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}.
+Alternatively it could simply be the result of a small SiO$_2$ deficit in that sample chamber.
 
@@
 The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
-At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.
-The preferred fit through the data is given by $G^{\circ} = 75 \cdot (1420-T)$ J/mol.}
+At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.}
   \label{fig:foH}
@@
 The melting of high pressure clinoenstatite in the presence of water is shown in Figure \ref{fig:eoH}.
-Unlike the depression of the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}.
-The liquidus is in excellent agreement with that estimated by \cite{YII2004}, as is the absence of stishovite in any of the sample chambers.
-We note that although we do not have any evidence for melt-fluid immiscibility, the very small observed dT/dX at ca.
-1500 $^{\circ}$C results in a very rapid change from water-rich to water-poor fluids with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}.
+Unlike the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}.
+The liquidus at higher water contents can be fit well with model parameters $G_{(1)} = 120 \cdot (1750-T)$ J/mol, $r=3/2$.
+The disagreement between this model and the experimental observations is small, and could plausibly be due to experimental error.
+However, given the results for stishovite (see next section), we feel that more complex models are required to explain the depression of melting in the MgSiO$_3$ system.
 
@@
   \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results.
-The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
-The line K=K(T) is given by $G^{\circ} = 120 \cdot (1750-T)$ J/mol.}
+The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.}
   \label{fig:eoH}
 \end{figure}
+
+Our experimentally determined liquidus is in excellent agreement with the experimental results of \cite{YII2004}, as is the absence of stishovite in any of the sample chambers.
+We note that although we do not have any evidence for melt-fluid immiscibility, the flat liquidus at ca.
+1500 $^{\circ}$C causes a very rapid change from water-rich to water-poor fluids with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}.
+The implication is that for a given bulk H$_2$O content, a sharp rise in the degree of partial melting will occur at about 1500 $^{\circ}$C in hydrous MgSiO$_3$ melts.
 
@@
 The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}.
-While the deviation between the predicted liquidus depression for molecular H$_2$O-free melts and the clinoenstatite observations were relatively small, the deviation for the stishovite data is extreme.
+While the deviation between the predicted enstatite liquidus depression for ideal molecular H$_2$O-free melts and the experimental observations are relatively small, the deviation for the stishovite data is extreme.
+The mixing model presented in Figure \ref{fig:SH} has parameters $G_{(1)} = 100 \cdot (1200-T)$ J/mol, $r=2$ is fit to the extrapolation of the experimentally defined liquiudus, but clearly fails to fit any of the experimental data.
 There are several possible explanations for this.
-First, the (metastable) melting point of anhydrous stishovite could be overestimated by ca.
+First, the melting point of anhydrous stishovite (which is metastable with respect to coesite) could be overestimated by ca.
 300 $^{\circ}$C, or the entropy of melting could be 10--20 J/K/mol, rather than 40 J/K/mol.
-Neither of these possibilities is very likely; the coesite and stishovite melting curves constrain the volume and entropy of SiO$_2$ melt in the region of the triple point at ca.
+Neither of these possibilities is very likely; the coesite and stishovite melting curves provide good constraints on the volume and entropy of SiO$_2$ melt in the region of the triple point at ca.
 13.5 GPa.
@@
 The thermodynamics of stishovite and the anhydrous melt are described in the text.
-At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.
-The line K=K(T) is given by $G^{\circ} = 100 \cdot (1200-T)$ J/mol.}
+At high temperatures, the stishovite liquidus is metastable with respect to coesite and melt.}
   \label{fig:SH}
@@
 
-%\citep{SU2001}
-%\citep{KOM2005}
-%\citep{AA1980}
-%\citep{Wunder1998}
-%\citep{AFRH2001}
-%\citep{Nagano2002}
-%\citep{MSUP2007}
-%\citep{MFY2002}
-%\citep{Sta2004}
-%\citep{MFY2004}
-%\citep{SUTG2001}
-%\citep{NM2008}
-%\citep{DM2010}
-%\citep{ZK1998}
+%\citep{SU2001, KOM2005, AA1980, Wunder1998, AFRH2001, Nagano2002, MFY2002, Sta2004, MFY2004, SUTG2001, NM2008, DM2010, ZK1998}
 
@@
 Thus, we cannot directly use our results to confirm or refute the viability of the water-filter model proposed by \cite{BK2003}, which requires neutrally buoyant or dense melts at the 410 km discontinuity.
-However, the extremely low dT/dX revealed by our experiments implies that between 1300 and 1700 $^{\circ}$C, water contents in the melt drop from 62 mol\% (34 wt\%) to around 30 mol\% (16 wt\%).
-Such large decreases in water content may lead to an abrupt sign change in density, where melts are buoyant at relatively low temperatures but dense at high temperatures.
-This would explain the preservation of high pressure phases in diamonds \citep{Pearsonetal2014}, which must have risen quickly from the mantle transition zone within a buoyant melt \citep{Keppler2014}.
-Large decreases in water contents are also likely to have a profound effect on melt viscosity, and may encourage reaction-infiltration instabilities encouraging melt channelisation \citep{SKA2001}.
+However, the flat liquidus curves revealed by our experiments are likely to also be observed in more natural chemical systems.
+The current results imply that the water contents of melts in equilibrium with forsterite and enstatite drop from 62 mol\% (34 wt\%) to around 30 mol\% (16 wt\%) between 1300 and 1700 $^{\circ}$C.
+For a given water content, this rapid change in composition must be balanced by a rapid increase in the mass fraction of partial melt, which could greatly enhance chances of detection by seismological or other geophysical techniques.
+Such large decreases in water content would also lead to large density changes; close to slabs, hydrous melts could be buoyant, while in warmer mantle of a similar composition they would be dense.
+This scenario might be expected to result in channelisation of hydrous melts close to the top interface of subducting slabs.
+Finally, large decreases in water contents are also likely to have a profound effect on melt viscosity, and may encourage reaction-infiltration instabilities encouraging melt channelisation \citep{SKA2001}.
 
@@
 
-Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that the thermodynamic properties of the melt remain essentially constant between 13 and 20 GPa.
+Here, we recalculate the wadsleyite/melt and ringwoodite/melt partitioning values with our melt model, assuming that our fitted values of $G_{(1)}$ and $r$ are suitable between 13 and 20 GPa.
+We justify this by noting that speciation is likely to be only weakly dependent on pressure under these conditions.
 Water concentrations in the solid are taken from the original studies.
@@
 We calculate the fugacity of pure water $f_{H_2O}$ from the equation of state of \citep{PS1995}.
-We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on forsterite, wadsleyite and ringwoodite using our melt model.
-The spike in olivine water concentrations at 1250 $^{\circ}$C reported by \cite{SFNHB2006} is difficult to fit with reasonable enthalpies.
+We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on forsterite, wadsleyite and ringwoodite using our melt model to provide estimates of $a_{H_2O}$.
+The sharp spike in olivine water concentrations at 1250 $^{\circ}$C reported by \cite{SFNHB2006} is difficult to fit with reasonable enthalpies, but we obtain good fits with the rest of their data, and those of \cite{LSKO2009}.
 The wadsleyite data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a low pressure effect on water content, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}.
@@
 These models allow us to predict the partition coefficients of water between olivine, wadsleyite and melt at the 410 km discontinuity ($\sim$14 GPa; Figure \ref{fig:partitioning_rw_wad}), and wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa; Figure \ref{fig:partitioning_rw_wad}).
-$D^{wad/fo}_{H_2O}$ displays a decrease from $\sim$4 at 1000 $^{\circ}$C to $\sim$1.8 at 2000 $^{\circ}$C, in good agreement with the estimate of \cite{LSOK2011} at low temperatures, but without a return to high values at high temperatures, where water contents are lower and prone to large relative errors.
+$D^{wad/fo}_{H_2O}$ displays a decrease from $\sim$4 at 1000 $^{\circ}$C to $\sim$1.8 at 2000 $^{\circ}$C, in good agreement with the estimates of \cite{LSOK2011} at low temperatures, but without a return to high values at high temperatures, where water contents are lower and prone to large relative errors.
 $D^{ring/wad}_{H_2O}$ decreases from $\sim$1.4 at 1000 $^{\circ}$C to $\sim$1.1 at 2000 $^{\circ}$C.
@@
   \caption{Modelled partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity.
-Note the separate scales for solid/solid and solid/melt partitioning.}
+Note the separate scales for solid/solid and solid/melt partitioning.
+The difference in melt compositions between this study and that of \cite{DDFK2005} results in much lower $D^{wad/melt}$ coefficients.
+Error bars on the wadsleyite datapoints are those reported by \cite{DDFK2005}.}
   \label{fig:partitioning_wad_fo}
@@
   \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity.
-Note the separate scales for solid/solid and solid/melt partitioning.}
+Note the separate scales for solid/solid and solid/melt partitioning.
+Error bars on the ringwoodite data points indicate the range of values reported by \cite{OMY2000}.
+Error bars on the wadsleyite data points indicate 1 s.d.
+uncertainties as reported by \cite{LSOK2011}.}
   \label{fig:partitioning_rw_wad}
@@
 
+\clearpage
 \section{Conclusions}
+
+\begin{itemize}
+\item We report the first experimental results bracketing the hydrous liquidus at conditions existing at the top of the mantle transition zone.
+\item Our results include the first high pressure data on the SiO$_2$-H$_2$O liquidus.
+\item The water-induced depression of melting in silica-rich systems greatly exceeds that of an ideal mixing model at high pressures, and requires significant stabilisation of the melt at relatively low water contents.
+\item Small temperature increases induce large increases in melt productivity as a result of flat liquidus curves. This is especially pronounced for melts in equilibrium with enstatite and stishovite.
+\item We produce the first thermodynamic models for water incorporation into forsterite, wadsleyite and ringwoodite at high pressure, where H$_2$O activity in the melt is much less than one.
+\item Solid/melt partition coefficients are well-constrained by the data, and in good agreement with reported mechanisms of water incorporation in the olivine polymorphs.
+\item Revised solid/solid partition coefficients suggests that small fraction melting may accompany upwelling of peridotite through the wadsleyite-ringwoodite transition.
+\end{itemize}
 

### 2015-08-02 MSH.tex
Update for submission

@@
 Water-rich starting compositions were created using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source.
-As MgO:SiO$_2$ ratios decrease, the liquidus curve develops an increasingly pronounced concave-down topology.
-At slab temperatures, the forsterite- and enstatite-water liquidii have similar water contents.
-Preliminary experiments on the SiO$_2$-H$_2$O binary reveal a pronounced melting point depression up to about 20 mol\% H$_2$O.
-This melting point depression exceeds estimates from published models, and requires either significant non-ideality between species in the melt, or distinct differences in the ease of protonation of bridging oxygens.
-We discuss the implications of this work on the migration of deep-seated slab fluids, and the partitioning of water between nominally anhydrous phases and melts.
+As MgO:SiO$_2$ ratios decrease, the T-$X_{H_2O}$ liquidus curve develops an increasingly pronounced concave-down topology.
+The melting point reduction of enstatite and stishovite at low water contents exceeds that predicted by simple ideal models of hydrogen speciation.
+We discuss the implications of this work on the behaviour of melts in the deep upper mantle and transition zone, and present new models describing the partitioning of water between the olivine polymorphs and associated hydrous melts.
 \end{abstract}
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/assembly}
+      %\includegraphics[width=0.8\textwidth]{figures/assembly}
   \caption{The 18/11 octahedral assembly design used in this study.}
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/Z1058_1400C_13GPa_fo}
+      %\includegraphics[width=0.8\textwidth]{figures/Z1058_1400C_13GPa_fo}
   \caption{BSE image of typical run products.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/periclase}
+      %\includegraphics[width=0.8\textwidth]{figures/periclase}
   \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/forsterite}
+      %\includegraphics[width=0.8\textwidth]{figures/forsterite}
       \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/enstatite}
+      %\includegraphics[width=0.8\textwidth]{figures/enstatite}
   \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/stishovite}
+      %\includegraphics[width=0.8\textwidth]{figures/stishovite}
       \caption{The silica-water phase diagram at 13 GPa, based on current experimental results.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
+      %\includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
   \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
+      %\includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
   \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
+  %\includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
   \caption{Modelled partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity.
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
+  %\includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
   \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity.
@@
 
-\clearpage
-\section{Conclusions}
-
-\begin{itemize}
-\item We report the first experimental results bracketing the hydrous liquidus at conditions existing at the top of the mantle transition zone.
-\item Our results include the first high pressure data on the SiO$_2$-H$_2$O liquidus.
-\item The water-induced depression of melting in silica-rich systems greatly exceeds that of an ideal mixing model at high pressures, and requires significant stabilisation of the melt at relatively low water contents.
-\item Small temperature increases induce large increases in melt productivity as a result of flat liquidus curves. This is especially pronounced for melts in equilibrium with enstatite and stishovite.
-\item We produce the first thermodynamic models for water incorporation into forsterite, wadsleyite and ringwoodite at high pressure, where H$_2$O activity in the melt is much less than one.
-\item Solid/melt partition coefficients are well-constrained by the data, and in good agreement with reported mechanisms of water incorporation in the olivine polymorphs.
-\item Revised solid/solid partition coefficients suggests that small fraction melting may accompany upwelling of peridotite through the wadsleyite-ringwoodite transition.
-\end{itemize}
+\section{Acknowledgements}
+The authors would like to thank Gerti Gollner for acquiring capsule materials, Stefan \"Ubelhack and Heinz Fischer for assembly and capsule cutting and Hubert Schulze for sample preparation.
+RM was supported by a Humboldt Postdoctoral Fellowship.
+This study was funded by the ACCRETE project (European Research Council Advanced Grant, contract number 290568).
 

### 2016-05-10 initial_submission/MSH.tex
Updated to revised version

--- 2015-08-02 MSH.tex
@@
 
-\journal{Geochemica and Cosmochemica Acta}
+\journal{Geochimica et Cosmochimica Acta}
 
@@
 
-The equilibrium constant is related to the energy $\Delta G_{(1)}$ required to protonate two bridging oxygens with a water molecule
+The equilibrium constant is related to the energy $\Delta G_{(1)}$ required to create two terminal OH groups from a bridging oxygen and a water molecule
 \begin{equation}
@@
 
-These models can be used to calculate melt compositions by equating the gibbs free energy of the solid $G^s_{solid}$ with the chemical potential of the component with the composition of that solid in the melt $\mu^s_{melt}$:
+These models can be used to calculate solid-melt equilibria by equating the Gibbs free energy of the solid $G^s_{solid}$ with the chemical potential of the melt component with the same composition $\mu^s_{melt}$:
 \begin{equation}
@@
 \end{equation}
-In this study, we use the thermodynamic dataset of \cite{SLB2011} to calculate $G^s_{solid}$ for periclase, forsterite and its high pressure polymorphs, orthoenstatite, high pressure clinoenstatite, stishovite and coesite.
+In this study, we use the thermodynamic dataset of \cite{SLB2011} to calculate $G^s_{solid}$ for periclase, forsterite and its high pressure polymorphs, orthoenstatite, high pressure clinoenstatite, coesite and stishovite.
 A modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} is used to calculate $G^s_{melt}$, and the \cite{SS1985} model is used to calculate $a^s_{melt}$.
@@
 The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
-At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.}
+At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.
+High temperature data come from the study of \cite{NDMF2015}.}
   \label{fig:foH}
@@
   \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results.
-The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.}
+The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
+High temperature data come from the study of \cite{NDMF2015}.}
   \label{fig:eoH}

### 2016-05-10 initial_submission/cover_letter/cover_letter.tex
Updated to revised version

% cover letter for GCA submission

\documentclass[10pt]{letter}

% Colors
\usepackage[table]{xcolor}
\definecolor{tableShade}{HTML}{F1F5FA}   %iTunes
\definecolor{tableShade2}{HTML}{ECF3FE} %Finder

\definecolor{cambridgedarkblue}{RGB}{0,62,114}
\definecolor{cambridgeblue}{RGB}{0,115,207}
\definecolor{cambridgelightblue}{RGB}{106,173,228}

\definecolor{cambridgedarkorange}{RGB}{200,78,0}
\definecolor{cambridgeorange}{RGB}{227,114,34}
\definecolor{cambridgelightorange}{RGB}{239,189,71}

\definecolor{cambridgedarkgreen}{RGB}{67,81,37}
\definecolor{cambridgegreen}{RGB}{88,166,24}
\definecolor{cambridgelightgreen}{RGB}{168,180,0}

\definecolor{cambridgedarkviolet}{RGB}{65,45,93}
\definecolor{cambridgeviolet}{RGB}{142,37,141}
\definecolor{cambridgelightviolet}{RGB}{181,147,155}

\definecolor{cambridgedarkazur}{RGB}{21,101,112}
\definecolor{cambridgeazur}{RGB}{0,179,190}
\definecolor{cambridgesportingblue}{RGB}{163,193,173}
\definecolor{Grey}{rgb}{0.5, 0.5, 0.5}
\newcommand{\grey}[1]{\textcolor{Grey}{#1}}

\RequirePackage{hyperref}
\hypersetup{pdftex, plainpages = false, pdfpagelabels, 
  pdfpagelayout = useoutlines,dvips, 
  pdfauthor={R. Myhill, myhill.bob@gmail.com},
  pdftitle={Cover letter},
  pdfsubject={Cover letter},
  pdfkeywords={},
  pdfproducer={pdfLaTeX with hyperref},
  pdfcreator={pdfLaTeX}
  bookmarks,
  bookmarksopen = false,
  bookmarksnumbered = true,
  breaklinks = true,
  linktocpage,
  colorlinks = true,
  linkcolor = cambridgeblue,
  urlcolor  = cambridgeviolet,
  citecolor = cambridgedarkorange,
  anchorcolor = cambridgedarkazur,
  hyperindex = true,
  hyperfigures}

\oddsidemargin=.2in
\evensidemargin=.2in
\textwidth=5.9in
\topmargin=-.5in
\textheight=9in

\name{Robert Myhill \\
Bayerisches Geoinstitut \\
Universit\"at Bayreuth \\
95440 Bayreuth \\
Deutschland \\
Tel: +49-(0) 921 55 3746 \\
\href{mailto:myhill.bob@gmail.com}{mailto: myhill.bob@gmail.com}}

\date{\today}

\begin{document}

\begin{letter}{}

\opening{Dear editor,}

Please find attached a manuscript describing our recent work on high pressure hydrous melts, which we hope you will consider for publication as an \emph{Article} in \emph{Geochimica et Cosmochimica Acta}. Our study is entitled \emph{Hydrous melting and partitioning in and above the mantle transition zone: insights from water-rich MgO-SiO$_2$-H$_2$O experiments}, with three authors:

\begin{itemize}
\item Dr Robert Myhill, corresponding author)
\item Prof. Daniel Frost, also from the Bayerisches Geoinstitut
\item Dr Davide Novella, Laboratoire Magmas et Volcans, Universit\'{e} Blaise Pascal
\end{itemize}

% Motivation for your study
Our manuscript describes the first ultrahigh pressure experiments on MgO-SiO$_2$-H$_2$O compositions more water-rich than the brucite-quartz binary. We accomplish this by using a new water source, dihydrogen hexahydroxyplatinate, which breaks down to PtO$_2$ and H$_2$O at high temperatures. Our compositions help us to explore regions of the ternary system which have been traditionally difficult to reach. These regions include the melt compositions expected within the Earth's interior, and so help us to constrain melt productivity within subducting slabs and around the mantle transition zone.

In addition to constraining the liquidus in water-rich systems, we build a simple melt model for the forsterite-water binary that allows us to model H$_2$O activities. Using this data, we construct thermodynamic models for water incorporation into the olivine polymorphs at high pressure.

% Work submitted elsewhere
We confirm that the accompanying manuscript is original work, not published elsewhere or under consideration for publication elsewhere. We have no financial conflicts.

\closing{Best wishes,}

\end{letter}

\end{document}

### 2016-05-10 revised_submission/MSH.tex
Updated to revised version

--- 2016-05-10 initial_submission/MSH.tex
@@
 %% Group authors per affiliation:
-\author{R. Myhill, D. J. Frost}
-\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany}
+\author{R. Myhill}
+\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany \\ now at School of Earth Sciences, University of Bristol, Wills Memorial Building, Queens Road, Bristol BS8 1RJ, United Kingdom}
 \cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
 \ead{myhill.bob@gmail.com}
+
+\author{D. J. Frost}
+\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany}
 
@@
 Hydrous melting at high pressures affects the physical properties, dynamics and chemical differentiation of the Earth.
-However, probing the compositions of hydrous melts at the conditions of the mantle transition zone has traditionally been challenging.
+However, probing the compositions of hydrous melts at the conditions of the deeper mantle such as the transition zone has traditionally been challenging.
 In this study, we conducted high pressure multianvil experiments at 13 GPa between 1200 and 1900 $^{\circ}$C to investigate the liquidus in the system MgO-SiO$_2$-H$_2$O.
 Water-rich starting compositions were created using platinic acid (H$_2$Pt(OH)$_6$) as a novel water source.
-As MgO:SiO$_2$ ratios decrease, the T-$X_{H_2O}$ liquidus curve develops an increasingly pronounced concave-down topology.
+As MgO:SiO$_2$ ratios decrease, the T-$X_{H_2O}$ liquidus curve develops an increasingly pronounced concave-up topology.
 The melting point reduction of enstatite and stishovite at low water contents exceeds that predicted by simple ideal models of hydrogen speciation.
-We discuss the implications of this work on the behaviour of melts in the deep upper mantle and transition zone, and present new models describing the partitioning of water between the olivine polymorphs and associated hydrous melts.
+We discuss the implications of these results with respect to the behaviour of melts in the deep upper mantle and transition zone, and present new models describing the partitioning of water between the olivine polymorphs and associated hydrous melts.
 \end{abstract}
@@
 In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage.
-It has also been argued that neutral or negatively buoyant water-rich melts are to blame for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.
+It has also been argued that neutral or negatively buoyant water-rich melts are responsible for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.
 
 Processes involving deep-seated hydrous melts are controlled to a great extent by melt composition.
-Most importantly, the equilibrium proportion of H$_2$O in melts controls the extent of partial melting for a given bulk H$_2$O content.
+Most importantly, the equilibrium proportion of H$_2$O in melts controls the melt fraction for a given bulk H$_2$O content.
 Together with solid-solid-melt dihedral angles, the extent of partial melting governs the permeability and therefore mobility of melt, and the ability to detect it via body wave seismology and electrical conductivity measurements.
-Composition is also a key variable in determining melt density and viscosity.
-Finally, the change in equilibrium melt compositions with pressure and temperature has a significant influence on melt channelisation, which reduces melt-solid interaction and increases ascent rates.
+Melt densities and viscosities are also governed by their composition.
+Finally, changes in equilibrium melt composition with pressure, temperature and bulk composition can enhance melt channelisation, which reduces melt-solid interaction and increases ascent rates \citep[e.g.][]{SKA2001,WK2012}.
 
 Melt compositions are also required to understand the stability of dense hydrous magnesium silicate (DHMS) phases and the water capacity of nominally anhydrous minerals (NAMS) at high pressure.
-Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth.
+Together, these control water transport and storage in the deep mantle, and as such are the primary inputs to models of the water budget of the Earth \citep{Hirschmann2006}.
 Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O.
-As pressure increases, immiscibility between fluid and melt decreases, and the thermal stability of hydrous phases increases.
+As pressure increases, immiscibility between fluid and melt decreases \citep{XXXX}, and the thermal stability of hydrous phases increases \citep{XXXX}.
 As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one.
@@
 The primary problem is that water-rich melts are unquenchable.
-During quench, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour on decompression and sample preparation.
+During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour on decompression and during sample preparation.
 The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}.
-Mass balance alone depends on knowing the volume proportions of solid phases and quench phases, being able to recognise and subtract quench overgrowths where they exist, and knowing the effective ambient-pressure density of the melt.
+Mass balance alone depends on being able to accurately estimate the volume proportions of solid phases and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt.
 All of these are associated with significant uncertainties.
-Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit would imply a large water content).
+Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit implies a large water content).
 However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam.
@@
 
-Bracketing the liquidus requires only that the bulk composition during the experiment is known.
-This technique therefore seems a good alternative to estimating melt compositions ex-situ.
-The only problem is that at slab or ambient mantle temperatures, water contents in melts at high pressure are very high.
-Making up solid starting compositions with more water than a brucite-quartz mix has traditionally been avoided.
-Existing studies either extrapolate high temperature liquidus curves to high water contents or add water in liquid form if the sample chambers are sufficiently large \citep[e.g.][]{HM2012}.
-Extrapolation is prone to large errors, and accurately adding liquid water to the small volumes of high pressure capsules is essentially impossible.
-In this study, we use high purity platinic acid, hexahydroxyplatinate(IV) (H$_2$Pt(OH)$_6$), as a novel water source.
-Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}.
-It breaks down at higher temperatures, releasing four molecules of H$_2$O and forming platinum (IV) oxide PtO$_2$.
-Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments.
+Here we use the approach of bracketing the liquidus surface to determine melt composition as a function of temperature in the system MgO-SiO$_2$-H$_2$O (MSH).
+This technique requires knowledge of the starting composition, techniques to effectively minimise water loss from the capsule and an ability to recognise the difference between liquidus and quench phases.
+We use this technique to develop thermodynamic models for water solubility in melts, and apply these to models of water partitioning between mantle phases and melts.
+We then apply these models to better understand hydrous melting under transition zone conditions.
+%XXXX what, in general, have you done in this paper?
 
 \section{Experimental and analytical methods}
-Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (H$_2$Pt(OH)$_6$).
+Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (hexahydroxyplatinate(IV); H$_2$Pt(OH)$_6$).
+We use platinic acid to increase water contents to those expected of hydrous melts in equilibrium with mantle phases under subduction/convecting mantle conditions.
+Platinic acid is a pale yellow compound which is stable to about 130 $^{\circ}$C \citep{Nagano2002}.
+Its relatively high stability, high water content and the inert nature of the breakdown product (in a system where redox reactions are absent) make it an excellent water source for high pressure hydrous melting experiments.
 Before weighing, quartz was dried at 1000 $^{\circ}$C for 12 hours, while brucite was heated to 250 $^{\circ}$C, also over 12 hours.
@@
 Powders were weighed and ground dry in an agate mortar for 30 minutes, using a mask, goggles and gloves to avoid physical contact with the platinic acid.
-Starting compositions were stored in glass vials in a vacuum desiccator.
-Compositions are listed in Table \ref{table:compositions}.
+Prepared powders were stored in glass vials in a vacuum desiccator until use.
+
+We chose starting compositions to cover the brucite-quartz-water subregion of the MSH ternary.
+Chosen compositions include points on the binaries between water and brucite, forsterite, enstatite and stishovite, and a few additional compositions to constrain the locations of cotectics (Table \ref{table:compositions}).
 
@@
 The temperature was recorded using a W$_{97}$Re$_{3}$--W$_{75}$Re$_{25}$ (Type D) thermocouple inserted axially into the assembly.
-To avoid water loss, higher temperature runs were heated for a shorter duration.
+To avoid water loss, higher temperature runs were heated for a shorter duration \citep[see][]{NDMPMF2016}.
 The experiments were quenched by cutting power to the heater.
@@
 \subsection{Background}
-Oxygen in neutral MSH melt species can exist either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group.
-The general equation describing the reaction between melt species is
+Oxygen in neutral MSH melt species exists either as molecular water, as a bridging oxygen between magnesium and/or silicon atoms, or as part of a terminal hydroxyl group (terminal oxygens would require Si=O double bonds, which are extremely unstable).
+The general equation describing the reaction between melt species is therefore
 \begin{equation}
@@
 These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers.
-A high proportion of monomers and dimers exist in relatively dilute solutions, but not in the concentrated solutions investigated in this study.
+A high proportion of monomers and dimers exist in relatively dilute solutions \citep{XXXX}, but not in the concentrated solutions investigated in this study.
 
-A number of models have been presented to investigate the behaviour of concentrated hydrous melts.
+A number of models have been presented to investigate the behaviour of water-bearing silicate melts.
 \cite{SS1985} made the assumption that H$_2$O, O$_{br}^{2-}$ and OH$_{tm}^-$ mix ideally, with a parameter $r$ representing the number of oxygen atoms in each formula unit which are available for protonation.
@@
 \subsection{The anhydrous melt model}
+\subsubsection{Endmembers}
 Thermodynamic models of MgO-SiO$_2$ melts are currently in their infancy.
@@
 This leads to a periclase melting point of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}.
-We notes that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}.
+We note that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}.
 Current thermodynamic models for periclase cannot be extrapolated to the extreme temperatures predicted by the \cite{ZF2008} data.
@@
 
-The MD-determined entropy of stishovite is significantly lower than the experimentally determined value, so we increased the standard state entropy of the melt by 16.5 J/K/mol.
+The MD-determined entropy of stishovite is significantly lower than the experimentally determined value \citep{SLB2011}, so we increased the standard state entropy of the melt by 16.5 J/K/mol.
 Conveniently, we also have a way to test the volumes predicted by the model of \cite{DKS2013} at around 13 GPa.
@@
 Unfortunately, the MD-derived SiO$_2$ melt model has its density crossover with coesite about 4 GPa lower.
-Therefore, we adjusted the bulk modulus and reference pressure so satisfy this constraint, and the volume at zero pressure.
-The adjustments required here may not be a failing of the MD simulations; it is likely that the coordination changes in the SiO$_2$ melt require a higher order Birch Murnaghan EoS than that used by \citep{DKS2013}.
+Therefore, we adjusted the bulk modulus and reference pressure to satisfy this constraint, and the volume at zero pressure.
+The adjustments required here may not be a failing of the MD simulations; it is likely that the coordination changes in the SiO$_2$ melt require a higher order Birch Murnaghan EoS than that used by \cite{DKS2013}.
 Our adjusted model is probably only accurate in a range of about $\pm$5 GPa around the triple point.
@@
 
+\subsubsection{Solution model}
 A subregular solution model was adopted to describe the free energy of intermediate-composition MgO-SiO$_2$ melts.
-The two interaction parameters W$_{MS}$ and W$_{SM}$ at 14 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions \citep{PG1990, PW1993}.
+The two interaction parameters W$_{MS}$ and W$_{SM}$ at 13 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions \citep{PG1990, PW1993}.
 We assumed an ideal entropic contribution to mixing.
@@
 
+\begin{equation}
+p' = \frac{p_{H_2O}}{p_{H_2O}  + r \left(1.-p_{H_2O} \right)} 
+\end{equation}
 \begin{eqnarray}
-p' = \frac{p_{H_2O}}{p_{H_2O}  + r \left(1.-p_{H_2O} \right)} \\
 X_O = 1 - p' - \frac{ 0.5 - \sqrt{ 0.25 - \frac{K_{(1)}-4}{K_{(1)}} \left(p'-p'^2 \right)}}{\frac{K_{(1)}-4}{K_{(1)}} }\\
@@
 Stishovite also formed equant to blocky grains, but crystals only separated from the melts in chambers with a high water content.
-In addition to the difference in texture and shape between quench crystals liquidus phases, subliquidus phases typically contained many small inclusions of Pt-Rh oxides.
+In addition to the difference in texture and shape between quench crystals and liquidus phases, subliquidus phases typically contained many small inclusions of Pt-Rh oxides.
 A typical set of run products is shown in Figure \ref{fig:sem}.
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/assembly}
+      \includegraphics[width=0.8\textwidth]{figures/assembly}
   \caption{The 18/11 octahedral assembly design used in this study.}
@@
 Compositions are listed in order of increasing molar H$_2$O content.
-Unused compositions for each run are marked by an en-dash.
+Unused compositions for each run are marked by a dash.
 Minor solids are listed in brackets.
-Mineral abbreviations are as follows: br - brucite, per - periclase, co - hydroxychondrodite, en - clinoenstatite, s - stishovite.}
+Mineral abbreviations are as follows: br - brucite, per - periclase, co - hydroxychondrodite, en - clinoenstatite, s - stishovite.
+Melts are present in all run products.}
 \label{table:experiments}
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/Z1058_1400C_13GPa_fo}
+      \includegraphics[width=0.8\textwidth]{figures/Z1058_1400C_13GPa_fo}
   \caption{BSE image of typical run products.
@@
 Experimental results for the system MgO-H$_2$O are shown in Figure \ref{fig:MH}.
-The periclase-brucite-water invariant at 13 GPa is well constrained at 62--66 mol\% H$_2$O and $\sim$1200 $^{\circ}$C.
-The MgO content of the fluid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase.
+The periclase-brucite-water invariant at 13 GPa is well constrained at 62--66 mol\% H$_2$O and $\sim$1250 $^{\circ}$C.
+The MgO content of the liquid increases gradually with increasing pressure, as a result of the extremely high melting temperature of periclase.
 In addition to the thermodynamic models described in the methods section, the activity of H$_2$O at the brucite-periclase-melt triple point is calculated using the thermodynamic dataset of \citep{HP2011}, which includes models for brucite, periclase and water \citep{PS1995}.
@@
 \end{equation}
-\noindent where the right hand side is the chemical potential of Mg(OH)$_2$ in the melt, and the right hand side is the Gibbs free energy of brucite with a correction for the different standard state energies of MgO between the \cite{HP2011} and \cite{SLB2011} datasets.
+\noindent where the left hand side is the chemical potential of Mg(OH)$_2$ in the melt, and the right hand side is the Gibbs free energy of brucite with a correction for the different standard state energies of MgO of the \cite{HP2011} and \cite{SLB2011} datasets.
 $G_{(1)} = RT \ln K_{(1)} = -75000$ J/mol ($r$=1) provides a good fit to the data if additional asymmetry is added via a binary subregular model (W$_{MgO-H_2O}$=55000 J/mol, W$_{H_2O-MgO}$=0 J/mol).
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/periclase}
+      \includegraphics[width=0.8\textwidth]{figures/periclase}
   \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results.
@@
 They reported that the stability of brucite reaches a maximum at about 9--10 GPa and 1200 $^{\circ}$C, decomposing at 1100--1150 $^{\circ}$C at 13 GPa.
-The temperature difference between their study and ours is therefore 50 $^{\circ}$C, within experimental uncertainties.
+The temperature difference between their study and ours is thus 50 $^{\circ}$C and therefore within experimental uncertainties.
 
@@
 The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}.
-The model parameters shown are $G_{(1)} = 75 \cdot (1420-T)$ J/mol, $r=4/3$.
-The anhydrous melting of forsterite is incongruent, so the high temperature part of the liquidus is metastable with respect to a mixture of periclase and liquid (see also Figure \ref{fig:ternary}).
+The model parameters shown are $G_{(1)} = 75 \cdot (1200-T)$ J/mol, $r=4/3$.
+The anhydrous melting of forsterite is incongruent, so at low water contents the liquidus phase along the Mg$_2$SiO$_4$-H$_2$O binary is periclase (see also Figure \ref{fig:ternary}).
 Water-rich chambers from experiments run at 1300 $^{\circ}$C revealed small quantities of enstatite, indicating that the forsterite-enstatite cotectic crosses the Mg$_2$SiO$_4$-H$_2$O tie line at 1300-1400 $^{\circ}$C.
@@
 At 1200 $^{\circ}$C, one chamber contained large crystals of hydroxychondrodite.
-The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained an MgO-deficit in forsterite resulting from the incorporation of hydrogen via the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}.
+The presence of this MgO-rich phase in the chamber with the lowest water contents may be explained by an MgO-deficit in forsterite resulting from the incorporation of hydrogen via the substitution mechanism Mg$^{2+}$ $\rightarrow$ 2H$^+$ \citep{KB2006}.
 Alternatively it could simply be the result of a small SiO$_2$ deficit in that sample chamber.
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/forsterite}
+      \includegraphics[width=0.8\textwidth]{figures/forsterite}
       \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results.
@@
 At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.
-High temperature data come from the study of \cite{NDMF2015}.}
+High temperature data come from the study of \cite{NDMPMF2016}.}
   \label{fig:foH}
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/enstatite}
+      \includegraphics[width=0.8\textwidth]{figures/enstatite}
   \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results.
 The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
-High temperature data come from the study of \cite{NDMF2015}.}
+High temperature data come from the study of \cite{NDMPMF2016}.}
   \label{fig:eoH}
@@
 Our experimentally determined liquidus is in excellent agreement with the experimental results of \cite{YII2004}, as is the absence of stishovite in any of the sample chambers.
-We note that although we do not have any evidence for melt-fluid immiscibility, the flat liquidus at ca.
-1500 $^{\circ}$C causes a very rapid change from water-rich to water-poor fluids with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}.
-The implication is that for a given bulk H$_2$O content, a sharp rise in the degree of partial melting will occur at about 1500 $^{\circ}$C in hydrous MgSiO$_3$ melts.
+We note that although we do not observe any evidence for melt-fluid immiscibility, the flat liquidus at ca.
+1500 $^{\circ}$C causes a very rapid change from water-rich to water-poor compositions with increasing temperature, in agreement with \cite{Inoue1994} and \cite{YII2004}.
+One possibility is that this reflects relatively high water loss from the water rich bulk compositions.
+However, there is very little evidence for melt loss from the capsules; the lids on the capsules were well-sealed and showed no evidence of melt escape.
+The implication is that for a given bulk H$_2$O content, a sharp rise in the degree of partial melting occurs when bulk compositions along the en+H$_2$O join are heated above 1500 $^{\circ}$C.
 
@@
 13.5 GPa.
-A third possibility is that approximately 30\% volume liquidus-phase stishovite crystals were missed during SEM analysis.
+A third possibility is that liquidus-phase stishovite crystals were missed during SEM analysis.
+In the most extreme cases, the missing fraction of crystals would have been on the order of 30\%.
 This is also extremely unlikely; the equant and dendritic morphologies of liquidus and quench stishovite are instantly recognisable, and the thermal gradients in the capsule lead to melt separation on the timescales of the experiments, such that even a small (5\%) volume proportion of solids should be detected.
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/stishovite}
+      \includegraphics[width=0.8\textwidth]{figures/stishovite}
       \caption{The silica-water phase diagram at 13 GPa, based on current experimental results.
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
+      \includegraphics[width=0.8\textwidth]{figures/experimental-ternary}
   \caption{A preliminary ternary liquidus diagram at 13 GPa, based on current experimental results.
-Contours are in $^{\circ}$C, and mineral abbreviations are liquidus fields as follows: per - periclase, fo - forsterite, en - clinoenstatite, stv - stishovite, coe - coesite, phB - phase B, br - brucite, chond - hydroxychondrodite, ice - ice VII.}
+Contours are in $^{\circ}$C, and mineral abbreviations are liquidus fields as follows: per - periclase, fo - forsterite, en - clinoenstatite, stv - stishovite, coe - coesite, phB - phase B, br - brucite, chond - hydroxychondrodite, ice - ice VII.
+The locations of the contours are constrained by observed liquidus phases and the liquidus-bracketing results presented here.
+Previous work on hydrous phase stability and Mg:Si ratios for cotectic melts are also included \citep{MSUP2007}.}
   \label{fig:ternary}
@@
 However, the flat liquidus curves revealed by our experiments are likely to also be observed in more natural chemical systems.
-The current results imply that the water contents of melts in equilibrium with forsterite and enstatite drop from 62 mol\% (34 wt\%) to around 30 mol\% (16 wt\%) between 1300 and 1700 $^{\circ}$C.
-For a given water content, this rapid change in composition must be balanced by a rapid increase in the mass fraction of partial melt, which could greatly enhance chances of detection by seismological or other geophysical techniques.
-Such large decreases in water content would also lead to large density changes; close to slabs, hydrous melts could be buoyant, while in warmer mantle of a similar composition they would be dense.
-This scenario might be expected to result in channelisation of hydrous melts close to the top interface of subducting slabs.
-Finally, large decreases in water contents are also likely to have a profound effect on melt viscosity, and may encourage reaction-infiltration instabilities encouraging melt channelisation \citep{SKA2001}.
+The current results imply that the water contents of melts in equilibrium with forsterite and enstatite drop from 62 mol\% (34 wt\%) to around 30 mol\% (16 wt\%) over a temperature interval of about 400 degrees between 1300 and 1700$^{\circ}$C.
+At $\sim$400 km depth, the interiors of slabs could be as cold as 600-800$^{\circ}$C \citep{Frohlich2006}, while surrounding mantle has a temperature of $\sim$1400$^{\circ}$C \citep{Anderson1982}.
+Mantle plumes have temperatures up to $\sim$1700$^{\circ}$C.
+The large increases in silicate content in melts at the top of the transition zone would lead to large density and viscosity changes; close to slabs, hydrous melts could be buoyant and of relatively low viscosity, while in warmer mantle of a similar composition they would be dense and more viscous.
+Dissolution of silicate material into melt would also promote channelisation of hydrous melts close to the top interface of subducting slabs \citep{SKA2001,WK2012}, which could increase seismic visibility by significantly increasing melt fraction and connectivity of melt.
 
@@
 In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA.
-In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to hydrogen.
+In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to the fraction of oxygen bonded to hydrogen.
 In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits.
@@
 We note that two of the three melt compositions in equilibrium with ringwoodite fall above the temperatures indicating that hydrogen exists only as molecular H$_2$O \citep{SS1985}.
-In addition, the calculated liquidus temperatures in the wadsleyite studies at 15 GPa are lower than those reported at 5.5 GPa \citep{Inoue1994}.
-Both of these observations appear highly unlikely.
+In addition, the calculated liquidus temperatures in the wadsleyite studies at 15 GPa are lower than those reported for olivine at 5.5 GPa \citep{Inoue1994}.
+Both of these observations, therefore, raise inconsistencies which are difficult to address within a self-consistent thermodynamic model.
 
@@
   \centering
-      %\includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
+      \includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
   \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
@@
 
-We now create thermodynamic models for incorporation of H$_2$O into forsterite, wadsleyite and ringwoodite.
+We now create thermodynamic models for the incorporation of H$_2$O into forsterite, wadsleyite and ringwoodite.
 For forsterite, we use high pressure (12--14 GPa) experimental data on the solids from \cite{SFNHB2006} and \cite{LSKO2009}; neither of these studies estimate coexisting melt compositions.
@@
 \begin{tabular}{lllll}
-Phase & $P_{\textit{\tiny{ref}}}$ (Pa) & A [kg/(kgPa)] & $\Delta H_{P_{\textit{\tiny{ref}}}}$ [J/mol] & $\Delta V$ [m$^3$/mol] \\
+Phase & $P_{\textit{\tiny{ref}}}$ ($10^{10}$ Pa) & A [$10^{-12}$ kg/(kgPa)] & $\Delta H_{P_{\textit{\tiny{ref}}}}$ [$10^{5}$ J/mol] & $\Delta V$ [$10^{-5}$ m$^3$/mol] \\
 \hline
@@
 % Olivine & 1.2e10 & 5e-12 & 1.70e5 & 1e-5 \\ Zhao et al., 2004, water contents much too high
-Olivine & 1.2e10 & 3.23e-12 & 1.50e5 & 1e-5 \\
-Wadsleyite & 1.5e10 & 1.97e-12 & 1.60e5 & 1e-5 \\
-Ringwoodite & 2.0e10 & 1.46e-12 & 2.03e5 & 1e-5 (fixed)
+Olivine & 1.2 & 3.23 & 1.50 & 1 \\
+Wadsleyite & 1.5 & 1.97 & 1.60 & 1 \\
+Ringwoodite & 2.0 & 1.46 & 2.03 & 1 (fixed)
 \end{tabular}
@@
 
-These models allow us to predict the partition coefficients of water between olivine, wadsleyite and melt at the 410 km discontinuity ($\sim$14 GPa; Figure \ref{fig:partitioning_rw_wad}), and wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa; Figure \ref{fig:partitioning_rw_wad}).
+These models allow us to predict the partition coefficients of water between olivine, wadsleyite and melt at the 410 km discontinuity ($\sim$14 GPa; Figure \ref{fig:partitioning_wad_fo}), and wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa; Figure \ref{fig:partitioning_rw_wad}).
 $D^{wad/fo}_{H_2O}$ displays a decrease from $\sim$4 at 1000 $^{\circ}$C to $\sim$1.8 at 2000 $^{\circ}$C, in good agreement with the estimates of \cite{LSOK2011} at low temperatures, but without a return to high values at high temperatures, where water contents are lower and prone to large relative errors.
 $D^{ring/wad}_{H_2O}$ decreases from $\sim$1.4 at 1000 $^{\circ}$C to $\sim$1.1 at 2000 $^{\circ}$C.
-These values imply that upwelling across both the 410 and the 520 km discontinuity will cause partial melting.
+These values imply that partial melting will accompany upwelling of H$_2$O-saturated ringwoodite and wadsleyite across both the '520'- and the '410'-km discontinuities.
 Both boundaries have been associated with low seismic velocities, which may indicate zones of deep melting \citep[e.g.][]{JDH2010}.
@@
   \centering
-  %\includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
-  \caption{Modelled partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity.
+  \includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
+  \caption{Modelled H$_2$O partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity.
 Note the separate scales for solid/solid and solid/melt partitioning.
@@
   \centering
-  %\includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
-  \caption{Modelled partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity.
+  \includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
+  \caption{Modelled H$_2$O partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity.
 Note the separate scales for solid/solid and solid/melt partitioning.
@@
 \end{figure}
+
+\section{Concluding Remarks}
+% Summary
+In this study, we have conducted multi-anvil experiments to constrain the topology of the liquidus in the MgO-SiO$_2$-H$_2$O system at 13 GPa.
+We employ a simple thermodynamic model to predict the activities of water in hydrous melts in equilibrium with the mantle, and to create models for incorporation of water into forsterite, wadsleyite and ringwoodite.
+
+% Future work
+We hope that this work forms the foundation for more detailed thermodynamic models of hydrous mantle melts.
+Particularly intriguing are the concave up topologies of the forsterite, enstatite and stishovite liquidus surfaces, and the large depression of the melting point at low water contents in silica-rich systems.
+Both of these observations imply efficient hydroxylation of the silicate network and strong interactions between species, which would be interesting subjects for molecular dynamic studies.
+We also hope that a combination of our liquidus constraints with parameterisations for density and viscosity will be used to inform more realistic models of melt migration above the mantle transition zone and during subduction.
 

### 2016-05-11 revised_submission/MSH.tex
update

@@
 \author{R. Myhill}
-\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany \\ now at School of Earth Sciences, University of Bristol, Wills Memorial Building, Queens Road, Bristol BS8 1RJ, United Kingdom}
+\address{Bayerisches Geoinstitut, Universit\"{a}t Bayreuth, Universit\"{a}tsstrasse 30, 95447 Bayreuth, Germany \\ Now at School of Earth Sciences, University of Bristol, Wills Memorial Building, Queens Road, Bristol BS8 1RJ, United Kingdom}
 \cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
@@
 Melting at subduction zones is primarily driven by the release of hydrous fluids from the warming slab at about 100 km depth, producing the world's arc volcanoes.
-The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to penetrate into the mantle transition zone or deeper, providing it remains low enough in temperature \citep{KOM2005}.
+The water cycle does not stop here, however; hydrous minerals within the crust should allow some water to be transported to 250 km \citep{PS2002}, and peridotite hydration should allow water to reach the mantle transition zone or deeper, providing temperatures remain sufficiently low \citep{KOM2005}.
 This deep water cycle must extend to at least 500 km depth, as implied by the discovery of water-saturated ringwoodite in a diamond from Juina, Brazil \citep{Pearsonetal2014}.
-In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage.
+In addition to kimberlites and arc volcanoes, hydrous melts are known to metasomatise and refertilise the lithospheric mantle, creating the so-called MARID (mica, amphibole, rutile, ilmenite, diopside) assemblage \citep{GHKR2010}.
 It has also been argued that neutral or negatively buoyant water-rich melts are responsible for low velocity and high conductivity layers observed at the 410 km discontinuity, and that these small fraction melts could act to filter out incompatible elements during upwelling \citep{BK2003}.
@@
 Hydrous fluids at elevated pressures and temperatures cannot be treated as pure H$_2$O.
-As pressure increases, immiscibility between fluid and melt decreases \citep{XXXX}, and the thermal stability of hydrous phases increases \citep{XXXX}.
-As a result, fluids/melts released by the breakdown of hydrous phases have H$_2$O activities significantly lower than one.
+As pressure increases, immiscibility between fluid and melt decreases \citep{MSUP2007, SUTG2001, HM2012}, and the thermal stability of Mg-rich hydrous phases increases \citep{KO2006}.
+As a result, the high pressure breakdown of hydrous phases releases fluids which have H$_2$O activities significantly lower than one.
 Understanding and predicting the \emph{P-T} conditions of DHMS phase breakdown and the water storage capacity at of NAMS therefore requires good activity-composition models for hydrous melts.
 
-Constraining melt compositions remains a major challenge in high-pressure experimental petrology.
+Constraining melt compositions remains a major challenge for the field of high-pressure experimental petrology.
 The primary problem is that water-rich melts are unquenchable.
-During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour on decompression and during sample preparation.
+During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour during decompression and subsequent sample preparation.
 The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}.
-Mass balance alone depends on being able to accurately estimate the volume proportions of solid phases and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt.
-All of these are associated with significant uncertainties.
+Using mass balance to calculate H$_2$O contents depends on being able to estimate the volume proportions of solid and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt.
+Such estimates are typically associated with significant uncertainties.
 Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit implies a large water content).
@@
 These equations can be used to describe in an average sense a range of melt-melt equilibria involving monomers (Mg(OH)$_2$ and Si(OH)$_4$), dimers (Mg$_2$O(OH)$_2$, MgSiO(OH)$_4$, Si$_2$O(OH)$_6$) and higher oligomers.
-A high proportion of monomers and dimers exist in relatively dilute solutions \citep{XXXX}, but not in the concentrated solutions investigated in this study.
+Concentrated solutions such as those investigated in this study are likely to be dominated by highly polymerised oligomers \citep[see][]{HM2012}.
 
@@
 A modified version of the molecular dynamics-derived anhydrous melt model of \cite{DKS2013} is used to calculate $G^s_{melt}$, and the \cite{SS1985} model is used to calculate $a^s_{melt}$.
-The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/}][]{CHRU2014}.
+The calculations in the work were undertaken with the mineral physics toolkit Burnman \citep[available from \url{https://geodynamics.org/cig/software/burnman/};][]{CHRU2014}.
 
-\subsection{The anhydrous melt model}
-\subsubsection{Endmembers}
+\subsection{The hydrous melt model}
+\subsubsection{Anhydrous melt endmembers}
 Thermodynamic models of MgO-SiO$_2$ melts are currently in their infancy.
-Many of the techniques used to constrain the thermodynamic properties (particularly the equations of state) of solids cannot be used on liquids as a result of their highly dynamic behaviour.
-Therefore in this study, we use MgO and SiO$_2$ melt models derived from molecular dynamics simulations \citep{DKS2013}, and fit a simplified mixing curve based on experimental data.
+Many of the experimental techniques used to constrain the thermodynamic properties (particularly the equations of state) of solids cannot be used on liquids as a result of their highly dynamic behaviour.
+In this study, we instead use MgO, Mg$_2$SiO$_4$, MgSiO$_3$ and SiO$_2$ melt models derived from molecular dynamics simulations \citep{DKS2013}.
 First, however, the thermodynamic properties of the endmember melts need to be adjusted to fit the experimentally derived thermodynamic data at the melting point.
@@
 This leads to a periclase melting point of $\sim$ 4180 K at 13 GPa, similar to that determined by previous ab-initio studies \citep{Alfe2005}.
-We note that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa][]{ZB1994,ZF2008}.
-Current thermodynamic models for periclase cannot be extrapolated to the extreme temperatures predicted by the \cite{ZF2008} data.
-However, assuming that the volume change of melting is roughly constant with temperature (1.8 cm$^3$/mol at 13 GPa), the entropy of melting at a melting point of 5373 K would be about 50\% of that for a melting point of 4180 K (13.25 vs.
+We note that experimental results vary by about 1000 K about this value \citep[3100 -- 5373 K at 13 GPa;][]{ZB1994,ZF2008}.
+It is therefore worth considering the effect of different melting curves on the free energy difference between melt and solid.
+Assuming that the volume change of melting is roughly constant with temperature (1.8 cm$^3$/mol at 13 GPa), the entropy of melting at a melting point of 5373 K would be about 50\% of that for a melting point of 4180 K (13.25 vs.
 27.5 J/K/mol).
-At 13 GPa and 2000 K, MgO melt would actually be 15 kJ/mol \emph{more} stable as a result of this lower entropy.
+The difference in free energy would therefore increase more slowly as temperature decreased, reducing the discrepancy between the two models at relatively low temperatures.
 
-The MD-determined entropy of stishovite is significantly lower than the experimentally determined value \citep{SLB2011}, so we increased the standard state entropy of the melt by 16.5 J/K/mol.
-Conveniently, we also have a way to test the volumes predicted by the model of \cite{DKS2013} at around 13 GPa.
-The experimentally-derived slope of the coesite melting curve is essentially zero at the triple point with stishovite at 13.7 GPa \citep{ZLGHF1993}.
-Therefore, from the Clausius-Clapeyron relationship $dT/dP = \Delta V / \Delta S$ the volumes of coesite and SiO$_2$ melt must be zero at this point.
-Unfortunately, the MD-derived SiO$_2$ melt model has its density crossover with coesite about 4 GPa lower.
-Therefore, we adjusted the bulk modulus and reference pressure to satisfy this constraint, and the volume at zero pressure.
-The adjustments required here may not be a failing of the MD simulations; it is likely that the coordination changes in the SiO$_2$ melt require a higher order Birch Murnaghan EoS than that used by \cite{DKS2013}.
+The treatment for Mg$_2$SiO$_4$ and MgSiO$_3$ was similar to that for periclase; the standard state energies of the liquids were adjusted to fit the melting points at 13 GPa for forsterite \citep[2290 $^{\circ}$C][]{PW1993} and HP-clinoenstatite \citep[2240 $^{\circ}$C][]{PG1990}, using the endmember models of \cite{SLB2011}.
+
+In the case of SiO$_2$ melt, we used the experimentally-derived melting curve to assess the \cite{DKS2013} model.
+The slope of the coesite melting curve is essentially zero at the triple point with stishovite at 13.7 GPa \citep{ZLGHF1993}, such that $V$ (coesite) = $V$ (SiO$_2$ melt) at this point.
+The slope of the stishovite melting curve at the triple junction \citep{SL1995} then provides an estimate of the entropy of melting (via the Clausius-Clapeyron relationship $dT/dP = \Delta V / \Delta S$), which is $\sim$40 J/K/mol.
+We note that the stishovite melting curve recently proposed by \cite{Millotetal2015} is much shallower than that constrained by experiments at 15--100 GPa \citep{SL1995, LAM1983}.
+Their two-parameter Simon's equation fit between the melting points at 13.7 and 500 GPa is probably inappropriate over such extreme pressure ranges, especially where the melting point may be rather sensitive to changes in cation coordination.
+
+At the triple point, the MD-derived entropy of melting is in excellent agreement with the experimental value, but the entropy of stishovite is 16.4 J/K/mol lower than the experimentally determined value \citep{SLB2011}.
+For this reason, we increased the standard state entropy of the SiO$_2$ melt by 16.4 J/K/mol.
+We also adjusted the bulk modulus and reference pressure to obtain the correct pressure at the coesite-melt density crossover.
+The adjustments required here may not be a failing of the MD simulations; the coordination changes in the SiO$_2$ melt may require a higher order EoS than that used by \cite{DKS2013}.
 Our adjusted model is probably only accurate in a range of about $\pm$5 GPa around the triple point.
-The experimental data can also be used to judge the accuracy of the entropy of the SiO$_2$ melt model.
-The MD-derived entropy of melting is $\sim$40 J/K/mol at the triple point.
-This value is in perfect agreement with the value derived from the Clapeyron slope of the stishovite liquidus \citep{SL1995}.
-We note that the slope recently proposed by \cite{Millotetal2015} is much shallower, but their slope is a two parameter Simon's equation fit based on the triple point temperature at 13.7 GPa and the melting point at 500 GPa, and fits the data at 15--100 GPa \citep{SL1995, LAM1983} rather poorly.
-A Simon's equation fit is probably inappropriate over such extreme pressure ranges, especially where the melting point may be rather sensitive to changes in cation coordination.
 
-\subsubsection{Solution model}
-A subregular solution model was adopted to describe the free energy of intermediate-composition MgO-SiO$_2$ melts.
-The two interaction parameters W$_{MS}$ and W$_{SM}$ at 13 GPa were derived from the melt composition at the forsterite (fo)-HP clinoenstatite (hen) eutectic \citep{PWMW1998} and melting temperatures at the fo, hen and fo-hen cotectic compositions \citep{PG1990, PW1993}.
-We assumed an ideal entropic contribution to mixing.
-The pressure dependencies were fit to the melting curves.
-It should be noted that although the excess volumes are similar to the results of \cite{DKS2013}, the excess entropies are some 50\% lower.
-However, they provide an excellent fit to the forsterite and orthoenstatite melting curves (see Supplementary Materials).
-
-\subsection{The hydrous melt model}
+\subsubsection{The hydrous melt model}
 
@@
 We assume that all bridging oxygens are equally likely to be protonated.
-The number of oxygens per formula unit $r$ thus increases monotonously from 1 to 2 across the MgO-SiO$_2$ binary ($r_{\textrm{per}}$ = 1, $r_{\textrm{fo, wad, rw}}$ = 4/3, $r_{\textrm{oen, cen}}$ = 3/2, and $r_{\textrm{coe, stv}}$ = 2).
+The number of oxygens per formula unit $r$ thus increases monotonously from 1 to 2 across the MgO-SiO$_2$ binary ($r_{MgO}$ = 1, $r_{Mg_2SiO_4}$ = 4/3, $r_{MgSiO_3}$ = 3/2, and $r_{SiO_2}$ = 2).
 The ideal activities of the anhydrous and water components can then be calculated:
@@
 \noindent where the subscripts $i$ and $j$ correspond to the components $s$ or H$_2$O.
-Equilibrium with the solid can now be found using Equation \ref{eqn:equilibrium}, noting that the dataset-derived values of $G^s_{solid}$ of the forsterite (Mg$_2$SiO$_4$) and enstatite (Mg$_2$Si$_2$O$_6$) polymorphs should be divided by 3 and 4 to reduce them to their one-cation equivalents.
+Equilibrium with the solid can now be found using Equation \ref{eqn:equilibrium}, noting that the dataset-derived values of $G^s_{solid}$ of the forsterite (Mg$_2$SiO$_4$) and enstatite (Mg$_2$Si$_2$O$_6$) polymorphs \citep{SLB2011} should be divided by 3 and 4 to reduce them to their one-cation equivalents.
+Similarly, the values of $G^s_{melt}$ for the MD-derived Mg$_2$SiO$_4$ and MgSiO$_3$ models \citep{DKS2013} should be divided by 3 and 2 respectively.
 
@@
 The melting point depression of forsterite in the presence of water at 13 GPa is illustrated in Figure \ref{fig:foH}.
-The model parameters shown are $G_{(1)} = 75 \cdot (1200-T)$ J/mol, $r=4/3$.
+The model parameters shown are $G_{(1)} = 75 \cdot (1420-T)$ J/mol, $r=4/3$.
 The anhydrous melting of forsterite is incongruent, so at low water contents the liquidus phase along the Mg$_2$SiO$_4$-H$_2$O binary is periclase (see also Figure \ref{fig:ternary}).
@@
 Unlike the forsterite or periclase liquidi, the observed depression of melting at low water contents exceeds that predicted by even the most extreme commonly-used ideal mixing model, where all oxygens in the melt are equivalent and available for protonation, and molecular H$_2$O is absent \citep{SS1985}.
-The liquidus at higher water contents can be fit well with model parameters $G_{(1)} = 120 \cdot (1750-T)$ J/mol, $r=3/2$.
+The liquidus at higher water contents can be fit well with model parameters $G_{(1)} = 190 \cdot (1700-T)$ J/mol, $r=3/2$.
 The disagreement between this model and the experimental observations is small, and could plausibly be due to experimental error.
-However, given the results for stishovite (see next section), we feel that more complex models are required to explain the depression of melting in the MgSiO$_3$ system.
+However, given the results for stishovite (see next section), we feel that MgSiO$_3$-H$_2$O melts exhibit significant nonideality even at 13 GPa.
 
@@
 However, there is very little evidence for melt loss from the capsules; the lids on the capsules were well-sealed and showed no evidence of melt escape.
-The implication is that for a given bulk H$_2$O content, a sharp rise in the degree of partial melting occurs when bulk compositions along the en+H$_2$O join are heated above 1500 $^{\circ}$C.
+The implication is that for a given bulk H$_2$O content at 13 GPa, a sharp rise in the degree of partial melting occurs when bulk compositions along the en+H$_2$O join are heated above 1500 $^{\circ}$C.
 
@@
 The 13 GPa SiO$_2$-H$_2$O binary phase diagram is shown in Figure \ref{fig:SH}.
-While the deviation between the predicted enstatite liquidus depression for ideal molecular H$_2$O-free melts and the experimental observations are relatively small, the deviation for the stishovite data is extreme.
-The mixing model presented in Figure \ref{fig:SH} has parameters $G_{(1)} = 100 \cdot (1200-T)$ J/mol, $r=2$ is fit to the extrapolation of the experimentally defined liquiudus, but clearly fails to fit any of the experimental data.
+While the deviation of the experimental results from the predicted enstatite liquidus depression for ideal molecular H$_2$O-free melts was relatively small, the deviation for the stishovite results is extreme.
+The mixing model presented in Figure \ref{fig:SH} has parameters $G_{(1)} = 100 \cdot (1200-T)$ J/mol, $r=2$ is fit to the extrapolation of the experimentally defined liquidus, but clearly fails to fit any of the experimental data.
 There are several possible explanations for this.
-First, the melting point of anhydrous stishovite (which is metastable with respect to coesite) could be overestimated by ca.
+The melting point of anhydrous stishovite (which is metastable with respect to coesite at 13 GPa) could be overestimated by ca.
 300 $^{\circ}$C, or the entropy of melting could be 10--20 J/K/mol, rather than 40 J/K/mol.
-Neither of these possibilities is very likely; the coesite and stishovite melting curves provide good constraints on the volume and entropy of SiO$_2$ melt in the region of the triple point at ca.
+Neither of these possibilities seems very likely; the coesite and stishovite melting curves \citep{ZLGHF1993} provide good constraints on the volume and entropy of SiO$_2$ melt in the region of the triple point at ca.
 13.5 GPa.
-A third possibility is that liquidus-phase stishovite crystals were missed during SEM analysis.
-In the most extreme cases, the missing fraction of crystals would have been on the order of 30\%.
+An alternative possibility is that liquidus-phase stishovite crystals were missed during SEM analysis.
+In the most extreme cases, the missing fraction of crystals would have to have been on the order of 30\%.
 This is also extremely unlikely; the equant and dendritic morphologies of liquidus and quench stishovite are instantly recognisable, and the thermal gradients in the capsule lead to melt separation on the timescales of the experiments, such that even a small (5\%) volume proportion of solids should be detected.
@@
 We justify this by noting that speciation is likely to be only weakly dependent on pressure under these conditions.
+We also approximate the activity of Mg$_2$SiO$_4$ in the solid phases to be 1, such that the additional water in wadsleyite and ringwoodite does not affect the calculated liquidus curves.
 Water concentrations in the solid are taken from the original studies.
@@
 \begin{equation}
-H_2O + O = (OH)_2
+H_2O_{melt} + O_{solid} \rightleftharpoons (OH)_{2\, solid}
 \end{equation}
@@
 Both of these assumptions are probably reasonable at low values of $c_{H_2O}$.
+The activity of H$_2$O in the melt can be obtained from our melt model.
 Rearranging now produces the expression:
@@
 
-We calculate the fugacity of pure water $f_{H_2O}$ from the equation of state of \citep{PS1995}.
-We can now fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on forsterite, wadsleyite and ringwoodite using our melt model to provide estimates of $a_{H_2O}$.
-The sharp spike in olivine water concentrations at 1250 $^{\circ}$C reported by \cite{SFNHB2006} is difficult to fit with reasonable enthalpies, but we obtain good fits with the rest of their data, and those of \cite{LSKO2009}.
-The wadsleyite data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a low pressure effect on water content, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}.
+We calculate the fugacity of pure water $f_{H_2O}$ from the equation of state of \cite{PS1995} and then fit the values of A, $\Delta H$ and $\Delta V$ to the experimental data on forsterite, wadsleyite and ringwoodite (Table \ref{table:partitioning}).
+The forsterite model of \cite{ZGK2004} is in extremely good agreement with the experimental data \citep{SFNHB2006, LSKO2009}, notwithstanding a sharp spike in olivine water concentrations at 1250 $^{\circ}$C reported by \cite{SFNHB2006}.
+The data of \cite{DDFK2005} at 1200 $^{\circ}$C and 14--17 GPa are consistent with a weak pressure dependence of water content in wadsleyite, implying that $\Delta V \sim 10^{-5}$ m$^3$/mol, similar or slightly lower than estimates for olivine \citep[1.00 -- 1.06 $\cdot 10^{-5}$ m$^3$/mol][]{KKR1996, ZGK2004, MDAR2006}.
 A single data point at 18 GPa has much lower water contents, which would require $\Delta V \sim 1.25 \cdot 10^{-5}$ m$^3$/mol.
@@
 We therefore favour the lower estimate of $\Delta V$, and assume the same value for ringwoodite.
-The model values are shown in Table \ref{table:partitioning}.
 
@@
 \centering
-\caption{Thermodynamic models for water solubility in wadsleyite and ringwoodite}
+\caption{Thermodynamic models for water solubility in forsterite, wadsleyite and ringwoodite. The forsterite model is taken from the literature \cite{ZGK2004}.}
 \label{table:partitioning}
 \begin{tabular}{lllll}
-Phase & $P_{\textit{\tiny{ref}}}$ ($10^{10}$ Pa) & A [$10^{-12}$ kg/(kgPa)] & $\Delta H_{P_{\textit{\tiny{ref}}}}$ [$10^{5}$ J/mol] & $\Delta V$ [$10^{-5}$ m$^3$/mol] \\
+Phase & $P_{\textit{\tiny{ref}}}$ (GPa) & A [$10^{-12}$ kg/(kgPa)] & $\Delta H_{P_{\textit{\tiny{ref}}}}$ [kJ/mol] & $\Delta V$ [cm$^3$/mol] \\
 \hline
@@
 % Olivine & 1.2e10 & 5e-12 & 1.70e5 & 1e-5 \\ Zhao et al., 2004, water contents much too high
-Olivine & 1.2 & 3.23 & 1.50 & 1 \\
-Wadsleyite & 1.5 & 1.97 & 1.60 & 1 \\
-Ringwoodite & 2.0 & 1.46 & 2.03 & 1 (fixed)
+Olivine & 12 & 13.5 & 170 & 10 \\
+Wadsleyite & 15 & 1.8 & 160 & 10 \\
+Ringwoodite & 20 & 1.2 & 200 & 10 (fixed)
 \end{tabular}
@@
 These models allow us to predict the partition coefficients of water between olivine, wadsleyite and melt at the 410 km discontinuity ($\sim$14 GPa; Figure \ref{fig:partitioning_wad_fo}), and wadsleyite, ringwoodite and melt at the 520 km discontinuity ($\sim$18 GPa; Figure \ref{fig:partitioning_rw_wad}).
-$D^{wad/fo}_{H_2O}$ displays a decrease from $\sim$4 at 1000 $^{\circ}$C to $\sim$1.8 at 2000 $^{\circ}$C, in good agreement with the estimates of \cite{LSOK2011} at low temperatures, but without a return to high values at high temperatures, where water contents are lower and prone to large relative errors.
-$D^{ring/wad}_{H_2O}$ decreases from $\sim$1.4 at 1000 $^{\circ}$C to $\sim$1.1 at 2000 $^{\circ}$C.
+$D^{wad/fo}_{H_2O}$ decreases from $\sim$6 at 1000 $^{\circ}$C to $\sim$1.1 at 2000 $^{\circ}$C, in good agreement with the estimates of \cite{LSOK2011} at low temperatures, but without a return to high values at high temperatures, where water contents are lower and prone to large relative errors.
+$D^{ring/wad}_{H_2O}$ decreases from $\sim$1.7 at 1000 $^{\circ}$C to $\sim$1.1 at 2000 $^{\circ}$C.
 These values imply that partial melting will accompany upwelling of H$_2$O-saturated ringwoodite and wadsleyite across both the '520'- and the '410'-km discontinuities.
@@
 Both of these observations imply efficient hydroxylation of the silicate network and strong interactions between species, which would be interesting subjects for molecular dynamic studies.
-We also hope that a combination of our liquidus constraints with parameterisations for density and viscosity will be used to inform more realistic models of melt migration above the mantle transition zone and during subduction.
+We also hope that a combination of our liquidus constraints with parameterisations for density and viscosity will be used to inform more realistic models of hydrous melt migration in the deep mantle.
 
@@
 This study was funded by the ACCRETE project (European Research Council Advanced Grant, contract number 290568).
+Python files recreating the results from this paper may be downloaded from the BurnMan webpage (\url{https://geodynamics.org/cig/software/burnman/}), and are also available from RM.
 

### 2016-05-12 revised_submission/cover_letter/cover_letter.tex
Submitted

% cover letter for GCA submission

\documentclass[10pt]{letter}

% Colors
\usepackage{fancyvrb}
\usepackage[table]{xcolor}
\definecolor{tableShade}{HTML}{F1F5FA}   %iTunes
\definecolor{tableShade2}{HTML}{ECF3FE} %Finder

\definecolor{cambridgedarkblue}{RGB}{0,62,114}
\definecolor{cambridgeblue}{RGB}{0,115,207}
\definecolor{cambridgelightblue}{RGB}{106,173,228}

\definecolor{cambridgedarkorange}{RGB}{200,78,0}
\definecolor{cambridgeorange}{RGB}{227,114,34}
\definecolor{cambridgelightorange}{RGB}{239,189,71}

\definecolor{cambridgedarkgreen}{RGB}{67,81,37}
\definecolor{cambridgegreen}{RGB}{88,166,24}
\definecolor{cambridgelightgreen}{RGB}{168,180,0}

\definecolor{cambridgedarkviolet}{RGB}{65,45,93}
\definecolor{cambridgeviolet}{RGB}{142,37,141}
\definecolor{cambridgelightviolet}{RGB}{181,147,155}

\definecolor{cambridgedarkazur}{RGB}{21,101,112}
\definecolor{cambridgeazur}{RGB}{0,179,190}
\definecolor{cambridgesportingblue}{RGB}{163,193,173}
\definecolor{Grey}{rgb}{0.5, 0.5, 0.5}
\newcommand{\grey}[1]{\textcolor{Grey}{#1}}

\RequirePackage{hyperref}
\hypersetup{pdftex, plainpages = false, pdfpagelabels, 
  pdfpagelayout = useoutlines,dvips, 
  pdfauthor={R. Myhill, myhill.bob@gmail.com},
  pdftitle={Cover letter},
  pdfsubject={Cover letter},
  pdfkeywords={},
  pdfproducer={pdfLaTeX with hyperref},
  pdfcreator={pdfLaTeX}
  bookmarks,
  bookmarksopen = false,
  bookmarksnumbered = true,
  breaklinks = true,
  linktocpage,
  colorlinks = true,
  linkcolor = cambridgeblue,
  urlcolor  = cambridgeviolet,
  citecolor = cambridgedarkorange,
  anchorcolor = cambridgedarkazur,
  hyperindex = true,
  hyperfigures}

\oddsidemargin=.2in
\evensidemargin=.2in
\textwidth=5.9in
\topmargin=-.5in
\textheight=9in

\name{Robert Myhill \\
School of Earth Sciences\\
University of Bristol\\
Wills Memorial Building\\
Queens Road\\
Bristol BS8 1RJ\\
United Kingdom \\
Tel: +44 (0) 117 954 5400 \\
\href{mailto:bob.myhill@bristol.ac.uk}{bob.myhill@bristol.ac.uk}}

\date{\today}

\begin{document}

\begin{letter}{}

\opening{Dear editor,}

Please find attached our revised manuscript entitled \emph{Hydrous
  melting and partitioning in and above the mantle transition zone:
  insights from water-rich MgO-SiO$_2$-H$_2$O experiments}, written
for potential publication as an \emph{Article} in \emph{Geochimica et
  Cosmochimica Acta}.

We have addressed all of the comments made by the reviewers. 
I have also made some minor changes to the text to improve
readability, and simplified some of the analysis so that it can be
more easily reproduced by others. The simplification (using the
intermediate composition melts from de Koker et
al., rather than creating a solution model) has a
negligible effect on the figures and conclusions in the paper.

\textbf{Responses to editorial and reviewer comments}
\begin{Verbatim}[commandchars=+\[\]]
Line 28: could use a reference at the end of this sentence (e.g. Spiegelman).
+underline[Two references added]

Line 33: could use a reference at the end of sentence about water
budget 
(e.g. Hirschmann)
+underline[Reference added]

Line 60: I think you might want a better segue here into the approach
you take in this study. Something like "Here we use the approach of 
bracketing the liquidus surface to determine melt composition as a 
function of temperature in the system MSH…".
+underline[The end of the introduction has been rewritten, starting with this suggestion.]

I found the introduction to be well written but in the end it fails to
let the reader know exactly what the tack is and what the goals of
this paper are, and even what system you will be working in and why.  
You could leave the business about how to make starting materials 
with high water contents until the methods, and end the introduction 
by telling the reader that you will use the liquidus phase relations
in MSH to develop thermodynamic models for water solubility in melts 
and partitioning models, which you will apply to understanding hydrous 
melting at transition zone conditions.
+underline[I agree. I've rewritten this paragraph]

Lines 70-75 (end of the introduction): Again, I agree with Mike that
the end of the intro section is a bit abrupt and suddenly becomes too
technical about the choice of starting mix water source. Such
information should become part of the methods (starting mix)
section. 
The authors are better of ending the introduction by saying generally 
what they have done in this study.
+underline[Paragraph rewritten.]

Line 78: "Starting compositions in the system MSH…"  
It also might be good to provide some explanation of the rationale 
for choosing the starting compositions, i.e. "We chose compositions 
along the bounding binaries, and within the ternary in order to 
contrain phase relations…"
+underline[Good suggestion. Detail added.]

Line 105: Any idea about water loss in any of the experiments? 
How much time do you have as a function of temperature, or is 
this intuition from experience?
+underline[Added reference to Novella et al. (2016), which has more details]

Line 137: "…concentrated hydrous melts". Presumably you mean water-rich.
+underline[Changed to “water-bearing silicate melts”]

Line 193: typo, should be 'note' not 'notes'
+underline[Typo corrected]

Line 203: I am probably just being thick, but here and other places 
you talk about "the experimentally determined value". 
What value are you referring to and from what source?
+underline[These values are from Stixrude and Lithgow-Bertelloni (2011).]
+underline[I cite the study for periclase (line 190), but have added another]
+underline[reference to the paper here.]

Line 227: You might consider a new section here as currently the 
segue is a bit clunky.
+underline[Section split into endmember and solution model subsections]

Line 269: Should read "…between quench crystals and liquidus phases…."
+underline[Corrected]

Table 1:  It took me a little while to get my head around the way the 
results are tabulated in Table 1, but in the end I think it works. You 
may want to add in the table caption that melts are present in all run 
products. There is also a typo in the caption "en-dash" should just by 
"dash".
+underline[Done]

Line 276: The invariant is drawn at ~1250 C rather than 1200. 
Also, why call it a fluid here… I would stick with liquid. You may
also want to clarify the key either by adding e.g. Brucite-L 
(or maybe br-L), etc.
+underline[Text changed to 1250C rather than 1200. ]
+underline[Reference to fluid changed to liquid. I haven't changed the key, ]
+underline[as all legend entries implicitly include liquid. The addition to ]
+underline[the caption of Table 1 should be sufficient (although I can]
+underline[change this if you think I should).]

Line 304. I am not understanding the use of the term 'metastable'
here.  It may be a good idea as well to be clear that Fig. 4 is not a binary
join, but is a T-X section depicting equilibrium crystallization. 
Actually, the figure is not drawn correctly based on Fig. 7, as there 
should be a fields for fo-en-liq and en-liq…
+underline[This sentence has been rephrased. I've also added the low]
+underline[temperature en-bearing fields to Figure 7]

Line 312: "…explained by an MgO…"
+underline[Corrected!]

The concave up nature if the liquidii in this study is very
interesting and convincing. Is there any chance this in part reflects
H2O loss in higher water content experiments? I don't think so, 
but just a thought.
+underline[Interesting thought. Given the different temperatures at which the ]
+underline[liquidus curves turn over in the fo-H2O, en-H2O and stv-H2O systems, ]
+underline[I suspect not. Yamada et al. (2004) saw something very similar, and we ]
+underline[didn't see any evidence for melt loss. I've made a comment to this]
+underline[effect at the end of the MgSiO3-H2O subsection.]

Fig. 7. I don't like the way you label the T contours on the
sidelines. I think you can label them within the ternary as is
typically done, and then you should plot the phases that occur on the 
bounding binary joins (e.g. br, fo, en). The light grey points are not 
explained in he caption and are a distraction. I also found a few 
inconsistencies with the liquidus phase relations depicted and the 
results in Table 1, but it was not clear to me whether this diagram 
is based on your best guess or is made from the thermodynamic 
model.
+underline[Changed labelling]
+underline[Removed grey points]
+underline[Added description of how the figure was created]

Line 388. I am not sure about this sentence. What you say is true, 
but the amount of melt will always depend on the water content of 
the source, and so a rapid increase in mass fraction is a relative
thing. If the source contains a few hundred ppm water, then there 
will never be a lot of melt, even though the mass fraction may
increase from say 0.0001 to 0.01, it is still small and hard to
detect. 
Seismic detection will depend mostly on their alignment 
The change in composition and density is of course very important to
its mobility.
+underline[I've reworked these sentences to focus on the channelisation,]
+underline[which will be promoted by the increase in silicate solubility in the]
+underline[melt and increase melt fraction and connectivity.]

Lines 385-390: Similar to the comments of the reviewer, I also note 
that this argument is unlikely to be correct. The melt proportion is 
controlled by the bulk water content of the source and the D of water.
 So even if water content in melt changes significantly, that does not 
necessarily mean generation of seismically detectable melt fraction. 
+underline[See above]

Plus, a temperature change of 1300 to 1700 degree C is huge. Is it 
likely that such a large change in T is realized even for deeply 
subducted slab-mantle interface, which should already be thermally 
equilibrated more than in typical subduction zone conditions?
+underline[I've now changed the sentence to be more specific about temperature]
+underline[ranges in the mantle at ~400 km.]

Line 395. Need a ref here regarding channelization.
+underline[Reworded to link the two comments on channelization.]

Conclusion: This may simply be a writing style issue, but the 
discussion section seems to end somewhat abruptly. Would the 
paper benefit from a small concluding remarks section or something 
similar?
+underline[Agreed. I've added a concluding remarks section, as suggested.]

Line 476; Figure # should be Figure 9
+underline[Corrected]

Figure 9 and 10: In the figure captions write that these are partition 
coefficients for H2O. Of course you mentioned in the text but from 
the figures one can't tell its partition coefficient for what.
+underline[Done]

Figure 9 and 10 y-axes: Again, write H2O as subscript to the Ds 
so that the figures are self-explanatory.
+underline[Corrected]
\end{Verbatim}
\closing{Best wishes,}

\end{letter}

\end{document}

### 2016-05-12 revised_submission/manuscript/MSH.tex
Submitted

--- 2016-05-11 revised_submission/MSH.tex
@@
 \cortext[mycorrespondingauthor]{Corresponding author: R. Myhill}
-\ead{myhill.bob@gmail.com}
+\ead{bob.myhill@bristol.ac.uk}
 
@@
 
+Here we use the approach of bracketing the liquidus surface to determine melt composition as a function of temperature in the system MgO-SiO$_2$-H$_2$O (MSH).
+This investigation is focussed on the behaviour at 13 GPa, which corresponds to the pressure just above the mantle transition zone, and is therefore relevant to the release of hydroxyl-bearing melts during deep subduction and during upwelling of hydrous wadsleyite.
+We use these results to develop thermodynamic models for water solubility in melts, and apply these to investigate water partitioning between melts and the olivine polymorphs (olivine, wadsleyite and ringwoodite).
+We then use our results to discuss the composition and properties of hydrous melts under transition zone conditions.
+
+\section{Experimental and analytical methods}
 Constraining melt compositions remains a major challenge for the field of high-pressure experimental petrology.
@@
 During quenching, some of the water in the melt crystallises as hydrous minerals such as brucite, but a significant proportion crystallises as water ice at high pressure, which is then lost as water vapour during decompression and subsequent sample preparation.
-The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in microprobe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}.
+The original H$_2$O contents of the partial hydrous melts are normally estimated by mass balance and/or by deficits in electron probe totals \citep[e.g.][]{OMY2000, DDFK2005, LSOK2011, MSUP2007}.
 Using mass balance to calculate H$_2$O contents depends on being able to estimate the volume proportions of solid and quench phases, recognise and subtract quench overgrowths where they exist, and calculate the effective ambient-pressure density of the melt.
 Such estimates are typically associated with significant uncertainties.
-Deficits in microprobe totals are sometimes reasonable for estimating water contents in solid phases, provided the microprobe standards are suitable, and the microprobe is exceptionally well-calibrated (even a small deficit implies a large water content).
+Deficits in electron probe totals are sometimes reasonable for estimating water contents in solid phases, provided the electron probe standards are suitable, and the electron probe is exceptionally well-calibrated (even a small deficit implies a large water content).
 However, for quench melts the deficits are attributable to water held in quench phases \emph{and} porosity within the volume sampled by the defocused beam.
@@
 
-Here we use the approach of bracketing the liquidus surface to determine melt composition as a function of temperature in the system MgO-SiO$_2$-H$_2$O (MSH).
-This technique requires knowledge of the starting composition, techniques to effectively minimise water loss from the capsule and an ability to recognise the difference between liquidus and quench phases.
-We use this technique to develop thermodynamic models for water solubility in melts, and apply these to models of water partitioning between mantle phases and melts.
-We then apply these models to better understand hydrous melting under transition zone conditions.
-%XXXX what, in general, have you done in this paper?
-
-\section{Experimental and analytical methods}
+We avoid the above pitfalls by instead bracketting the liquidus surface with a series of distinct starting compositions.
+This technique is not without its own challenges.
+For example, the composition must be well-constrained and not change during the experiment, which requires techniques to effectively minimise water loss from the capsule.
+It must also be possible to recognise the difference between liquidus and quench phases.
 Starting compositions were created from a mixture of high-purity brucite, quartz and platinic acid (hexahydroxyplatinate(IV); H$_2$Pt(OH)$_6$).
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/periclase}
+      \includegraphics[width=0.8\textwidth]{figures/periclase_edited}
   \caption{The periclase-water phase diagram at 13 GPa, based on current experimental results.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/forsterite}
+      \includegraphics[width=0.8\textwidth]{figures/forsterite_edited}
       \caption{The forsterite-water phase diagram at 13 GPa, based on current experimental results.
-The thermodynamics of pure forsterite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
+The thermodynamics of pure forsterite and the anhydrous melt are taken from the literature \citep{SLB2011, DKS2013}.
 At high temperatures, the forsterite liquidus is metastable with respect to periclase + melt.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/enstatite}
+      \includegraphics[width=0.8\textwidth]{figures/enstatite_edited}
   \caption{The clinoenstatite-water phase diagram at 13 GPa, based on current experimental results.
-The thermodynamics of pure high pressure clinoenstatite are taken from the literature \citep{SLB2011}, and the properties of the anhydrous melt are from the mixing model described in the text.
+The thermodynamics of pure high pressure clinoenstatite and the anhydrous melt are taken from the literature \citep{SLB2011, DKS2013}.
 High temperature data come from the study of \cite{NDMPMF2016}.}
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/stishovite}
+      \includegraphics[width=0.8\textwidth]{figures/stishovite_edited}
       \caption{The silica-water phase diagram at 13 GPa, based on current experimental results.
@@
 The causes of this additional stability (and, to a lesser extent, that of hydrous MgSiO$_3$ melts) are well beyond the scope of the current study, but may be a fruitful line of investigation for molecular dynamics simulations \citep[e.g.][]{KS2010}.
-
-%\citep{SU2001, KOM2005, AA1980, Wunder1998, AFRH2001, Nagano2002, MFY2002, Sta2004, MFY2004, SUTG2001, NM2008, DM2010, ZK1998}
 
@@
 The locations of the contours are constrained by observed liquidus phases and the liquidus-bracketing results presented here.
-Previous work on hydrous phase stability and Mg:Si ratios for cotectic melts are also included \citep{MSUP2007}.}
+Previous work on hydrous phase stability and Mg:Si ratios for cotectic melts are also included \citep{MSUP2007}.
+The MgO-SiO$_2$ binary is constructed using a Margules solution model for the melt (see Supplementary Material).}
   \label{fig:ternary}
@@
 
-% Densities
-%\cite{MF2010}
-%\cite{BK2003}
-%\cite{MSK2008}
-
-%So at least in my models there are three quantities that are important for the pattern of melt migration, the permeability, the compaction viscosity and the shear viscosity (if you have elasticity/plasticity, other things come in, too).
-Everywhere in the equations where you have the permeability, it's divided by the melt viscosity, and the permeability itself is normally assumed to scale with phi^3 (or less common phi^2), and at least I didn't see any parametrisations describing the effect of temperature on the permeability, so I would assume it's small.
-So for the permeability, you can just use the values you get for your different P-T-X conditions, and see if the overall term permeability/melt_viscosity increases or decreases.
-
-%For the viscosities of the solid you have the influence of temperature on the viscosity (which I think is similar for shear and compaction viscosity) and you have the effect of porosity, eta ~ exp(alpa * phi) with phi~-27 for the shear viscosity (I think this is what Rich Katz uses in his models, but I would have to look it up to be sure), and xi ~ phi0/phi for the compaction viscosity, and you can do the same here, compare how they change for your different P-T-X conditions.
-
 \subsection{Water partitioning in the mantle transition zone}
@@
 In quenchable phases, there are a number of ways that concentration can be measured, including FTIR, SIMS and ERDA.
-In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in microprobe totals is due to the fraction of oxygen bonded to hydrogen.
-In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from microprobe deficits.
+In the case of melts, which are unquenchable, $c^{melt}_{H_2O}$ in partitioning studies has most commonly been estimated by mass balance or by assuming that the deficit in electron probe totals is due to the fraction of oxygen bonded to hydrogen.
+In two studies on Fe-free wadsleyite, \cite{DDFK2005} and \cite{LSOK2011} present estimates of water content in melts calculated from electron probe deficits.
 A similar study on ringwoodite was undertaken by \cite{OMY2000}.
@@
   \centering
-      \includegraphics[width=0.8\textwidth]{figures/fo_wad_rw}
-  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa. Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.}
+      \includegraphics[width=0.8\textwidth]{figures/fo_wad_rw_edited}
+  \caption{Water concentrations in wadsleyite, ringwoodite and melt, modelled at 15 and 20 GPa.
+Experimental data is taken from three published studies \citep{OMY2000,DDFK2005,LSOK2011}.
+Unfilled and filled symbols correspond respectively to the estimates of solid and melt H$_2$O contents.}
   \label{fig:fo_wad_rw_melt}
@@
 \hline
-% Olivine & 1.2e10 & 1.74e-11 & 1.70e5 & 1e-5 \\ Not a bad fit for the Zhao enthalpy
-% Olivine & 1.2e10 & 5e-12 & 1.70e5 & 1e-5 \\ Zhao et al., 2004, water contents much too high
 Olivine & 12 & 13.5 & 170 & 10 \\
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410}
+  \includegraphics[width=0.8\textwidth]{figures/wad_fo_partitioning_410_edited}
   \caption{Modelled H$_2$O partition coefficient between forsterite, wadsleyite and melt at 14 GPa, approximately the pressure of the 410 km discontinuity.
 Note the separate scales for solid/solid and solid/melt partitioning.
-The difference in melt compositions between this study and that of \cite{DDFK2005} results in much lower $D^{wad/melt}$ coefficients.
+The higher melt hydroxyl contents in this study compared to that of \cite{DDFK2005} results in much lower $D^{wad/melt}$ coefficients.
 Error bars on the wadsleyite datapoints are those reported by \cite{DDFK2005}.}
@@
   \centering
-  \includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520}
+  \includegraphics[width=0.8\textwidth]{figures/rw_wad_partitioning_520_edited}
   \caption{Modelled H$_2$O partition coefficient between wadsleyite, ringwoodite and melt at 18 GPa, approximately the pressure of the 520 km discontinuity.
@@
 
+The authors thank Raj Dasgupta and Mike Walter for their insightful comments which greatly improved the manuscript.
+
 \clearpage
